---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Supplementary materials
```{r echo=FALSE, message=FALSE}
#| output: false

# Source the founctions in simulaton.qmd with a temporary R script containing only code blocks
temp_r_file <- tempfile(fileext = ".R")
knitr::purl("simulation.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
knitr::purl("index.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
unlink(temp_r_file)
rm(temp_r_file)
```

## Example opioid equivalence calculation {#sec-conversion-example}

**Question**: What dose of intravenous (IV) fentanyl will provide equal the analgesic effect of 5 mg of IV morphine, assuming IV fentanyl is 100 times more potent than IV morphine.

**Calculation:**

**Step 1:** Use the potency value to find the conversion value to be 0.01

$$\begin{aligned}
\rho_{\text{fentanyl,IV}} = 100 \\
c_{\text{fentanyl,IV}} = \frac{1}{\rho_{\text{fentanyl,IV}}} \\
c_{\text{fentanyl,IV}} = \frac{1}{100} = 0.01 \\
\end{aligned}$$

**Step 2:** Multiple the reference dose of 5mg of morphine by the conversion value to find the equivalent dose of fentanyl to be equal to 0.05 mg or 50 mcg.

$$\begin{aligned}
d_{\text{morphine}} \times c_{\text{fentanyl,IV}} = d_{\text{fentanyl}} \\[1em]
5 \times 0.01 = 0.05\text{mg} = 50\text{mcg}
\end{aligned}$$

This approach is used clinically for opioid conversion or rotation and is the origin of the opioid equivalence system. With a small set of opioid equivalence values, it's possible to convert doses between a wide range of different opioids, often via the reference medication as an intermediate.

Medications administered in different dose units are generally converted to a single dose unit (often **mg**) and the conversion values updated to reflect this however in some cases they may be kept in their administered unit (**e.g. mcg**) and the conversion shifted by an appropriate factor.


## Mathematical proofs

### Rearrangement of @eq-mu-ome-final {#sec-equation-proof}

The equation for the group mean OME provided @eq-mu-ome-1 is is a weighted sum of the OME values for patients $j$ in group $G$.

$$
\begin{aligned}
\text{OME}_j &= \sum_{i\in\mathbf{M}} d_i c_{i,r} \\
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j=G} \text{OME}_j
\end{aligned}
$${#eq-mu-ome-1}
First the the expression for $\text{OME}_j$ is substituted into the group mean. Then the order of summation is rearranged as both the sets $G$ and $\mathbf{M}$ are finite. Following this, the constant $c_{i,r}$ is factored out of the inner sum whilst the scaling term $\frac{1}{|G|}$ of group size is distributed to the inner term. 

$$
\begin{aligned}
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j \in G}  \left( \sum_{i\in\mathbf{M}} d_{j,i} c_{i,r} \right) \\
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j \in G}  \sum_{i\in\mathbf{M}} d_{j,i} c_{i,r} \\
\mu_{\text{OME}}^{(G)} &= \sum_{i\in\mathbf{M}} \left( \frac{1}{|G|} \sum_{j=1}  d_{j,i}  \right) c_{i,r}
\end{aligned}
$$

We can then define the mean dose $\bar{d}$ of medication $i$ as the arithmetic mean of doses for patients $j \in G$. 

$$
\bar{d_i} = \frac{1}{|G|} \sum_{j \in G} d_{j,i}
$$
This mean dose is then substituted back into the equation to yield the final rearrangement as seen in @eq-mu-ome-final. 

$$
\mu_{\text{OME}}^{(G)}  = \sum_{i\in\mathbf{M}} \bar{d_{i}} c_{i,r} 
$$

## Literature search

### PUBMED search terms {#sec-appendix-pubmed-1}

On 14th of July 2026, a search was performed on the PUBMED online search tool using the following search term to identify the volume of literature using opioid equivalence metrics in research. The search term was "(Milligram morphine equivalent) OR (Oral morphine equivalent)" with a 10 year filter was used and yielded the following full search string. There was 4,031 results. 

::: {.callout-note appearance="simple" title = "PUBMED search term"}
((("milligram"[All Fields] OR "milligrams"[All Fields]) AND ("morphine derivatives"[Supplementary Concept] OR "morphine derivatives"[All Fields] OR "morphines"[All Fields] OR "morphine derivatives"[MeSH Terms] OR ("morphine"[All Fields] AND "derivatives"[All Fields]) OR "morphine"[Supplementary Concept] OR "morphine"[All Fields] OR "morphin"[All Fields] OR "morphine"[MeSH Terms] OR "morphine s"[All Fields] OR "morphinic"[All Fields]) AND ("equivalence"[All Fields] OR "equivalences"[All Fields] OR "equivalencies"[All Fields] OR "equivalency"[All Fields] OR "equivalent"[All Fields] OR "equivalently"[All Fields] OR "equivalents"[All Fields])) OR (("mouth"[MeSH Terms] OR "mouth"[All Fields] OR "oral"[All Fields]) AND ("morphine derivatives"[Supplementary Concept] OR "morphine derivatives"[All Fields] OR "morphines"[All Fields] OR "morphine derivatives"[MeSH Terms] OR ("morphine"[All Fields] AND "derivatives"[All Fields]) OR "morphine"[Supplementary Concept] OR "morphine"[All Fields] OR "morphin"[All Fields] OR "morphine"[MeSH Terms] OR "morphine s"[All Fields] OR "morphinic"[All Fields]) AND ("equivalence"[All Fields] OR "equivalences"[All Fields] OR "equivalencies"[All Fields] OR "equivalency"[All Fields] OR "equivalent"[All Fields] OR "equivalently"[All Fields] OR "equivalents"[All Fields]))) AND (y_10[Filter])
:::


### PUBMED search terms {#sec-appendix-pubmed}

The search term was executed on the PUBMED online search tool on 7th July 2026. 

::: {.callout-note appearance="simple" title = "PUBMED search term"}
(((((("opioid equivalent" OR "oral morphine equivalence" OR "oral morphine equivalent" OR "oral morphine equivalents" OR "OME" OR "morphine milligram equivalents" OR "morphine milligram equivalence" OR "MME" OR "morphine equivalent daily dose" OR "MEDD") AND ("perioperative" OR "peri-operative" OR "postoperative" OR "post-operative" OR "surgical" OR "surgery" OR "post-surgical" OR "postsurgical"))) AND ("2025/01/01"[Date - Publication] : "2025/12/31"[Date - Publication]))) AND ("clinical trial"[Filter] OR "clinical trial" OR "controlled clinical trial" OR "randomized controlled trial" OR "interventional" OR "intervention study" OR "clinical trials as topic"[MeSH Terms] OR "cohort studies"[MeSH Terms] OR "cohort" OR "prospective" OR "retrospective" OR "case-control" OR "comparative study" OR "evaluation study"))
:::

### Literature search inclusion criteria {#sec-appendix-inclusion}

The literature was screen with an eligibility criteria following the PCC framework. 

```{r}
#| label: tbl-screening
#| tbl-cap: "Litrature extraction proforma"
#| output: TRUE

tbl_screening <- data.frame("Criteria" = c("Population", "Concept","Context", "Other"),
                            "Include" =c("All patient groups undergoing a surgical procedure", "Studies using a total opioid burden metric as a primary or secondary outcome (Such as oral morphine equivalence (OME), milligram morphine equivalence (MME) or opioid equivalence)", "Patients undergoing surgical procedures", "Any studies comparing a control against and intervention or two interventions."),
                            "Exclude" = c("Patients not undergoing surgical procedures. Non-human studies", "Outcomes not including opioid totals", "Palliative care studies or Pharmacology studies", "Studies with no difference between intervention groups" ))

knitr::kable(tbl_screening)

```
### Covidence extraction form {#sec-appendix-cov-extraction}

Data was extracted from selected papers using the following proforma by a single author. Authors could leave notes to highlight specific features for later analysis required.

```{r}
#| label: tbl-extraction-form
#| tbl-cap: "Litrature extraction proforma"
#| output: TRUE

tbl_extraction_form <- data.frame("Variable" = c("Study type",
                          "Opioid conversion set",
                          "Intervention",
                          "Opioid proportions reported between intervetnion groups",
                          "Outcome",
                          "Data avalibility"),
           "Selection options" = c("Randomised controlled trial, Cohort study, Case controlled study, Other",
                         "Cited, Included in paper, Not described",
                         "Surgical intervention, Anaesthetic intervnetion, Other",
                         "Described or Not described",
                         "Difference in opioid requirement/treatment, No difference",
                         "Publically avalible, Not publically avaliable, Database study"),
           "Explanatory text" = c("What is the format of the study?",
                                  "In the methods section, is how is the OME metric generated described and how are the conversion values referenced?",
                                  "What type of intervention is given. For pre/post-operative interventions that include changes to medication select anaesthesia. For descriptive studies analysing differences select other.",
                                  "Assess the results section, Is it possible to establish the proprotion, mean or total doses of each opioid between intervention groups. If authros explicilty state only one type of opioid used in the OME/MME metric select 'Described'.",
                                  "In the primary outcome of OME/MME, is there a significant difference reported between intervention groups.",
                                  "Is data avalible for readers to ascertain the proportions of opioids between groups? If OME/MME metric only avaliable then select 'Not avalible'."))

knitr::kable(tbl_extraction_form,
    col.names = c("Variable", "Options", "Explanatory text")
)

```

### Grid search of plausible conversion values to studies {#sec-appendix-neilsen-ranges}

The values of the possible plausible conversion values are taken from Table 1 of Neilsen's paper on a unified conversion value set. These conversion values are sourced from national clinical guideline documents from a range of countries. 

Medication's were assumed to be administered IV with the exception of Tramadol and Hydrocodone as Neilsen only reports oral conversion values for these. 

```{r ranges}

neilsen_ranges_table <- data.frame(
  Medication = names(neilsen_ranges),
  `Conversion values` = sapply(neilsen_ranges, paste, collapse = ", "),
  check.names = FALSE,
  row.names = NULL
)

knitr::kable(neilsen_ranges_table)
```


```{r}
#| label: fig-lit-outcome-dist
#| fig-cap: "Density plot of the possible treatment effects calculated de-novo across the grid values for opioid conversion values."
#| eval: false

### EVAL IS FALSE ###

library(ggplot2)

# 2. Plot with facets
ggplot(plot_data, aes(x = delta_ome)) +
  # Use geom_density for a smooth distribution
  geom_density(fill = "steelblue", alpha = 0.6, color = "black") +
  # Use facet_wrap with free scales so each study's range is visible
  facet_wrap(~ authors, scales = "free", ncol = 3) + 
  theme_minimal(base_size = 12) +
  labs(
    title = "Sensitivity Analysis: OME Relative Difference by Study",
    subtitle = "Variation based on independent uncertainty in conversion factors (c_i)",
    x = "Relative Difference (delta_ome)",
    y = "Density"
  ) +
  theme(
    strip.background = element_rect(fill = "gray95"),
    panel.spacing = unit(1, "lines")
  )

```


## Simulation study


### Effect of $\mu_{\text{OME}}^{\text{(control)}}$ on simple simulation {#sec-appendix-sim-1}

To investigate the scaling term of $\mu_{\text{OME}}^{\text{(control)}}$ on the observed treatment effect, 6 further simulations were performed, each as described in @sec-manuscript-simple-sim but with a different $\mu_{\text{OME}}^{\text{(control)}}$ between 0 OME mg and 10 OME mg. The treatment effect was fixed at 0. 

```{r}
#| echo: false
#| message: false
#| fig-width: 6
#| fig-height: 6

library(ggplot2)
library(dplyr)
library(latex2exp)

p_min <- -1
p_max <- 1
e_min <- -1
e_max <- 1
sigma = 0.1

if (PRODUCTION_QUALITY) {
  n = 10000
  space = 0.05
} else {
  n = 1000
  space = 0.1
}

simple_results_mu <- data.frame(p = numeric(), e = numeric(), diff = numeric(), mu=numeric())
for(mu in seq(0, 10, length.out = 6)){
for(p in seq(p_min, p_max, by = space)){
  for(e in seq(e_min, e_max, by = space)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu,
      potency_oxycodone = potency_oxycodone,
      potency_morphine = potency_morphine,
      error_in_c_oxycodone = e,
      treatment_effect = 0
    )
    
    simple_results_mu <- rbind(simple_results_mu, data.frame(p = p, e = e, diff = diff, mu=mu))
  }
}
}
```

```{r}
#| label: fig-appendix-simple-sim-mu
#| fig-title: Effect of scale of the control group average dose (OME mg) on observed effect size
#| fig-cap: Plots of the difference in proportion over the error in conversion values provided faceted by the expected mean dose of the control group. Colour scale demonstrates the observed difference in the treatment group. True effect size is set to 0. 
#| fig-width: 6
#| fig-height: 8

label_function <- function(val) {
  paste0('mu["OME"] == ', val)
}

plot_simple_grid_mu <- ggplot(simple_results_mu, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)'),
       caption = TeX(r'(True $\delta \mu_{{OME}} = 0$ )')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
symmetric_fill_scale(30)+
  facet_wrap(~mu, labeller = as_labeller(label_function, default = label_parsed)
             )
  
plot_simple_grid_mu

```

### Effect of $\Delta \mu_{\text{OME}}$ on simple simulation {#sec-appendix-sim-2}

To investigate the interaction of a true treatment effect on the observed treatment effect, 9 further simulations were performed, each as described in @sec-manuscript-simple-sim but with a different true treatment effect between the values of -1 and 1. 

```{r}

if (PRODUCTION_QUALITY) {
  n = 10000
  space = 0.05
} else {
  n = 1000
  space = 0.1
}

simple_results_t <- data.frame(p = numeric(), e = numeric(), diff = numeric(), t=numeric())
for(t in seq(-1, 1, length.out = 9)){
for(p in seq(p_min, p_max, by = space)){
  for(e in seq(e_min, e_max, by = space)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = 10,
      potency_oxycodone = potency_oxycodone,
      potency_morphine = potency_morphine,
      error_in_c_oxycodone = e,
      treatment_effect = t
    )
    
    simple_results_t <- rbind(simple_results_t, data.frame(p = p, e = e, diff = diff, t=t))
  }
}
}


```

```{r}
#| label: fig-appendix-simple-sim-t
#| fig-title: Effect of differing true effect size on observed effect size
#| fig-cap: Plots of the difference in proportion over the error in conversion values provided faceted by the expected mean dose of the control group. Colour scale demonstrates the observed difference in the treatment group. The mean value of the control group is set to 10 OME mg . 
#| ig-width: 6
#| fig-height: 8

label_function <- function(val) {
  paste0('delta["OME"] == ', val)
}

plot_simple_grid_t <- ggplot(simple_results_t, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)'),
       caption  = TeX(r'($\mu_{{OME}}^{(control)}$ (Mean dose of control group) $= 10$ )')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
symmetric_fill_scale(30)+
  facet_wrap(~t, labeller = as_labeller(label_function, default = label_parsed)
             )
  
plot_simple_grid_t
```

::: {.content-visible when-format="pdf"}
\listoffigures
:::
