---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Supplementary materials
```{r echo=FALSE, message=FALSE}
#| output: false

# Source the founctions in simulaton.qmd with a temporary R script containing only code blocks
temp_r_file <- tempfile(fileext = ".R")
knitr::purl("simulation.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
knitr::purl("index.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
unlink(temp_r_file)
rm(temp_r_file)
```

## Example opioid equivalence calculation {#sec-conversion-example}

**Question**: What dose of intravenous (IV) fentanyl will provide equal the analgesic effect of 5 mg of IV morphine, assuming IV fentanyl is 100 times more potent than IV morphine.

**Calculation:**

**Step 1:** Use the potency value to find the conversion value to be 0.01

$$\begin{aligned}
\rho_{\text{fentanyl,IV}} = 100 \\
c_{\text{fentanyl,IV}} = \frac{1}{\rho_{\text{fentanyl,IV}}} \\
c_{\text{fentanyl,IV}} = \frac{1}{100} = 0.01 \\
\end{aligned}$$

**Step 2:** Multiple the reference dose of 5mg of morphine by the conversion value to find the equivalent dose of fentanyl to be equal to 0.05 mg or 50 mcg.

$$\begin{aligned}
d_{\text{morphine}} \times c_{\text{fentanyl,IV}} = d_{\text{fentanyl}} \\[1em]
5 \times 0.01 = 0.05\text{mg} = 50\text{mcg}
\end{aligned}$$

This approach is used clinically for opioid conversion or rotation and is the origin of the opioid equivalence system. With a small set of opioid equivalence values, it's possible to convert doses between a wide range of different opioids, often via the reference medication as an intermediate.

Medications administered in different dose units are generally converted to a single dose unit (often **mg**) and the conversion values updated to reflect this however in some cases they may be kept in their administered unit (**e.g. mcg**) and the conversion shifted by an appropriate factor.


## Mathematical proofs

### Rearrangement of @eq-mu-ome-final {#sec-equation-proof}

The equation for the group mean OME provided @eq-mu-ome-1 is is a weighted sum of the OME values for patients $j$ in group $G$.

$$
\begin{aligned}
\text{OME}_j &= \sum_{i\in\mathbf{M}} d_i c_{i,r} \\
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j=G} \text{OME}_j
\end{aligned}
$${#eq-mu-ome-1}
First the the expression for $\text{OME}_j$ is substituted into the group mean. Then the order of summation is rearranged as both the sets $G$ and $\mathbf{M}$ are finite. Following this, the constant $c_{i,r}$ is factored out of the inner sum whilst the scaling term $\frac{1}{|G|}$ of group size is distributed to the inner term. 

$$
\begin{aligned}
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j \in G}  \left( \sum_{i\in\mathbf{M}} d_{j,i} c_{i,r} \right) \\
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j \in G}  \sum_{i\in\mathbf{M}} d_{j,i} c_{i,r} \\
\mu_{\text{OME}}^{(G)} &= \sum_{i\in\mathbf{M}} \left( \frac{1}{|G|} \sum_{j=1}  d_{j,i}  \right) c_{i,r}
\end{aligned}
$$

We can then define the mean dose $\bar{d}$ of medication $i$ as the arithmetic mean of doses for patients $j \in G$. 

$$
\bar{d_i} = \frac{1}{|G|} \sum_{j \in G} d_{j,i}
$$
This mean dose is then substituted back into the equation to yield the final rearrangement as seen in @eq-mu-ome-final. 

$$
\mu_{\text{OME}}^{(G)}  = \sum_{i\in\mathbf{M}} \bar{d_{i}} c_{i,r} 
$$


### Sensitivity of the arithmetic mean to differing proportions of $\bar{d_i}$ {#sec-ome-difference-proof}

:::{.callout-warning}
This section should be deleted.
:::

#### True treatment effect

Equation @eq-mu-ome-final outlines the calculation of a sample mean OME within a finite group. At the population level, using a simplified linear, confounder-free treatment effect, the expected OME of the treated population $\mathbb{E}[\text{OME} \mid T = 1]$ is modelled as a function of the baseline expectation $\mathbb{E}[\text{OME} \mid T = 0]$ and a standardized treatment effect ($t$) scaled by the baseline population standard deviation $\sigma_p$ (assuming homoscedasticity)

$$
\mathbb{E}[\text{OME} \mid T = 1] = \mathbb{E}[\text{OME} \mid T = 0] + t \cdot \sigma_{p}
$$

To estimate $t$ we will often measure the difference in the the observed mean of each group ($\mu_{\text{OME}}^{(T = 1)} - \mu_{\text{OME}}^{(T = 0)}$) divided by the variance $\sigma$.

$$
\begin{aligned}
\mu_{\text{OME}}^{(T = 1)} &= \mu_{\text{OME}}^{(T = 0)} + t \cdot \sigma_{p} \\[1em]
 t &= \frac{\mu_{\text{OME}}^{(T = 1)} - \mu_{\text{OME}}^{(T = 0)} }{\sigma_p} = \frac{\Delta \mu_{\text{OME}}}{\sigma_p} \\
\end{aligned}
$$ {#eq-t-estimator}

#### Conditional Component Definition

For any medication $i \in \mathbf{M}$ within a study cohort $G \in \{T=1, T=0\}$, the unconditional mean dose $\mu_{d_i}^{(G)}$  (as used in @eq-mu-ome-final ) can be decomposed as the product of the prescribing proportion $p_i$ and the conditional mean dose $\hat{d}_i$ amoung treated patients. 

$$
\bar{d}_i^{(G)} = p_i^{(G)} \cdot \hat{d}_i^{(G)}
$$ {#eq-conditional-mean}

where:

* $p_i^{(G)} = P(d_i > 0 \mid G)$ represents the proportion of patients in group $G$ receiving medication $i$.

* $\hat{d}_i^{(G)} = E[d_i \mid d_i > 0, G]$ represents the conditional mean dose among actively treated patients.

#### Group Difference Expansion

Substituting @eq-conditional-mean into @eq-mu-ome-final for groups $T=1$ and $T=0$:

$$
\Delta\mu_{\text{OME}} = \mu_{\text{OME}}^{(T=1)} - \mu_{\text{OME}}^{(T=0)} = \sum_{i\in\mathbf{M}} \left( p_i^{(T=1)}\hat{d}_i^{(T=1)} - p_i^{(T=0)}\hat{d}_i^{(T=0)} \right) c_{i,r}
$$ {#eq-expanded-difference}

#### Counterfactual Counter-Balancing

By adding and subtracting the cross-term $p_i^{(T=0)}\hat{d}_i^{(T=0)}$

$$
\Delta \mu_{\text{OME}} = \sum_{i\in\mathbf{M}} \left( p_i^{(T=1)}\hat{d}_i^{(T=1)} - p_i^{(T=1)}\hat{d}_i^{(T=0)} + p_i^{(T=1)}\hat{d}_i^{(T=0)} - p_i^{(T=0)}\hat{d}_i^{(T=0)} \right) c_{i,r}
$$ {#eq-algebraic-bridge}

#### Final Structural Decomposition

Factoring out the shared terms isolates the two independent drivers of macroscopic OME divergence:

$$
\Delta\mu_{\text{OME}} = \sum_{i\in\mathbf{M}} \underbrace{p_i^{(T=1)}\left(\hat{d}_i^{(T=1)} - \hat{d}_i^{(T=0)}\right)c_{i,r}}_{\text{treatment effect}} + \sum_{i\in\mathbf{M}} \underbrace{\hat{d}_i^{(T=0)}\left(p_i^{(T=1)} - p_i^{(T=0)}\right)c_{i,r}}_{\text{proportion effect}}
$$ {#eq-final-decomposition}
The first term reflects genuine dose-intensity change among treated patients: this is the component the estimator in @eq-t-estimator intends to recover as $t \cdot \sigma_p$. The second term arises purely from differing prescribing proportions $p_i^{(G)}$ between groups and is not attributable to treatment. This means that the the difference in group means $\Delta\mu_{\text{OME}}$ will only be equal to $t \cdot \sigma_p$ in the special case where the prescribing portions between groups are equal $p_i^{(T=1)} = p_i^{(T=0)}$, otherwise it only approximates the treatment effect. 



### Simulation study

:::{.callout-warning}
This section should be deleted.
:::

The simple simulation find that the observed difference is a hyperbolic parabola of $\Delta p_i$ and $\delta c_i$ with a scaling constant $k$.

$$
\begin{aligned}
\Delta \mu_{\text{OME}} &= k \cdot \Delta p_i \cdot  \epsilon \\
k &= \frac{\Delta \mu_{\text{OME}}}{\Delta p_i \cdot \epsilon }
\end{aligned}
$$ {#eq-results-simple-2-proof}

**Step 1: Expand $\Delta\mu_{\text{OME}}$ using @eq-expanded-difference.**

Substituting the group-difference expansion, where $\Delta p_i = p_i^{(T=1)} - p_i^{(T=0)}$:

$$
k = \frac{\displaystyle\sum_{i\in\mathbf{M}} \left( p_i^{(T=1)}\hat{d}_i^{(T=1)} - p_i^{(T=0)}\hat{d}_i^{(T=0)} \right) c_{i,r}}{\Delta p_i \cdot \delta c_i}
$$

**Step 2: Apply the null treatment effect assumption.**

In the simple simulation there is no true treatment effect, so conditional mean doses are equal across groups: $\hat{d}_i^{(T=1)} = \hat{d}_i^{(T=0)} = \hat{d}_i$. The treatment effect term in @eq-final-decomposition vanishes and only the proportion effect remains:

$$
\Delta\mu_{\text{OME}} = \sum_{i\in\mathbf{M}} \hat{d}_i \cdot \Delta p_i \cdot c_{i,r}
$$

**Step 3: Decompose the conversion factor into its true value and its error.**

Let $c_{i,r} = c_{i,r}^{\text{true}} + \delta c_i$, where $\delta c_i$ is the bias in the conversion value for medication $i$. For morphine ($i = M$), the reference drug: $c_M^{\text{true}} = 1$ and $\delta c_M = 0$ by definition (morphine is the OME reference and carries no conversion error). For oxycodone ($i = O$):

$$
c_{O,r} = c_{O,r}^{\text{true}} + \delta c_O
$$

**Step 4: Expand for the two-drug system $\mathbf{M} = \{M, O\}$.**

$$
\begin{aligned}
\Delta\mu_{\text{OME}} &= \hat{d}_M \cdot \Delta p_M \cdot c_M^{\text{true}} + \hat{d}_O \cdot \Delta p_O \cdot \left(c_{O,r}^{\text{true}} + \delta c_O\right) \\
&= \hat{d}_M \cdot \Delta p_M + \hat{d}_O \cdot \Delta p_O \cdot c_{O,r}^{\text{true}} + \hat{d}_O \cdot \Delta p_O \cdot \delta c_O
\end{aligned}
$$

**Step 5: Apply the proportional constraint.**

Since every patient receives exactly one medication, prescribing proportions within each group sum to unity. Therefore a change in oxycodone proportion is exactly offset by an equal and opposite change in morphine proportion: $\Delta p_M = -\Delta p_O$. Substituting:

$$
\begin{aligned}
\Delta\mu_{\text{OME}} &= -\hat{d}_M \cdot \Delta p_O + \hat{d}_O \cdot \Delta p_O \cdot c_{O,r}^{\text{true}} + \hat{d}_O \cdot \Delta p_O \cdot \delta c_O \\
&= \Delta p_O \underbrace{\left(\hat{d}_O \cdot c_{O,r}^{\text{true}} - \hat{d}_M \right)}_{\text{cancel under equal analgesic requirement}} + \hat{d}_O \cdot \Delta p_O \cdot \delta c_O
\end{aligned}
$$

**Step 6: Apply the equal analgesic requirement assumption.**

In the simple simulation, patients have the same underlying pain requirement regardless of which opioid is prescribed. Expressed in morphine equivalents, the conditional mean dose of morphine equals the conditional mean dose of oxycodone scaled by the true conversion factor:

$$
\hat{d}_M = \hat{d}_O \cdot c_{O,r}^{\text{true}}
$$

The bracketed term therefore cancels:

$$
\hat{d}_O \cdot c_{O,r}^{\text{true}} - \hat{d}_M = \hat{d}_O \cdot c_{O,r}^{\text{true}} - \hat{d}_O \cdot c_{O,r}^{\text{true}} = 0
$$

**Step 7: Collect the remaining term.**

$$
\Delta\mu_{\text{OME}} = \hat{d}_O \cdot \Delta p_O \cdot \delta c_O
$$

This confirms the empirically observed hyperbolic paraboloid form, where the spurious signal arises exclusively from the interaction of the proportion shift $\Delta p_i$ and the conversion bias $\delta c_i$.

**Step 8: Solve for $k$.**

$$
\begin{aligned}
k &= \frac{\hat{d}_O \cdot \Delta p_O \cdot \delta c_O}{\Delta p_O \cdot \delta c_O} \\[0.5em]
k &= \hat{d}_i
\end{aligned}
$$ {#eq-k-final}

The scaling constant is therefore the **conditional mean dose** of medication $i$ among the patients who receive it. Using @eq-conditional-mean, this can be expressed in terms of the unconditional mean dose $\bar{d}_i$ and prescribing proportion $p_i^{(G)}$:

$$
k = \hat{d}_i = \frac{\bar{d}_i^{(G)}}{p_i^{(G)}} = E\!\left[d_i \mid d_i > 0,\, G\right]
$$ {#eq-k-expanded}

And since $\hat{d}_M = \hat{d}_O \cdot c_{O,r}^{\text{true}}$ (Step 6), we can also express $k$ in terms of the underlying analgesic requirement $\hat{d}_M$ and the true potency of oxycodone via @eq-potency:

$$
k = \frac{\hat{d}_M}{c_{O,r}^{\text{true}}} = \hat{d}_M \cdot \rho_{O,r}^{\text{true}}
$$ {#eq-k-potency}

### Clinical Significance

* **Associational difference:** Quantifies the true statistical association by isolating shifts in the active treatment magnitude while holding the selection rate constant ($p_i^{(A)}$).
* **Medication proportion Effect:** Quantifies the structural selection bias. A shift in the prevalence of medication usage ($p_i^{(A)} \neq p_i^{(B)}$) alters $\mu_{\text{OME}}$ linearly via $c_{i,r}$, even if the active proportions of medications are perfectly homogeneous ($\bar{d}_i^{(A)} = \bar{d}_i^{(B)}$).

## Literature search

### PUBMED search terms {#sec-appendix-pubmed-1}

On 14th of July 2026, a search was performed on the PUBMED online search tool using the following search term to identify the volume of literature using opioid equivalence metrics in research. The search term was "(Milligram morphine equivalent) OR (Oral morphine equivalent)" with a 10 year filter was used and yielded the following full search string. There was 4,031 results. 

::: {.callout-note appearance="simple" title = "PUBMED search term"}
((("milligram"[All Fields] OR "milligrams"[All Fields]) AND ("morphine derivatives"[Supplementary Concept] OR "morphine derivatives"[All Fields] OR "morphines"[All Fields] OR "morphine derivatives"[MeSH Terms] OR ("morphine"[All Fields] AND "derivatives"[All Fields]) OR "morphine"[Supplementary Concept] OR "morphine"[All Fields] OR "morphin"[All Fields] OR "morphine"[MeSH Terms] OR "morphine s"[All Fields] OR "morphinic"[All Fields]) AND ("equivalence"[All Fields] OR "equivalences"[All Fields] OR "equivalencies"[All Fields] OR "equivalency"[All Fields] OR "equivalent"[All Fields] OR "equivalently"[All Fields] OR "equivalents"[All Fields])) OR (("mouth"[MeSH Terms] OR "mouth"[All Fields] OR "oral"[All Fields]) AND ("morphine derivatives"[Supplementary Concept] OR "morphine derivatives"[All Fields] OR "morphines"[All Fields] OR "morphine derivatives"[MeSH Terms] OR ("morphine"[All Fields] AND "derivatives"[All Fields]) OR "morphine"[Supplementary Concept] OR "morphine"[All Fields] OR "morphin"[All Fields] OR "morphine"[MeSH Terms] OR "morphine s"[All Fields] OR "morphinic"[All Fields]) AND ("equivalence"[All Fields] OR "equivalences"[All Fields] OR "equivalencies"[All Fields] OR "equivalency"[All Fields] OR "equivalent"[All Fields] OR "equivalently"[All Fields] OR "equivalents"[All Fields]))) AND (y_10[Filter])
:::


### PUBMED search terms {#sec-appendix-pubmed}

The search term was executed on the PUBMED online search tool on 7th July 2026. 

::: {.callout-note appearance="simple" title = "PUBMED search term"}
(((((("opioid equivalent" OR "oral morphine equivalence" OR "oral morphine equivalent" OR "oral morphine equivalents" OR "OME" OR "morphine milligram equivalents" OR "morphine milligram equivalence" OR "MME" OR "morphine equivalent daily dose" OR "MEDD") AND ("perioperative" OR "peri-operative" OR "postoperative" OR "post-operative" OR "surgical" OR "surgery" OR "post-surgical" OR "postsurgical"))) AND ("2025/01/01"[Date - Publication] : "2025/12/31"[Date - Publication]))) AND ("clinical trial"[Filter] OR "clinical trial" OR "controlled clinical trial" OR "randomized controlled trial" OR "interventional" OR "intervention study" OR "clinical trials as topic"[MeSH Terms] OR "cohort studies"[MeSH Terms] OR "cohort" OR "prospective" OR "retrospective" OR "case-control" OR "comparative study" OR "evaluation study"))
:::

### Covidence extraction form {#sec-appendix-cov-extraction}

Data was extracted from selected papers using the following proforma by a single author. Authors could leave notes to highlight specific features for later analysis required.

```{r}
#| label: tbl-extraction-form
#| tbl-cap: "Litrature extraction proforma"
#| output: TRUE

tbl_extraction_form <- data.frame("Variable" = c("Study type",
                          "Opioid conversion set",
                          "Intervention",
                          "Opioid proportions reported between intervetnion groups",
                          "Outcome",
                          "Data avalibility"),
           "Options" = c("Randomised controlled trial, Cohort study, Case controlled study, Other",
                         "Cited, Included in paper, Not described",
                         "Surgical intervention, Anaesthetic intervnetion, Other",
                         "Described or Not described",
                         "Difference in opioid requirement/treatment, No difference",
                         "Publically avalible, Not publically avaliable, Database study"),
           "Explanatory text" = c("What is the format of the study?",
                                  "In the methods section, is how is the OME metric generated described and how are the conversion values referenced?",
                                  "What type of intervention is given. For pre/post-operative interventions that include changes to medication select anaesthesia. For descriptive studies analysing differences select other.",
                                  "Assess the results section, Is it possible to establish the proprotion, mean or total doses of each opioid between intervention groups. If authros explicilty state only one type of opioid used in the OME/MME metric select 'Described'.",
                                  "In the primary outcome of OME/MME, is there a significant difference reported between intervention groups.",
                                  "Is data avalible for readers to ascertain the proportions of opioids between groups? If OME/MME metric only avaliable then select 'Not avalible'."))

knitr::kable(tbl_extraction_form,
    col.names = c("Variable", "Options", "Explanatory text")
)

```

### Grid search of plausible conversion values to studies {#sec-appendix-neilsen-ranges}

The values of the possible plausible conversion values are taken from Table 1 of Neilsen's paper on a unified conversion value set. These conversion values are sourced from national clinical guideline documents from a range of countries. 

Medication's were assumed to be administered IV with the exception of Tramadol and Hydrocodone as Neilsen only reports oral conversion values for these. 

```{r ranges}

neilsen_ranges_table <- data.frame(
  Medication = names(neilsen_ranges),
  `Conversion values` = sapply(neilsen_ranges, paste, collapse = ", "),
  check.names = FALSE,
  row.names = NULL
)

knitr::kable(neilsen_ranges_table)
```


```{r}
#| label: fig-lit-outcome-dist
#| fig-cap: "Density plot of the possible treatment effects calculated de-novo across the grid values for opioid conversion values."
#| eval: false

### EVAL IS FALSE ###

library(ggplot2)

# 2. Plot with facets
ggplot(plot_data, aes(x = delta_ome)) +
  # Use geom_density for a smooth distribution
  geom_density(fill = "steelblue", alpha = 0.6, color = "black") +
  # Use facet_wrap with free scales so each study's range is visible
  facet_wrap(~ authors, scales = "free", ncol = 3) + 
  theme_minimal(base_size = 12) +
  labs(
    title = "Sensitivity Analysis: OME Relative Difference by Study",
    subtitle = "Variation based on independent uncertainty in conversion factors (c_i)",
    x = "Relative Difference (delta_ome)",
    y = "Density"
  ) +
  theme(
    strip.background = element_rect(fill = "gray95"),
    panel.spacing = unit(1, "lines")
  )

```


## Simulation study

### Hypothetical study simulation {#sec-appendix-sim-design}

```{r}

t_0 = as.Date("2020-01-01")
t_end = as.Date("2025-12-31")
t_robot = as.Date("2023-01-01")
study_length_years <- as.numeric(difftime(t_end, t_0,  unit="weeks"))/52.25
robot_length_years <- as.numeric(difftime(t_end, t_robot,  unit="weeks"))/52.25

case_volume <- 100
total_study_population <- case_volume * study_length_years
proportion_robotic <- 0.8
total_robotic <- robot_length_years * case_volume * proportion_robotic
total_non_robotic = total_study_population - total_robotic

proportion_morphine = 0.7
oxycodone_shortage = c(as.Date("2024-01-01"), as.Date("2024-12-31"))

mu_tot = 50
sd_tot = 10
potency_oxycodone_truth = 1.85
potency_oxycodone_ome = 2

robot_effect_size = 0.8

mu_robotic = mu_tot * robot_effect_size
mean_diff = mu_robotic - mu_tot
mean_diff_sd = mean_diff/sd_tot

```

:::{.callout-warning}
Ensure variables match main mauscript values
:::

A research team wish to understand if the use of a robot for a surgical procedure, reduces the post-operative pain.

The average case volume is `r case_volume` procedures per year, these were all performed using non-robotic surgery until `r format(t_robot, "%e %b %y")` when a robot was acquired, and since `r scales::percent(proportion_robotic)` of cases are performed with the robot. All EHR data was collected between `r format(t_0, "%e %b %y")` and `r format(t_end, "%e %b %y")`.

All patients are treated with a Patient Controlled Analgesia (PCA) after their operation and on average the anaesthetist will prescribe a morphine PCA to `r scales::percent(proportion_morphine)` of patients and an oxycodone PCA to the remainder. No other opioid medication was administered. 

They design a retrospective cohort study for patients treated with a robotic procedure and those without. The primary outcome is the total opioid requirement, as measured in OME mg, in the first 24 hours post-operatively as a surrogate for the total pain caused by the surgical procedure.

During the study there was a national shortage of oxycodone from `r oxycodone_shortage[1]` to `r oxycodone_shortage[2]` and at which time all patients were prescribed a morphine PCA. Whilst the authors described this in their methods section, they did feel it would impact their results as the differing opioids were already accounted for in their OME metric. 

### Effect of $\mu_{\text{OME}}^{\text{(control)}}$ on simple simulation {#sec-appendix-sim-1}

To investigate the scaling term of $\mu_{\text{OME}}^{\text{(control)}}$ on the observed treatment effect, 6 further simulations were performed, each as described in @sec-manuscript-simple-sim but with a different $\mu_{\text{OME}}^{\text{(control)}}$ between 0 OME mg and 10 OME mg. The treatment effect was fixed at 0. 

```{r}
#| echo: false
#| message: false
#| fig-width: 6
#| fig-height: 6

library(ggplot2)
library(dplyr)
library(latex2exp)

p_min <- -1
p_max <- 1
e_min <- -1
e_max <- 1
sigma = 0.1

if (PRODUCTION_QUALITY) {
  n = 10000
  space = 0.05
} else {
  n = 1000
  space = 0.1
}

simple_results_mu <- data.frame(p = numeric(), e = numeric(), diff = numeric(), mu=numeric())
for(mu in seq(0, 10, length.out = 6)){
for(p in seq(p_min, p_max, by = space)){
  for(e in seq(e_min, e_max, by = space)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = 0
    )
    
    simple_results_mu <- rbind(simple_results_mu, data.frame(p = p, e = e, diff = diff, mu=mu))
  }
}
}
```

```{r}
#| label: fig-appendix-simple-sim-mu
#| fig-title: Effect of scale of the control group average dose (OME mg) on observed effect size
#| fig-cap: Plots of the difference in proportion over the error in conversion values provided faceted by the expected mean dose of the control group. Colour scale demonstrates the observed difference in the treatment group. True effect size is set to 0. 
#| fig-width: 6
#| fig-height: 8

label_function <- function(val) {
  paste0('mu["OME"] == ', val)
}

plot_simple_grid_mu <- ggplot(simple_results_mu, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)'),
       caption = TeX(r'(True $\delta \mu_{{OME}} = 0$ )')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
symmetric_fill_scale(30)+
  facet_wrap(~mu, labeller = as_labeller(label_function, default = label_parsed)
             )
  
plot_simple_grid_mu

```

### Effect of $\Delta \mu_{\text{OME}}$ on simple simulation {#sec-appendix-sim-2}

To investigate the interaction of a true treatment effect on the observed treatment effect, 9 further simulations were performed, each as described in @sec-manuscript-simple-sim but with a different true treatment effect between the values of -1 and 1. 

```{r}

if (PRODUCTION_QUALITY) {
  n = 10000
  space = 0.05
} else {
  n = 1000
  space = 0.1
}

simple_results_t <- data.frame(p = numeric(), e = numeric(), diff = numeric(), t=numeric())
for(t in seq(-1, 1, length.out = 9)){
for(p in seq(p_min, p_max, by = space)){
  for(e in seq(e_min, e_max, by = space)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = 10,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = t
    )
    
    simple_results_t <- rbind(simple_results_t, data.frame(p = p, e = e, diff = diff, t=t))
  }
}
}


```

```{r}
#| label: fig-appendix-simple-sim-t
#| fig-title: Effect of differing true effect size on observed effect size
#| fig-cap: Plots of the difference in proportion over the error in conversion values provided faceted by the expected mean dose of the control group. Colour scale demonstrates the observed difference in the treatment group. The mean value of the control group is set to 10 OME mg . 
#| ig-width: 6
#| fig-height: 8

label_function <- function(val) {
  paste0('delta["OME"] == ', val)
}

plot_simple_grid_t <- ggplot(simple_results_t, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)'),
       caption  = TeX(r'($\mu_{{OME}}^{(control)}$ (Mean dose of control group) $= 10$ )')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
symmetric_fill_scale(30)+
  facet_wrap(~t, labeller = as_labeller(label_function, default = label_parsed)
             )
  
plot_simple_grid_t
```

### Power calculation {#sec-appendix-simulation-power}

In the first part of the experiment a large effect size is used.

For a study period of `r round(study_length_years,1)` years and a annual case volume of `r case_volume` the total number of expected study participants is `r total_study_population`, of which we would expect `r round(total_robotic,1)` to be robotic cases and `r round(total_non_robotic,1)`  given a `r round(robot_length_years,1)` year period where `r scales::percent(proportion_robotic)` of cases are performed robotically.

Assuming a `r mu_tot` mg (SD: `r sd_tot` mg) mean opioid requirement for the non-robotic population, we can calculate the expected mean OME requirement for robotic and non-robotic cases given a `r robot_effect_size` effect size. The expected mean OME requirement for robotic cases is `r mu_robotic` mg giving a mean difference of `r  mean_diff` mg or `r mean_diff_sd` standard deviations.

```{r}
library(pwr)
power_exp_1 <- pwr.t2n.test(n1 = total_non_robotic, 
             n2 = total_robotic, 
             d = mean_diff_sd, 
             sig.level = 0.05, 
             alternative = "two.sided")

```
Given the initial experimental set up it is clear that the study is very highly powered with the power value to be calculated at `r scales::percent(power_exp_1$power)`.

@fig-power-curve demonstrated the power curve for the comparison of OME between robotic and non-robotic surgical groups. The dashed lines indicate the minimum detectable effect size at 80% power.

```{r}
#| label: fig-power-curve
#| fig-cap: "Power curve for the comparison of OME between robotic and non-robotic surgical groups. The dashed lines indicate the minimum detectable effect size at 80% power."

library(pwr)
library(ggplot2)

d_seq <- seq(0.05, 1.0, by = 0.005)

power_seq <- sapply(d_seq, function(d) {
  pwr.t2n.test(n1 = total_non_robotic, n2 = total_robotic,
               d = d, sig.level = 0.05,
               alternative = "two.sided")$power
})

mdd <- pwr.t2n.test(n1 = total_non_robotic, n2 = total_robotic,
                    sig.level = 0.05, power = 0.8,
                    alternative = "two.sided")$d
mdd_mg = sd_tot * mdd

power_df <- data.frame(d = d_seq, power = power_seq)

ggplot(power_df, aes(x = d, y = power)) +
  geom_line() +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = mdd, linetype = "dashed", colour = "grey40") +
  annotate("text", x = mdd + 0.03, y = 0.05,
           label = paste0("MDD = ", round(mdd_mg, 2),"mg \n(SD =" , round(mdd, 3), ")"),
           hjust = 0, size = 3.5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent_format()) +
  labs(x = "Cohen's *d* (standardised effect size)",
       y = "Power") +
  theme_classic() +
  theme(axis.title.x = ggtext::element_markdown())
```

With the observed sample sizes, the minimum detectable effect size at 80% power is 
*d* = `r round(mdd, 3)`, which equates to a mean difference of `r round(mdd_mg, 2)` mg OME.


::: {.content-visible when-format="pdf"}
\listoffigures
:::
