---
output: html_document
editor_options: 
  chunk_output_type: console
params:
  quality: "low"
---
# Supplementary materials
```{r echo=FALSE, message=FALSE}
# Source the founctions in simulaton.qmd with a temporary R script containing only code blocks
temp_r_file <- tempfile(fileext = ".R")
knitr::purl("simulation.qmd", output = temp_r_file, quiet = TRUE)
knitr::purl("index.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
unlink(temp_r_file)
rm(temp_r_file)
```

## Opioid equivelence calculations in practice

**Question**: What dose of intravenous fentanyl will provide equal analgesic benefit to 5 mg of intravenous morphine, assuming IV fentanyl is 100 times more potent.

$$\begin{align}
\rho_{\text{fentanyl,IV}} = 100 \\
c_{\text{fentanyl,IV}} = \frac{1}{\rho_{\text{fentanyl,IV}}} \\
c_{\text{fentanyl,IV}} = \frac{1}{100} = 0.01 \\
\end{align}$$

First we use the potency value to find the conversion value to be 0.01.

$$\begin{align}
d_{\text{morphine}} \times c_{\text{fentanyl,IV}} = d_{\text{fentanyl}} \\
5 \times 0.01 = 0.05\text{mg} = 50\text{mcg}
\end{align}$$

Then we multiple our reference dose of 5mg of morphine by the conversion value to find the equivalent dose of fentanyl to be 0.05 mg or 50 mcg.

This approach can also be used to convert from two non-reference medications by firstly converting to the reference (often oral morphine) and then to the target medication.





## Mathematical proofs

### Proof for @eq-mu-ome-final {#sec-equation-proof}
 @eq-mu-ome-1 can then be substituted into the the arithmetic mean to yield @eq-mu-ome-2.
$$
\mu = \frac{1}{N} \sum^N_{j=1}  \left( \sum_{i\in\mathbf{M}} d_{j,i} c_{i,r} \right)
$${#eq-mu-ome-2}
As finite summation is linear, we sun cam across the medications $i$ and then across the patients $p$. As the conversion vector $c_{i,r}$ is constant with respect to $j$ it can be factored out to yeild @eq-mu-ome-3
$$
\mu = \sum_{i\in\mathbf{M}} \left( \frac{1}{N} \sum^N_{j=1}  d_{j,i}  \right) c_{i,r}
$${#eq-mu-ome-3}


### Proof to demonstrate the sensitivity of the arthimetic mean to proportions {#sec-ome-difference-proof}

To describe how the variance between group means ($\mu_{\text{OME}}^{(A)} - \mu_{\text{OME}}^{(B)}$) is driven by both the statistical association between treatment groups (Associational difference)^[This is not a causal quantitiy but rather the combination of the causal effect of the treatement on the OME metric combined with the statical association of the treatment on the outcome] and composition of the medications within each group (medication proportion), we employed a structural decomposition.

#### Conditional Component Definition

For any medication $i \in \mathbf{M}$ within a study cohort $G \in \{A, B\}$, the unconditional mean dose $\mu_{d_i}^{(G)}$ is defined as the product of the prescription probability and the expected dose:

$$
\mu_{d_i}^{(G)} = p_i^{(G)} \cdot \bar{d}_i^{(G)}
$$ {#eq-conditional-mean}

where:

* $p_i^{(G)} = P(d_i > 0 \mid G)$ represents the proportion of patients in group $G$ receiving medication $i$.

* $\bar{d}_i^{(G)} = E[d_i \mid d_i > 0, G]$ represents the conditional mean dose among actively treated patients.

#### Group Difference Expansion

Substituting @eq-conditional-mean into the aggregate group difference formula yields:

$$
\mu_{\text{OME}}^{(A)} - \mu_{\text{OME}}^{(B)} = \sum_{i\in\mathbf{M}} \left( p_i^{(A)}\bar{d}_i^{(A)} - p_i^{(B)}\bar{d}_i^{(B)} \right) c_{i,r}
$$ {#eq-expanded-difference}

#### Counterfactual Counter-Balancing

By adding and subtracting the counterfactual cross-term $p_i^{(A)}\bar{d}_i^{(B)}c_{i,r}$ within the summation, we map the trajectory between the two distinct states:

$$
\mu_{\text{OME}}^{(A)} - \mu_{\text{OME}}^{(B)} = \sum_{i\in\mathbf{M}} \left( p_i^{(A)}\bar{d}_i^{(A)} - p_i^{(A)}\bar{d}_i^{(B)} + p_i^{(A)}\bar{d}_i^{(B)} - p_i^{(B)}\bar{d}_i^{(B)} \right) c_{i,r}
$$ {#eq-algebraic-bridge}

#### Final Structural Decomposition

Factoring out the shared terms isolates the two independent drivers of macroscopic OME divergence:

$$
\mu_{\text{OME}}^{(A)} - \mu_{\text{OME}}^{(B)} = \sum_{i\in\mathbf{M}} \underbrace{p_i^{(A)}\left(\bar{d}_i^{(A)} - \bar{d}_i^{(B)}\right)c_{i,r}}_{\text{Associational difference}} + \sum_{i\in\mathbf{M}} \underbrace{\bar{d}_i^{(B)}\left(p_i^{(A)} - p_i^{(B)}\right)c_{i,r}}_{\text{Medication proportion}}
$$ {#eq-final-decomposition}

### Simulation study

$$
\begin{aligned}
\Delta \mu_{\text{OME}} = -1 \cdot \Delta p_i \cdot \delta c_i \\
\Delta \mu_{\text{OME}} = -\mu_{\text{OME}}^\text{(control)} \cdot \Delta p_i \cdot \delta c_i \\
\\
p_i^{\text{(control)}} - p_i^{\text{(intervention)}} 
\\

\mu_{\text{OME}}^{\text{(control)}} - \mu_{\text{OME}}^{\text{(intervention)}} = -\mu_{\text{OME}}^\text{(control)} \cdot \Delta p_i \cdot \delta c_i \\
\mu^\text{intervention} = \mu^\text{control}[ 1 + \Delta p_i \cdot \delta c_i]
\end{aligned}
$$ {#eq-results-simple-2-proof}

### Clinical Significance

* **Associational difference:** Quantifies the true statistical association by isolating shifts in the active treatment magnitude while holding the selection rate constant ($p_i^{(A)}$).
* **Medication proportion Effect:** Quantifies the structural selection bias. A shift in the prevalence of medication usage ($p_i^{(A)} \neq p_i^{(B)}$) alters $\mu_{\text{OME}}$ linearly via $c_{i,r}$, even if the active proportions of medications are perfectly homogeneous ($\bar{d}_i^{(A)} = \bar{d}_i^{(B)}$).

## Literature search

## PUBMED search terms {#sec-appendix-pubmed-1}

On 14th of July 2026, a search was performed on the PUBMED online search tool using the following search term to identify the volume of literature using opioid equivalence metrics in research. The search term was "(Milligram morphine equivalent) OR (Oral morphine equivalent)" with a 10 year filter was used and yielded the following full search string. There was 4,031 results. 

::: {.callout-note appearance="simple" title = "PUBMED search term"}
((("milligram"[All Fields] OR "milligrams"[All Fields]) AND ("morphine derivatives"[Supplementary Concept] OR "morphine derivatives"[All Fields] OR "morphines"[All Fields] OR "morphine derivatives"[MeSH Terms] OR ("morphine"[All Fields] AND "derivatives"[All Fields]) OR "morphine"[Supplementary Concept] OR "morphine"[All Fields] OR "morphin"[All Fields] OR "morphine"[MeSH Terms] OR "morphine s"[All Fields] OR "morphinic"[All Fields]) AND ("equivalence"[All Fields] OR "equivalences"[All Fields] OR "equivalencies"[All Fields] OR "equivalency"[All Fields] OR "equivalent"[All Fields] OR "equivalently"[All Fields] OR "equivalents"[All Fields])) OR (("mouth"[MeSH Terms] OR "mouth"[All Fields] OR "oral"[All Fields]) AND ("morphine derivatives"[Supplementary Concept] OR "morphine derivatives"[All Fields] OR "morphines"[All Fields] OR "morphine derivatives"[MeSH Terms] OR ("morphine"[All Fields] AND "derivatives"[All Fields]) OR "morphine"[Supplementary Concept] OR "morphine"[All Fields] OR "morphin"[All Fields] OR "morphine"[MeSH Terms] OR "morphine s"[All Fields] OR "morphinic"[All Fields]) AND ("equivalence"[All Fields] OR "equivalences"[All Fields] OR "equivalencies"[All Fields] OR "equivalency"[All Fields] OR "equivalent"[All Fields] OR "equivalently"[All Fields] OR "equivalents"[All Fields]))) AND (y_10[Filter])
:::


### PUBMED search terms {#sec-appendix-pubmed}

The search term was executed on the PUBMED online search tool on 7th July 2026. 

::: {.callout-note appearance="simple" title = "PUBMED search term"}
(((((("opioid equivalent" OR "oral morphine equivalence" OR "oral morphine equivalent" OR "oral morphine equivalents" OR "OME" OR "morphine milligram equivalents" OR "morphine milligram equivalence" OR "MME" OR "morphine equivalent daily dose" OR "MEDD") AND ("perioperative" OR "peri-operative" OR "postoperative" OR "post-operative" OR "surgical" OR "surgery" OR "post-surgical" OR "postsurgical"))) AND ("2025/01/01"[Date - Publication] : "2025/12/31"[Date - Publication]))) AND ("clinical trial"[Filter] OR "clinical trial" OR "controlled clinical trial" OR "randomized controlled trial" OR "interventional" OR "intervention study" OR "clinical trials as topic"[MeSH Terms] OR "cohort studies"[MeSH Terms] OR "cohort" OR "prospective" OR "retrospective" OR "case-control" OR "comparative study" OR "evaluation study"))
:::

### Covidence extraction form {#sec-appendix-cov-extraction}

Data was extracted from selected papers using the following proforma by a single author. Authors could leave notes to highlight specific features for later analysis required.

```{r}
#| label: tbl-extraction-form
#| tbl-cap: "Litrature extraction proforma"
#| output: TRUE

tbl_extraction_form <- data.frame("Variable" = c("Study type",
                          "Opioid conversion set",
                          "Intervention",
                          "Opioid proportions reported between intervetnion groups",
                          "Outcome",
                          "Data avalibility"),
           "Options" = c("Randomised controlled trial, Cohort study, Case controlled study, Other",
                         "Cited, Included in paper, Not described",
                         "Surgical intervention, Anaesthetic intervnetion, Other",
                         "Described or Not described",
                         "Difference in opioid requirement/treatment, No difference",
                         "Publically avalible, Not publically avaliable, Database study"),
           "Explanatory text" = c("What is the format of the study?",
                                  "In the methods section, is how is the OME metric generated described and how are the conversion values referenced?",
                                  "What type of intervention is given. For pre/post-operative interventions that include changes to medication select anaesthesia. For descriptive studies analysing differences select other.",
                                  "Assess the results section, Is it possible to establish the proprotion, mean or total doses of each opioid between intervention groups. If authros explicilty state only one type of opioid used in the OME/MME metric select 'Described'.",
                                  "In the primary outcome of OME/MME, is there a significant difference reported between intervention groups.",
                                  "Is data avalible for readers to ascertain the proportions of opioids between groups? If OME/MME metric only avaliable then select 'Not avalible'."))

knitr::kable(tbl_extraction_form,
    col.names = c("Variable", "Options", "Explanatory text")
)

```

## Simulation study

### Effect of $\mu_{\text{OME}}^{\text{(control)}}$ on simple simulation {#sec-appendix-sim-1}

To investigate the scaling term of $\mu_{\text{OME}}^{\text{(control)}}$ on the observed treatment effect, 6 further simulations were performed, each as described in @sec-manuscript-simple-sim but with a different $\mu_{\text{OME}}^{\text{(control)}}$ between 0 OME mg and 10 OME mg. The treatment effect was fixed at 0. 

```{r}

library(ggplot2)
library(dplyr)
library(latex2exp)

p_min <- -1
p_max <- 1
e_min <- -1
e_max <- 1
sigma = 0.1

if (PRODUCTION_QUALITY) {
  n = 10000
  space = 0.05
} else {
  n = 1000
  space = 0.2
}

simple_results_mu <- data.frame(p = numeric(), e = numeric(), diff = numeric(), mu=numeric())
for(mu in seq(0, 10, length.out = 6)){
for(p in seq(p_min, p_max, by = space)){
  for(e in seq(e_min, e_max, by = space)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = 0
    )
    
    simple_results_mu <- rbind(simple_results_mu, data.frame(p = p, e = e, diff = diff, mu=mu))
  }
}
}
```

```{r}
#| label: fig-appendix-simple-sim-mu
#| fig-title: Effect of scale of the control group average dose (OME mg) on observed effect size
#| fig-cap: Plots of the difference in proportion over the error in conversion values provided faceted by the expected mean dose of the control group. Colour scale demonstrates the observed difference in the treatment group. True effect size is set to 0. 

label_function <- function(val) {
  paste0('mu["OME"]^"(control)" == ', val )
}

symmetric_fill_scale <- function(limit, type="fill") {
  
  if(type=="fill"){
  scale_fill_gradientn(
    colours = c("darkblue", "blue", "dodgerblue", "white", "orange", "red", "darkred"),
    values  = scales::rescale(c(-limit, -limit/2, 0, limit/2, limit)),
    limits  = c(-limit, limit)
  )
    }else{
    scale_colour_gradientn(
    colours = c("darkblue", "blue", "dodgerblue", "white", "orange", "red", "darkred"),
    values  = scales::rescale(c(-limit, -limit/2, 0, limit/2, limit)),
    limits  = c(-limit, limit)
  )
  }
}

plot_simple_grid_mu <- ggplot(simple_results_mu, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)'),
       caption = TeX(r'(True $\delta \mu_{{OME}} = 0$ )')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
symmetric_fill_scale(30)+
  facet_wrap(~mu, labeller = as_labeller(label_function, default = label_parsed)
             )
  
plot_simple_grid_mu

```

### Effect of $\Delta \mu_{\text{OME}}$ on simple simulation {#sec-appendix-sim-2}

To investigate the interaction of a true treatment effect on the observed treatment effect, 9 further simulations were performed, each as described in @sec-manuscript-simple-sim but with a different true treatment effect between the values of -1 and 1. 

```{r}

if (PRODUCTION_QUALITY) {
  n = 10000
  space = 0.05
} else {
  n = 1000
  space = 0.2
}

simple_results_t <- data.frame(p = numeric(), e = numeric(), diff = numeric(), t=numeric())
for(t in seq(-1, 1, length.out = 9)){
for(p in seq(p_min, p_max, by = space)){
  for(e in seq(e_min, e_max, by = space)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = 10,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = t
    )
    
    simple_results_t <- rbind(simple_results_t, data.frame(p = p, e = e, diff = diff, t=t))
  }
}
}


```

```{r}
#| label: fig-appendix-simple-sim-t
#| fig-title: Effect of differing true effect size on observed effect size
#| fig-cap: Plots of the difference in proportion over the error in conversion values provided faceted by the expected mean dose of the control group. Colour scale demonstrates the observed difference in the treatment group. The mean value of the control group is set to 10 OME mg . 

label_function <- function(val) {
  paste0('delta["OME"] == ', val)
}

plot_simple_grid_t <- ggplot(simple_results_t, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)'),
       caption  = TeX(r'($\mu_{{OME}}^{(control)}$ (Mean dose of control group) $= 10$ )')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
symmetric_fill_scale(30)+
  facet_wrap(~t, labeller = as_labeller(label_function, default = label_parsed)
             )
  
plot_simple_grid_t
```

### Power calculation {#sec-appendix-simulation-power}

In the first part of the experiment a large effect size is used.

```{r}

t_0 = as.Date("2020-01-01")
t_end = as.Date("2025-12-31")
t_robot = as.Date("2023-01-01")
study_length_years <- as.numeric(difftime(t_end, t_0,  unit="weeks"))/52.25
robot_length_years <- as.numeric(difftime(t_end, t_robot,  unit="weeks"))/52.25

case_volume <- 100
total_study_population <- case_volume * study_length_years
proportion_robotic <- 0.8
total_robotic <- robot_length_years * case_volume * proportion_robotic
total_non_robotic = total_study_population - total_robotic

proportion_morphine = 0.7
oxycodone_shortage = c(as.Date("2024-01-01"), as.Date("2024-12-31"))

mu_tot = 50
sd_tot = 10
potency_oxycodone_truth = 1.85
potency_oxycodone_ome = 2

robot_effect_size = 0.8

mu_robotic = mu_tot * robot_effect_size
mean_diff = mu_robotic - mu_tot
mean_diff_sd = mean_diff/sd_tot

```

For a study period of `r study_length_years` years and a annual case volume of `r case_volume` the total number of expected study participants is `r total_study_population`, of which we would expect `r total_robotic` to be robotic cases and `r total_non_robotic`  given a `r robot_length_years`year period where `r scales::percent(proportion_robotic)` of cases are performed robotically.

Assuming a `r mu_tot` mg (SD: `r sd_tot` mg) mean opioid requirement for the non-robotic population, we can calculate the expected mean OME requirement for robotic and non-robotic cases given a `r robot_effect_size` effect size. The expected mean OME requirement for robotic cases is `r mu_robotic` mg giving a mean difference of `r  mean_diff` mg or `r mean_diff_sd` standard deviations.

```{r}
library(pwr)
power_exp_1 <- pwr.t2n.test(n1 = total_non_robotic, 
             n2 = total_robotic, 
             d = mean_diff_sd, 
             sig.level = 0.05, 
             alternative = "two.sided")

```
Given the inital experimental set up it is clear that the study is very highly powered with the power value to be calculated at `r scales::percent(power_exp_1$power)`.

@fig-power-curve demonstrated the power curve for the comparison of OME between robotic and non-robotic surgical groups. The dashed lines indicate the minimum detectable effect size at 80% power.

```{r}
#| label: fig-power-curve
#| fig-cap: "Power curve for the comparison of OME between robotic and non-robotic surgical groups. The dashed lines indicate the minimum detectable effect size at 80% power."

library(pwr)
library(ggplot2)

d_seq <- seq(0.05, 1.0, by = 0.005)

power_seq <- sapply(d_seq, function(d) {
  pwr.t2n.test(n1 = total_non_robotic, n2 = total_robotic,
               d = d, sig.level = 0.05,
               alternative = "two.sided")$power
})

mdd <- pwr.t2n.test(n1 = total_non_robotic, n2 = total_robotic,
                    sig.level = 0.05, power = 0.8,
                    alternative = "two.sided")$d
mdd_mg = sd_tot * mdd

power_df <- data.frame(d = d_seq, power = power_seq)

ggplot(power_df, aes(x = d, y = power)) +
  geom_line() +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = mdd, linetype = "dashed", colour = "grey40") +
  annotate("text", x = mdd + 0.03, y = 0.05,
           label = paste0("MDD = ", round(mdd_mg, 2),"mg \n(SD =" , round(mdd, 3), ")"),
           hjust = 0, size = 3.5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2),
                     labels = scales::percent_format()) +
  labs(x = "Cohen's *d* (standardised effect size)",
       y = "Power (1 − β)") +
  theme_classic() +
  theme(axis.title.x = ggtext::element_markdown())
```

With the observed sample sizes, the minimum detectable effect size at 80% power is 
*d* = `r round(mdd, 3)`, which equates to a mean difference of `r round(mdd_mg, 2)` mg OME.

The effect size is use to inform the weakly powered study in 

```
