---
format: html
---

# Manuscript

```{r source_files}
#| echo: false
#| message: false

source("functions.R", echo = FALSE)

shared_objects <- readRDS("data/shared_objects_sim.rds")
plot_simple_grid_2 =shared_objects$plot_simple_grid_2
plot_simple_grid = shared_objects$plot_simple_grid
plot_grid_2 = shared_objects$plot_grid_2
bounds=shared_objects$bounds

shared_objects_lit <- readRDS("data/shared_objects_lit.rds")
n_lit_data <- shared_objects_lit$n_lit_data
lit_not_described <- shared_objects_lit$lit_not_described
lit_provide_doses <- shared_objects_lit$lit_provide_doses
plot_point_whisker <- shared_objects_lit$plot_point_whisker
```

## Introduction

There is a wide variety of opioids used across the range of clinical specialities and patient groups, each with their own potency to provide analgesia @bhatnagarOpioidEquivalency2026. Opioid conversion values offer the ability to convert a specific opioid into the equivalent units of a reference opioid, to produce equi-analgesic doses measured in oral morphine equivalent milligrams (OME mg) or milligram morphine equivalent (MME).  

We define the total opioid burden (TOB) to be the weighted sum of the dose $d$ for medication $i$ in the set of medications $\mathbf{M}$ multiplied by the conversion value $c$ for a given route $r$ (@eq-ome).

$$\text{TOB} = \sum_{i\in\mathbf{M}} d_i c_{i,r}
$${#eq-ome}

A number of conversion value tables have been published in an attempt to provide a *"best estimate"* of the conversion values between opioids [@mcphersonDemystifyingOpioidConversion2010; @nielsenSynthesisOralMorphine2016; @dowellCDCClinicalPractice2022]. Within clinical practice, these cited value tables are conventionally treated as a good surrogate for the true conversion value @mercadanteConversionRatiosOpioid2023. However, these values have been co-opted to produce metrics used in research despite a lack of clear methodological quality assurance. A commonly used, and often primary, outcome in perioperative analgesic studies is the difference in opioid requirement between groups, as measured by comparing the arithmetic mean of summed TOB. 

There is an implicit claim that once converted, it no longer matters which opioids a patient receives. However, the mathematical proof in @sec-proof demonstrates that this is only true if one of the two following conditions are satisfied (@eq-logic):

1) If the mixture of opioids in each group is the same 

2) There is no error in the conversion values used

$$
\Delta \mu = \Delta \mu^0 \iff \left(\Delta p_i = 0 \ \forall i\right) \lor \left(\epsilon_i = 0 \ \forall i\right) \tag{A.6}
$$

If neither condition is satisfied exactly, a previously undescribed bias arises that can skew or even invert the results of a study. In this simulation study we aim to quantify the extent of this bias and furthermore to estimate the impact on the current body of literature. 

## Methods

Two simulation studies were performed to understand the extent and impact of the bias on studies comparing group mean differences in TOB (@sec-sims). The first simulation explores the the parameter space between groups with inversely distributed opioids groups and conversion values with errors of $\pm$ 100% of a *"ground truth"* value. The second, simulates a realistic hypothetical study of `r shared_objects$n_second_sim` participants to analyse the effect of robotic surgery on pain, using an *"exogenous shock"* to vary the inter-group opioid proportions and varying conversion values to vary the error. The outcome in both simulations was the degree of bias, estimated by measuring the observed difference compared to the expected difference in group mean TOB.

To assess the current practices of TOB methodology and reporting in perioperative research a exploratory literature search was performed (@sec-lit-review). After screening, data was extracted from a random subset of 20% of the eligible studies. For manuscripts that reported the group mean doses of each opioid, it was attempted to recalculate the outcomes as a point estimate and then as a minimum and maximum using a range of different conversion values. A literature search protocol was then formalised (@sec-sm-lit-protocol) and registered on the [Open Science Framework](https://doi.org/10.17605/OSF.IO/EFJ8G).

No ethical approvals were required for this study. All simulations were performed in `r version$version.string` and code is available on [Github](https://github.com/DW10/ome_simulation_manuscript). Methods are reported with the relevant guideline [@morrisUsingSimulationStudies2019; @moherPreferredReportingItems2015].

## Results

The first simulation found that, as shown in @eq-logic, if either condition is met the observed difference is equal to the expected difference and there is no bias. Where there is a 10% error in the conversion value of oxycodone used, the resulting observed difference in outcome can be up to $\pm$ 10% the true difference if the opioids within the groups are completely mismatched. The linear bias that arises from either a difference in opioid proportions or conversion value error compound to diverge the observed and expected difference further.

The second simulation shows that this bias is relevant under the realistic conditions of the hypothetical study. The observed treatment effect may overestimate the intervention's true efficacy or paradoxically invert its direction, suggesting harm from a beneficial treatment.

The exploratory review included `r n_lit_data` eligible studies that underwent primary data extraction, of which `r n_lit_data* lit_not_described` (`r scales::percent(lit_not_described)`) stated the conversion values used while `r lit_provide_doses* n_lit_data` (`r scales::percent( lit_provide_doses)`) reported the proportion of opioids between groups. Of which, only 5 studies provided the within group mean opioid doses to allow the calculation of the outcome metrics independently. 

@fig-1 shows the study reported outcome alongside the recalculated outcome for each of the 5 studies. It demonstrates that for studies where there was no difference in opioid proportions (Fedorka 2025 @fedorkacjLiposomalBupivacaineBupivacaine2025) there is no bias, despite using a range of possible different conversion values . However, in studies where the two groups received different analgesic options (Carter 2025 @cartermEffectNonsteroidalAntiInflammatory2025), the observed value can fall within a 10% range. 

```{r}
#| label: fig-1
#| fig-cap:  'Forest style plot illustrating the percentage change in opioid use associated with an intervention. Blue cross denote the study reported outcome. Orange diamond and error bar denote the the primary point estimate and range using different conversion values.'
plot_point_whisker
```


## Discussion

In observational research, it will be vanishingly rare that the first condition of balanced groups is met exactly. An RCT may offer more hope of eliminating the TOB bias, however if a collider bias occurs (where the treatment itself has an effect on the choice of analgesia used @pearlBookWhyNew2019) or the post-operative analgesia treatment is not protocolised, the bias will persist.

The influence of the bias in the conversion value is a significant problem; the degree of error to the biological ground truth can never be known and estimates very widely [@rennickVariabilityOpioidEquivalence2016; @yessaillianSystematicReviewMorphine2024]. It is known that the conversion value tables are built on small, single opioid studies and have been formalised from clinical rules of thumb @mcphersonWhyEquianalgesicTables2020. We propose that data driven distributions of possible conversion value should instead be produced, this would allow quantification of uncertainty in the TOB metrics.

This study is limited by only simulating a two opioid paradigm and by of comparing group means; it is unclear how this bias will interact in multiple opioid studies or comparison of group medians. Furthermore, the literature search without a systemic approach isn't representative of the wider literature. 

We recommend:

1. All studies using opioid equivalence metrics must publish the proportions of the opioids within each group and conversion value tables used. 

2. More work is required to estimate the true population level opioid conversion value tables to reduce the error between the true potency and the value tables used. 


