---
format: html
---

# Manuscript

```{r source_files}
#| echo: false
#| message: false

source("functions.R", echo = FALSE)

shared_objects <- readRDS("data/shared_objects_sim.rds")
plot_simple_grid_2 =shared_objects$plot_simple_grid_2
plot_simple_grid = shared_objects$plot_simple_grid
plot_grid_2 = shared_objects$plot_grid_2
bounds=shared_objects$bounds

shared_objects_lit <- readRDS("data/shared_objects_lit.rds")
n_lit_data <- shared_objects_lit$n_lit_data
lit_not_described <- shared_objects_lit$lit_not_described
lit_provide_doses <- shared_objects_lit$lit_provide_doses
plot_point_whisker <- shared_objects_lit$plot_point_whisker
```

## Introduction

There are a wide range of different opioid used across the range of medical specialities and patient groups, each with their own potency to provide analgesia @bhatnagarOpioidEquivalency2026. Opioid conversion values have offered the ability to convert a specific opioid into the equivalent units of a reference opioid, usually morphine, to produce doses in oral morphine equivalent milligrams (OME mg) or milligram morphine equivalent (MME).  

A number of conversion value tables have been published in an attempt to provide a best estimate of the conversion values between opioids [@mcphersonDemystifyingOpioidConversion2010; @nielsenSynthesisOralMorphine2016; @dowellCDCClinicalPractice2022]. Within clinical practice, these cited value sets are a good surrogate for the true conversion value when used in opioid rotation @mercadanteConversionRatiosOpioid2023. Despite this, application of the value sets remains variable in both clinical and research practice [@rennickVariabilityOpioidEquivalence2016; @yessaillianSystematicReviewMorphine2024].

The total opioid burden (TOB) for patient $j$ is defined as the a weighted sum of the dose $d$ for medication $i$ in the set of medications $\mathbf{M}$ multiplied by the conversion value $c$ for a given route $r$ (@eq-ome).

$$\text{TOB}_j = \sum_{i\in\mathbf{M}} d_i c_{i,r}
$${#eq-ome}

For many studies, the TOB for each patient within a group, is often averaged and compared to draw conclusions about difference in opioid requirement. There is an implicit claim that once converted, it no longer matters which opioids a patient receives. However, this is only true if one of the two following conditions are true:

::: {.callout-important title='Conditions to use TOB metrics'}
1) If the mixture of opioids in each group are the same 

2) There is no error in the conversion values used
:::

We aim to quantify the degree of this bias and describe the methodologies of reporting of TOB metrics used in the perioperative literature. 

## Methods

Two simulation studies where performed as detailed in @sec-sims. In both, an observational comparative trial was simulated between a control and intervention group with both groups receiving varying proportions of morphine and oxycodone. Opioid requirement was converted to an oxycodone dose ($d_{\text{oxycodone}}$) using a *"ground truth"* conversion value ($c_\text{oxycodone}$) of `r shared_objects$potency_oxycodone` and then converted back to a OME mg using $c_\text{oxycodone}$ with an error term ($\epsilon$). The outcome measured was the observed difference in group mean TOB.

In the first simulation, 10,000 participants are simulated for each of the grid values across the parameter space ($\Delta p_i \in$ `r shared_objects$p_min`, `r shared_objects$p_max` and $\epsilon \in$ `r percent(shared_objects$e_min)`, `r percent(shared_objects$e_max)`) with no treatment effect. The second simulates a realistic hypothetical study of `r shared_objects$n_second_sim` participants, of the effect of robotic surgery on pain, to understand the extent of the bias under realistic study conditions. The robotic surgery has a true treatment effect of `r shared_objects$robot_effect_mg` OME mg.

To assess the current practices of TOB methodology and reporting in perioperative research a exploratory literature search was performed (@sec-lit-review). After screening, data was extracted from a random subset of 20% of eligable studies. Manuscripts that reported the group mean doses of each opioid, it was attempted to recalculate the TOB outcomes as a point estimate and then as a minimum and maximum using a range of different conversion values.

literature rapid review protocol was developed . To assess the protocol, a preliminary pilot was performed to screen and extract data from 20% of the the studies. From eligible studies  The protocol was then formalised ( @sec-sm-lit-protocol) and registered on the [Open Science Framework](https://doi.org/10.17605/OSF.IO/EFJ8G)..    

No ethics was required for this study. All simulations were performed in `r version$version.string` and code is available on [Github](https://github.com/DW10/ome_simulation_manuscript). Methods are reported with the relevant guideline [@morrisUsingSimulationStudies2019; @moherPreferredReportingItems2015].

## Results

The first simulation found that if either condition is met (equal group opioid proportion or no error in the conversion value), the observed difference is equal to the expected difference, thus there is no bias. When fixing the error in the conversion value, a linear bias arises with respect to any difference in opioid proportions between groups as show in @fig-1. When the error is also varied, this bias compounds. Under the hypothetical experimental conditions of a observational trial, the bias is able to significantly alter the results, mask and even invert the observed treatment effect as shown in @fig-2. 


::: {#fig-sim-ome layout-ncol=2}

```{r}
#| label: fig-1
#| fig-cap: "**Results of the simple simulation** - Given a treatment effect of 0mg across the enitre parameter space of inverted group proportions and 100% errors in the conversion values applied."
#| fig-width: 5
#| fig-height: 5

plot_simple_grid+ 
  geom_rect(ymin=bounds[1],
            ymax=bounds[2],
            xmin=bounds[3],
            xmax=bounds[4],
            colour="black",
            fill=alpha("grey",0))+
  geom_line(data = data.frame(x = c(0,0), y = bounds[1:2]), aes(x,y, fill=NULL), linetype = "dashed", color = "red")
```

```{r}
#| label: fig-2
#| fig-cap: "**Results of the realistic hypothetical simulation** - Given a treatment effect of -5mg across the parameter space of the hypothetical experiment space."
#| fig-width: 5
#| fig-height: 3

plot_grid_2
```
Results of simulation studies demonstrating the parameter space of the bias. Heat map of the observed difference between the control and intervention group. Plotted on an axis of the error in the conversion value over the difference in proportion of opioids between groups. Black outline box and red dashed line represents the hypothetical experimental space and ground truth value of oxycodone use.
:::

Across the perioperative literature, `r n_lit_data` studies had data extracted. Only `r scales::percent(lit_not_described)` of papers described the conversion values used and `r scales::percent(lit_provide_doses)` described the proportions of different opioids between intervention groups. 5 studies reported the minimum dataset to attempt to reproduce the TOB calculation. 

## Discussion

In observational research, it will be vanishingly rare that the first condition of balanced groups is met exactly. Whilst a randomised controlled trial may offer more hope, a collider bias may easily occur if the treatment itself has an effect on the choice of analgesia used, if the post-operative analgesia treatment is not protocolised @pearlBookWhyNew2019.

The influence of the error in the conversion value is a significant problem.  McPherson describes a range of limitations in the use of these equianalgesic tables for clinical care including the single opioid studies as source of the data and the lack of patient specific variables @mcphersonWhyEquianalgesicTables2020. We propose that distributions of possible conversion value should instead be produced, this would allow quantitation of uncertainty in the TOB metrics and also offer the opportunity to tailor conversion values to specific populations that have known pharmacokinetic differences @leeOpioidRequirementsResponses1997.

This study is limited by only simulating a two opioid paradigm and by of comparing group means; it is unclear how this bias may effect multiple opioid or comparison of group medians. Furthermore, the preliminary results from the literature search may not be representative of the wider literature.  

We recommend the following:

1. All studies using opioid equivalence metrics must publish the proportions of the differing opioids for each group and conversion value sets used. 

2. More work is required to estimate the true population level opioid conversion value sets to reduce the error between the true potency and the value sets used. 


