---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Reader Notes{.unnumbered}
 
```{r set quality}
PRODUCTION_QUALITY <- FALSE
```
::: {.callout-note}
```{r print quality statement}
#| output: asis
if (PRODUCTION_QUALITY) {
  cat("High quality document - For review")
} else {
  cat("Low quality document. Edit index.qmd to set `PRODUCTION_QUALITY <- TRUE` before review.")
}
```
:::

:::{.callout-danger}

The premise of the paper is that when the the proportions of opioid between groups are different **AND** OR there is an error in the conversions value used **THEN** the OME metric is not comparable between groups. This is *maybe* a version of Simpsons Paradox however it seems to be a metric commonly used as a primary outcome in studies. 

For this issue to have an impact there needs to be:

- Difference in proportions of opioids between groups

  - We show that this is not well reported so we cannot know the extent of this problem
  
- An error in the conversion value
  
  - We cannot know the "ground truth" so we cannot calculate the exact error
  
  - But we can use the "breath" of values used in the literature, this though is generally narrow (Almost everyone uses Neilson's value set)
  
The headline conclusions are:

For OME to be a reliable metric we need to report opioid proportions (and potentially find a way to correct post-hoc) AND find a way to estimate the "ground truth" conversion values better than the current estimates. 


:::


# Abstract {.unnumbered}



*Abstract to be written*
