---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Reader Notes{.unnumbered}
 
```{r set quality}
PRODUCTION_QUALITY <- FALSE

```
::: {.callout-note}
```{r print quality statement}

#| output: asis
if (PRODUCTION_QUALITY) {
  cat("High quality render - Ready for review")
} else {
  cat("Low quality render.  Edit index.qmd to set `PRODUCTION_QUALITY <- TRUE` before review.")
}
```
:::

:::{.callout-danger}

The premise of the paper is that when the the proportions of opioid between groups are different **AND** OR there is an error in the conversions value used **THEN** the OME metric is not comparable between groups. This is *maybe* a version of Simpsons Paradox however it seems to be a metric commonly used as a primary outcome in studies. 

For this issue to have an impact there needs to be:

- Difference in proportions of opioids between groups

  - We show that this is not well reported so we cannot know the extent of this problem
  
- An error in the conversion value
  
  - We cannot know the "ground truth" so we cannot calculate the exact error
  
  - But we can use the "breath" of values used in the literature, this though is generally narrow (Almost everyone uses Neilson's value set)
  
The headline conclusions are:

For OME to be a reliable metric we need to report opioid proportions (and potentially find a way to correct post-hoc) AND find a way to estimate the "ground truth" conversion values better than the current estimates. 


:::

# Lay summary

In peri-operative research we use **opioid burden metrics** such as oral morphine equivalent (OME) or milligram morphine equivalent (MME) to pool the opioid dose across different opioids and routes for a patient. Often this pooled result is used to calculate group means and inter-group differences to estimate a treatment/associational effect. It is assumed that it doesn't matter which opioids are in each group as it is accounted for using the opioid burden metric. 

We hypothesis that the differing opioids within the groups can bias the mean opioid metrics in un-equal ways between groups. We also hypothesis that if the conversion value is not exactly the same as the ground truth this will also contribute to bias. 

We describe the equations used in practice and highlight their dependence on the proportions and error in conversion values. We then carried out a simulations to understand this relationship and demonstrate the bias introduced. 

To establish the extent of the issue in research we carried out a literature review to show how many studies publish their conversion metrics and intra-group medication proportions. We demonstrate that only a low percentage report both of the key aspects. 

For papers that did supply the intra-group proportions, we simulated applying a range of different, but accepted, conversion values to demonstrate how the conclusions of the paper are dependant on the opioids used in each group. 

We conclude that going forward all studies using an opioid metric must report the conversion values and proportions of medication within each group. We also call for new approaches to estimate conversion values that are offer a clearer view of the uncertainty involved. 

# Abstract {.unnumbered}

*Abstract to be written*
