---
title: "Manuscript"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r source_files}
#| echo: FALSE
#| message: FALSE
#| output: FALSE

# Source the founctions in simulaton.qmd with a temporary R script containing only code blocks
temp_r_file <- tempfile(fileext = ".R")
knitr::purl("simulation.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
knitr::purl("index.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
unlink(temp_r_file)
rm(temp_r_file)
```

```{r set theme}
#| warning: false
#| message: false

library(scales)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(patchwork)
library(latex2exp)

theme_academic <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      text = element_text(family = "sans"),
      plot.title = element_text(face = "bold", size = 14),
      strip.background = element_rect(fill = "white", color = "black")
    )
}
# Add scale_y_continuous(expand = expansion(mult = c(0, 0.05))) to plots 
# to ensure the Y axis starts at 0 and bars touch the X axis.

scale_fill_academic <- function() {
  scale_fill_viridis_d(option = "magma", begin = 0.3, end = 0.8)
}

```


## Origin of opioid equivalence and eqipotency

Opioid medications are an important pillar of analgesia in medicine, used across a range of specialities and patients. There are a wide range of opioids, some naturally derived with others synthetic, each with their own potency. Whilst a wide range of factors may influence the potency of a specific opioid in a specific patient, there has long been a need to calculate the eqi-analgesic dose, or the dose of a medication that would provide an equal amount of analgesia compared to a reference. 

The need for agreed conversion values has lead to a number of publications over the last 20 years aiming to formalise the differing potencies. Much of this literature comes from the domain of palliative care in the context of opioid rotation and looked at multiple single opioid small studies to establish conversion values. McPherson described a range of limitations in the use of these equialangesic tables including the source of the data, lack of patient specific variables and bi-directionality of equivalence values. @mcphersonWhyEquianalgesicTables2020. Despite this, a set of approximate conversion values, taken with clinical experience offer a helpful set of starting points when prescribing different opioids 

The equi-analgesic conversion set have offered a new opportunity in research, the ability to quantify the total opioid burden for a group of patients.  With the increase in electronic health records, opportunities to now use total opioid use as either a feature or an outcome has become increasingly prevalent. In the last 5 years there have been 4,031 papers using the terms *"Milligram morphine equivalent"* (MME) or *"Oral morphine equivalent"* (OME) (@sec-appendix-pubmed-1). 

In 2016, Neilsen *et. al.* published on the first formal OME conversion tables for the intention of research @nielsenSynthesisOralMorphine2016. The authors aimed to source a range of national guidelines and recommended a single value for each opioid and route with a upper and lower quartile. This "common conversion set" paved the way for the idea that all opioid research should be using a single unifed conversion set. It was clear in the literature that using a range of different conversion sets would not produce equitable or comparable results\[**Add citations**\]. In an attempt to unify value sets, in 2022 the Center for disease control (CDC) produced their own value set an obligated any NIH researches working in opioid research to use their opioid conversions @dowellCDCClinicalPractice2022. 

A range of papers have highlighted some of the limitations and issues with opioid conversions sets in research. Namely the lack of accounting for inter-patient variability and the differential application of conversion sets. \[**Add citations**\]

Two aspects that have not been covered are the dependence on the proportion of each opioid and the error within the conversion values used. 
\[Add citations about who said/didn't say what\]


## Calculating opioid equivalence in clinical practise

In clinical practice the term **potency** is used to describe the relative strength of one opioid compared to another. The **eqi-potency** of a medication is commonly understood to be the dose required to achieve equal effect, and in the case of opioids, the equianalgesic effect.\[**Add citations**\] Thus @eq-potency describes the relationship between the conversion value $c$ and the potency $\rho$ of a given medication $i$ via a given route $r$.

$$c_{i,r} = \frac{1}{\rho_{i,r}}
$$ {#eq-potency} 

The value of $c$, then allows for the calculation of the equivalent effective dose of a target medication $d_{\text{target}}$, given a dose of a reference medication $d_{\text{referance}}$ (@eq-equivelent-dose). 

$$
d_{\text{referance}} \times c_{i,r} = d_{\text{target}}
$$ {#eq-equivelent-dose}

A worked example is provided in @sec-conversion-example.

## Calculating opioid equivalence in research

The technique of comparing opioid equivalent doses has lead to the development of a new metric, the **total opioid burden**. By converting each medication into it's equi-analgesic dose of the reference medication and then calculating the sum, the resulting value is an estimate of the total opioid analgesic effect administered. The total opioid burden metric is often presented as the weighted sum of opioids administered and provided in units of  of **oral morphine equivalent milligrams (OME mg)** or **milligram morphine equivalent (MME)**. 

In line with the literature we will refer to the total opioid burden as $\text{OME}$. We can define $\text{OME}_j$ as the sum of the dose $d$ for medication $i$ in the set of medications $\mathbf{M}$ multiplied by the conversion value $c$ for a given route $r$ for patient $j$ (@eq-ome). $\text{OME}_j$ is often  expressed as a total of all opioids administered or within a specific time frame (*e.g.* 24 hours post-operative).

$$\text{OME}_j = \sum_{i\in\mathbf{M}} d_i c_{i,r}
$${#eq-ome}

Assuming a patient's pain is well controlled and that the analgesia provided is due predominantly to the opioid burden then the total opioid burden of a patient $\text{OME}_j $ is equivalent to the total pain burden (@eq-pain-burden). 

$$
\begin{aligned}
\text{Pain}_{\text{total}} &= \text{Pain}_{\text{experienced}} + \text{Analgesia} \\[1em]
\text{Where} & \qquad  \text{Pain}_{\text{experienced}} \approx 0 \\
 \text{and } & \qquad \text{Opioid burden} \approx \text{Analgesia} \\
 \text{then: } & \qquad  \text{OME}_j   \approx \text{Pain}_{\text{total}}
\end{aligned}
$$ {#eq-pain-burden}

Given @eq-pain-burden and the likley parametric distribution of OME values within a group, the arithmetic mean of OME values $\mu_{\text{OME}}$ if often calculated and used as an outcome in studies aiming to estimate the total pain. 

As the $\mu_{\text{OME}}^{(G)}$ is a **mean of weighted sums** it can instead be expressed as a **weighted sum of mean doses** whereby the mean dose of a given medication within the group $\bar{d_i}$ is weighted by the conversion values $c_{i,r}$ (@eq-mu-ome-final - rearrangement provided in @sec-equation-proof).

$$\begin{aligned}
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j=G} \text{OME}_j\\
\mu_{\text{OME}}^{(G)}  &= \sum_{i\in\mathbf{M}} \bar{d_{i}} c_{i,r} 
\end{aligned}
$${#eq-mu-ome-final}

@eq-mu-ome-final demonstrates that the group mean ($\mu_{\text{OME}}^{(G)}$) is dependant on the relative proportions of the medications within each group. For two groups $A$ and $B$ where the mean dose of a given opioid $i$ is not then same then the resulting means OME values are not equivalent.
$$
\begin{aligned}
\text{where } \bar{d}_{i}^{(A)} \neq \bar{d}_{i}^{(B)} &\implies \mu_{\text{OME}}^{(A)} \neq \mu_{\text{OME}}^{(B)}
\end{aligned}
$$
Additionally, the conversion value ($c_{i,r}$) is a scaler applied to the mean dose ${\bar{d_i}}$. Where the conversion value provided is not equal to the ground truth, a further bias will be introduced. If there is a  difference in proportions of medication, the bias introduced by the error in the conversion value will cause the resulting error to diverge. 

## Aims

We will aim to:

1. Investigate the methods of calculating total opioid burden in the peri-operative and surgical literature.

2. Simulate and describe the dependence of differing opioid proportions and errors in conversion values on the the OME values calculated. 

# Methods

## Rapid Literature Review

A rapid literature review of the use of opioid equivalence metrics in peri-operative and surgical studies published in 2025 was performed. A structured search of the PUBMED database as detailed in @sec-appendix-pubmed was performed on 7th July 2026. Papers were screened and data extracted using the Covidence platform @zotero-item-670 with a custom proforma (@sec-appendix-cov-extraction).

```{r}
# To get screening data:
# Covidence -> Export -> References subsection -> Title and absract screening -> Export as .csv -> Drag to data foler -> Rename to lit_screen.csn
lit_screen <- read.csv("data/lit_screen.csv")
lit_select <- read.csv("data/lit_select.csv")
lit_included <- read.csv("data/lit_included.csv")
lit_excluded <- read.csv("data/lit_excluded.csv")
lit_data <- read.csv("data/lit_data.csv") %>% filter(Reviewer.Name == "Dylan Whitaker")
conversion <- read.csv("data/conversion_sets.csv")

n_lit_data <- nrow(lit_data)

```

## Simulated study

Two simulated studies were performed to demonstrate the sensitivity of the opioid burden metric to differences in the proportions of medications between groups and the conversion values used. In both simulations the outcome was the measured difference in the group mean total opioid burden.

The first (@sec-manuscript-simple-sim) uses a simplified simulation with no treatment effect between groups to understand the variation in results across the full range of parameters.

The second simulation (@sec-manuscript-exemplar-sim) uses a hypothetical experimental setup with realistic values to establish the extent of the variation in results under realistic conditions. 

### Simple simulation {#sec-manuscript-simple-sim}

```{r sim simple}

p_min <- -1
p_max <- 1
e_min <- -1
e_max <- 1
n = 10000
mu_opioid_requirment_control = 10
sigma = 0.1
treatment_effect = 0
potency_oxycodone = 2

simple_results <- data.frame(p = numeric(), e = numeric(), diff = numeric())
for(p in seq(p_min, p_max, by = 0.05)){
  for(e in seq(e_min, e_max, by = 0.05)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu_opioid_requirment_control,
      potency_oxycodone = potency_oxycodone,
      error_in_c_oxycodone = e,
      treatment_effect = treatment_effect
    )
    
    simple_results <- rbind(simple_results, data.frame(p = p, e = e, diff = diff))
  }
}
```

`r comma(n)` samples were divided equally between a control and intervention group. An opioid of either morphine or oxycodone was then samples from a distribution set by $\Delta p_i$ where $\Delta p_i = p_i^{\text{(control)}} - p_i^{\text{(intervention)}}$. An opioid requirement was sampled from a normal distribution ($\mathbb{E} =$   `r mu_opioid_requirment_control` with $\sigma =$ `r sigma`) for both groups enforcing a treatment effect of `r percent(treatment_effect)`. For samples where the opioid was oxycodone, the opioid requirement was converted to an oxycodone requirement with a ground truth potency equal to `r potency_oxycodone` and then back into an OME mg value using a conversion value with a error term set to $\epsilon$ where $\delta c_i = c_i + (c_i * \epsilon)$. For samples where the opioid was morphine the opioid requirement is equal to the OME mg as it is the reference medication. 

The simulation was then performed for the range of grid values for ($\Delta p_i \in$ `r p_min`, `r p_max` and $\epsilon \in$ `r e_min`, `r e_max` with all other parameters fixed. The result for each study was the measured difference between the group means of the control and intervention groups ($\Delta \mu_{\text{OME}} = \mu_{\text{OME}}^{\text{(control)}} - \mu_{\text{OME}}^{\text{(intervention)}}$).


### Exemplar simulation {#sec-manuscript-exemplar-sim}

```{r}
case_volume <- 100
proportion_robotic <- 0.8
t_0 = as.Date("2020-01-01")
t_end = as.Date("2025-12-31")
t_robot = as.Date("2023-01-01")
proportion_morphine = 0.7
oxycodone_shortage = c(as.Date("2024-01-01"), as.Date("2024-12-31"))

mu_tot = 50
sd_tot = 10
potency_oxycodone_truth = 1.85
potency_oxycodone_ome = 2

annual_n = 30
robot_effect_size = 0.90
robot_effect_mg = mu_tot * (robot_effect_size-1)

p_morph_high = 0.8
p_morph_low = 0.2
conversion_value_low = 1
conversion_value_high = 3
```

A exemplar simulation was performed with the parameters outlines in @nte-sim-design. 

::: {#nte-sim-design .callout-note title="Simulated trial design" collapse=true appearance="simple"}
A research team wish to understand if the use of a robot for a surgical procedure, reduces the post-operative pain.

The average case volume is `r case_volume` procedures per year, these were all performed using non-robotic surgery until `r format(t_robot, "%e %b %y")` when a robot was acquired, and since `r scales::percent(proportion_robotic)` of cases are performed with the robot. All EHR data was collected between `r format(t_0, "%e %b %y")` and `r format(t_end, "%e %b %y")`.

All patients are treated with a Patient Controlled Analgesia (PCA) after their operation and on average the anaesthetist will prescribe a morphine PCA to `r scales::percent(proportion_morphine)` of patients and an oxycodone PCA to the remainder. No other opioid medication was administered. 

They design a retrospective cohort study for patients treated with a robotic procedure and those without. The primary outcome is the total opioid requirement, as measured in OME mg, in the first 24 hours post-operatively as a surrogate for the total pain caused by the surgical procedure.

During the study there was a national shortage of oxycodone from `r oxycodone_shortage[1]` to `r oxycodone_shortage[2]` and at which time all patients were prescribed a morphine PCA. Whilst the authors described this in their methods section, they did feel it would impact their results as the differing opioids were already accounted for in their OME metric. 
:::

The difference in proportion ($\Delta p_i$) between groups is simulated using a hypothetical medication shortage that due to temporal differences in the groups, effects the treatment group to a greater extent. The ground truth potency of oxycodone is set to `r potency_oxycodone_truth`. The post-operative opioid requirement was sampled from a normal distribution ($\mathbb{E} =$ `r mu_tot` with $\sigma =$ `r sd_tot`) for the non-robotic group with the average treatment effect of the intervention is set to `r robot_effect_mg` OME mg as informed by a power analysis in @sec-appendix-simulation-power. OME was calculated as per described previously.

The simulation was performed once with parameters specified in @nte-sim-design and then across the grid values for the rate of post-operative morphine prescribed between `r percent(p_morph_high)` and `r percent(p_morph_low)` with conversion values of oxycodone:morphine between `r conversion_value_low`:1 and `r conversion_value_high`:1. The result for each study was the measured difference between the group means of the control and intervention groups ($\Delta \mu_{\text{OME}}$), the observed difference in proportion ($\Delta p_i$) and the error in the conversion value ($\epsilon$).  

All simulations were performed in `r version$version.string` and all code is available in @sec-sim. 

# Results

## Rapid Literature Review 

Of the 480 studies identified in the rapid literature search, `r nrow(lit_data)` fit the eligibility criteria and data was extracted. 

The majority of studies were cohort studies and most commonly looked at anaesthetic type interventions (@fig-study-types)

```{r}
#| label: fig-study-types
#| fig-cap: "**Study design and intervention types in included literature** - Bar charts summarising the study design and intervention type of studies included in the rapid literature review. Left panel shows count by study design; right panel shows count by intervention type."
#| fig-width: 10
#| fig-height: 5

# Plot for Study.type proportions
plot_lit_study_type <- ggplot(lit_data, aes(x = Study.type)) +
  geom_bar(aes(fill = Study.type)) +
  labs(x = "Study design",
       y = "Count") +
  theme_minimal() +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_academic()+
  theme(legend.position = "none")

# Plot for Intervention proportions
plot_lit_intervnetion_type <- ggplot(lit_data, aes(x = Intervention)) +
  geom_bar(aes(fill = Intervention)) +
  labs(x = "Intervention",
       y = "Count") +
  theme_minimal() +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_academic()+
  theme(legend.position = "none")

plot_lit_study_type + plot_lit_intervnetion_type

```

With regards to the opioid conversion sets used, `r lit_data %>% filter(Opioid.conversion.set == "Cited") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.conversion.set == "Cited") %>% nrow()/n_lit_data)`) cited another source, whilst `r lit_data %>% filter(Opioid.conversion.set == "Included in text") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.conversion.set == "Included in text") %>% nrow()/n_lit_data)`) included the conversion values within the paper. `r lit_data %>% filter(Opioid.conversion.set == "Not described") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.conversion.set == "Not described") %>% nrow()/n_lit_data)`) papers did not include any citation or description of the opioid conversion sets used. 

Upon examining the results section, it was found that only `r lit_data %>% filter(Opioid.proportions.within.group == "Described") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.proportions.within.group == "Described") %>% nrow()/n_lit_data)`) papers described the proportions of different opioids between intervention groups(@fig-study-results). 

```{r}
#| label: fig-study-results
#| fig-cap: "**Opioid conversion set reporting and within-group proportion documentation** - Heatmap showing the count of studies by opioid conversion set type and whether opioid proportions within groups were reported. Cell counts are displayed as text labels."

# Prepare data for heatmap: sum of occurrences per combination
heatmap_data <- lit_data |>
  group_by(Opioid.conversion.set, Opioid.proportions.within.group) |>
  summarise(count = n(), .groups = "drop")

# Prepare data with a conditional color column for labels
ggplot(heatmap_data, aes(y = Opioid.proportions.within.group, x = Opioid.conversion.set, fill = count)) +
  geom_tile() +
  geom_label(aes(label = paste0("n=",count),
                 color = after_scale(ifelse(fill > 0.5, "black", "white"))),
             fill = alpha("white", 0.7), size = 3.5) +
  labs(y = "Proportion of opioids within groups",
       x = "Opioid conversion set",
       fill = "Count") +
  theme_minimal()+
 theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(angle = 45),
    panel.grid = element_blank(),
    legend.position = "none")+
  symmetric_fill_scale(max(heatmap_data$count))
```

### Conversion values

Of the `r lit_data %>% filter(Opioid.conversion.set == "Cited") %>% nrow()` papers where a conversion value set was cited, the `r nrow(conversion)` unique source value sets where reviewed. Generally across the literature, conversion values with the intention of research were in the minority (`r percent((nrow(conversion %>% filter(aim == "for research")))/(nrow(conversion)))`) and most had compiled the conversion value sets for use in clinical care. 

:::{.callout-warning}
Can expand this section but not sure if its worth it?
:::

## Simulation study 

### Simple simulation {#sec-manuscript-results-simple}

`r comma(nrow(simple_results))` different simulations where performed across the grid values of $\Delta p_{\text{oxycodone}}$ and $\epsilon$. The observed difference across the grid values demonstrates a normal distribution, centered around the true treatment effect of `r treatment_effect` (@fig-simple-density).

```{r}
#| label: fig-simple-density
#| fig-cap: "**Distribution of observed group mean difference across the simple simulation** - Kernel density of the observed difference in group means (ΔOME) across all grid parameter combinations. The dashed red line indicates the true treatment effect of zero."

ggplot(simple_results, aes(x=diff))+
  geom_density(fill="blue", alpha = 0.8)+
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed")+
  labs(y = "Density",
       x = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) 

```

When plotted against the grid values, it becomes clear that when the proportion of medication in the control and intervention groups are the same ($\Delta p_i = 0$) then the true effect size of 0 is recovered, regardless of any error in the conversion value ($\delta c_i$) used (@fig-simple-grid-1). Furthermore, where the conversion values used are correct to the ground truth, any difference in proportion is accounted for and the true effect is again recovered.

However, as the proportions between the control and treatment group deviate, resulting difference observed is biased by any error in the conversion values. Where the groups are completely unmatched ($\Delta p_i =$ 1 or -1) the resulting observed difference is the due entirely to the error in potency.

```{r}
#| label: fig-simple-grid-1
#| fig-cap: "**Observed OME group mean difference across opioid proportion difference and conversion value error** - Heatmap of the observed difference in group means (ΔOME) across grid values of Δp_i (difference in opioid proportion between groups) and ε (error in the oxycodone conversion value). Bias is zero when either parameter is zero and increases as both deviate from zero."


plot_simple_grid <- ggplot(simple_results, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
 # scale_fill_gradientn(
#  colours = c("blue", "dodgerblue", "white", "orange", "red"))+
#  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
#  geom_hline(yintercept = 0, linetype = "dashed", color = "black")+
#  annotate("text",x=-0.7, y=0.05, label = "Proportion of medication equal")+
#  annotate("text",x=-0.7, y=0.95, label = "Control group: 100% morphine\nIntervention group: 0% morphine")+
 # annotate("text",x=-0.7, y=-0.95, label = "Control group: 100% oxycodone\nIntervention group: 0% oxycodone")+
  #annotate("text",x=-0.05, y=-0.7, label = "No error in conversion value", angle=90)+
  #annotate("text",x=0.7, y=-0.05, label = "Overestimate potency ->" )+
  #annotate("text",x=-0.7, y=-0.05, label = "<- Underestimate potency")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  symmetric_fill_scale(round(max(simple_results$diff))+0.01)

plot_simple_grid

```

:::{.callout-warning}

REVIEW MARKER
:::

@eq-results-simple-1 demonstrates the hyperbolic parabolic relationship observed in @fig-simple-grid-1. 

$$
\Delta \mu_{\text{OME}} = -1 \cdot \Delta p_i \cdot \delta c_i
$$ {#eq-results-simple-1}

@fig-appendix-simple-sim-mu shows that the scaling constant (1 in this simulation) is the expected dose for the control group. This relationship can be rearranged as @eq-results-simple-2 (as expanded in in @eq-results-simple-2-proof).

$$
\mu^\text{intervention} = \mu^\text{control}[ 1 + \Delta p_i \cdot \delta c_i]
$$ {#eq-results-simple-2}
 
However, in all of the simulations described thus far the treatment effect was set at 0 meaning $\mathbb{E}(\text{OME}^\text{control}) = \mathbb{E}(\text{OME}^\text{intervention})$. This demonstrated that despite have no true treatment effect, difference in the proportion and error in conversion size, can lead to the observation of a treatment effect. This phenomenon can be explained by Simpsons paradox, whereby the observed treatment effect is conditioned on the proportion of the the medications within each group, thus creating the observation of a false treatment effect @pearlBookWhyNew2019. 

@sec-appendix-sim-2 demonstrates that as the treatment effect is increased (of decreased) it stretched the hyperbolic parabola in the Z direction. 

This saddle geometry illustrates a critical vulnerability in aggregate OME comparisons: if a conversion factor for the medication $i$ is over-estimated ($\epsilon_i > 0$) and prevalence is simultaneously higher in the intervention group ($\Delta p_i > 0$), the error in the effect size compounds quadratically. Conversely, opposing directional errors will collapse and at the extreme values may invert the direction of the observed treatment effect.

### Exemplar simulation study

```{r }
df <- sim_patient(
  mu_tot = mu_tot,
  potency_oxycodone_truth = potency_oxycodone_truth,
  potency_oxycodone_ome = potency_oxycodone_ome,
  proportion_morphine = proportion_morphine,
  oxycodone_shortage = oxycodone_shortage,
  t_0 = t_0,
  t_end = t_end,
  t_robot = t_robot,
  annual_n = annual_n,
  robot_effect_size = robot_effect_size
)

```

`r nrow(df)` procedures were simulated using the parameters described in @nte-sim-design. The OME value for each patient was calculated use $c_{\text{morphine}}$ as 1 (as the reference medication) and $c_{\text{oxycodone}}$ as `r 1/potency_oxycodone_ome`, in contract to the ground truth value of `r 1/potency_oxycodone_truth`. The mean dose (mg), opioid and OME value for each group is shown in @tbl-sim-demo.

```{r}
#| label: tbl-sim-demo
#| tbl-title: Demographic table of simulated study population

library(tableone)

table_1 <- CreateTableOne(data=df,
               vars = c("dose", "opioid", "ome"),
               strata = "robotic",
               addOverall = TRUE,
               test=FALSE)

kableone(table_1, format="latex")
```

::: {#fig-sim-density layout-ncol=3}
```{r}
#| label: fig-sim-dose-density
#| fig-cap: "**Dose distribution by opioid** - Kernel density of administered dose (mg) for each opioid type."

ggplot(df, aes(x=dose, fill=opioid))+
  geom_density(alpha=0.5)+
  theme_academic()+
  labs(,
        x="Dose (mg)",
        y="Density",
        fill="Opioid")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) 
```

```{r}
#| label: fig-sim-ome-density
#| fig-cap: "**OME distribution by surgery type** - Kernel density of calculated OME (OME mg) for robotic and non-robotic surgery groups."

ggplot(df, aes(x=ome, fill=robotic))+
  geom_density(alpha=0.5)+
  theme_academic()+
  labs(x="OME (OME mg)",
        y="Density",
        fill="Robotic Surgery")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) 

```

```{r}
#| label: fig-sim-ome-proportion
#| fig-cap: "**Opioid prescribing proportions by surgery type** - Proportion and count of each opioid prescribed within robotic and non-robotic surgery groups."

df %>%
  count(robotic, opioid) %>%
  group_by(robotic) %>%
  mutate(proportion = n / sum(n)) %>%
  ggplot(aes(x = robotic, y = proportion, fill = opioid)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = paste0("n=",n)),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 3
  )+
  labs(y = "Proportion", x = "", fill = "Opioid") +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent_format()) 

```

**Simulated patient dose, OME, and opioid mix by surgery type** - Three-panel figure showing (A) dose distribution by opioid, (B) OME distribution by surgery type, and (C) opioid prescribing proportions by surgery group.
:::

@fig-sim-density Demonstrates the distribution of the dose (in mg) of each medication (@fig-sim-dose-density) and the distribution of OME values (in OME mg) by surgery type (@fig-sim-ome-density). @fig-sim-ome-proportion shoes that within the robotic and non-robotic groups the proportion of each opioid is not equal, and the effect of the oxycodone shortage has had a larger impact on the robotic group.

```{r}
obs_robotic_ome <- df %>% filter(robotic == "Robotic") %>% summarise(mean(ome)) %>% pull() 
obs_non_robotic_ome <- df %>% filter(robotic != "Robotic") %>% summarise(mean(ome)) %>% pull()
obs_robot_effect_mg <- obs_robotic_ome - obs_non_robotic_ome
```

The mean OME value for the robotic surgery group was `r round(obs_robotic_ome, 1)` OME mg and for the non-robotic surgery group was `r  round(obs_non_robotic_ome, 1)` OME mg. This gives an observed effect size of the robot as `r round(obs_robot_effect_mg,1)` OME mg. 

```{r}
result <- analyse_dataset(df)
```

Using a two sided t-test to assess if there is a significant difference in mean OME between the robotic and standard surgery groups, the mean difference was `r round(result$mean_diff_ome,1)` OME mg (95% CI `r paste(round(result$conf_int_ome, 1), collapse = " to ")`) with a p-value of `r signif(result$p_value_ome, 3)`.

Thus the researched from St Somewhere Hospital were able to conclude that the robotic surgery was associated with a significantly different post-operative OME dose. 

### Impact of conversion set and proportions

The study was then recreated using a grid search of across different conversion values and different proportion of post-operative oxycodone prescribing. The prior was informed by ranges observed in the literature and the latter to infulence the degree of different in opioid proportions between groups.

```{r}
#| label: fig-sim-grid-search-1
#| fig-cap: "**Sensitivity of observed treatment effect to conversion value and opioid proportion — large effect** - Grid search results showing the observed treatment effect across oxycodone conversion values and post-operative morphine prescribing proportions. True treatment effect is large. The dashed line indicates the ground truth conversion value."

if (PRODUCTION_QUALITY) {
  trials = 100
} else {
  trials = 3
}

simulation1 <- run_simulation(trials=trials,
               conversion_values = seq(from=conversion_value_low, to=conversion_value_high, by = 0.2),
               morphine_proportion = seq(from=p_morph_low, to= p_morph_high, by = 0.05)
               )
plot_grid(simulation1,
          potency_oxycodone_truth = potency_oxycodone_truth,
          robot_effect_mg = robot_effect_mg)

```

Given a `r mu_tot` mg expected opioid requirement and a treatment effect of `r scales::percent(robot_effect_size-1)` mg we would expect the mean difference in OME of the robotic group to be `r (mu_tot * robot_effect_size)-mu_tot` mg OME.

The results shown in @fig-sim-grid-search-1 show that where the proportions of morphine morphine prescribed post-op, (thus meaning a shortage of oxycodone has little influence on prescribing patterns and hence the effect of the exogenous shock is little) then across the values of $c_{\text{Oxycodone}}$, the true difference in treatment groups is well observed. With all situations being statistically significant. However as the proportion of post-operative prescribing shifts towards oxycodone, the observed difference in OME between the two groups becomes highly sensitive to the conversion value used. This is due to the exogenous effect of the oxycodone shortage impacting the robotic group more than the the standard surgery group. 

In this study set up, the true effect of the robot is powerful with an true effect of -10mg which is equal to 1 SD. Power calculations in @sec-appendix-simulation-power demonstrate that this is a very highly powered study. 

```{r}
robot_effect_mg = -4
robot_effect_size = (mu_tot + robot_effect_mg)/mu_tot

```

The experiment was run again but with a more modest effect of `r robot_effect_mg`mg or a post-operative OME `r scales::percent(robot_effect_size)` percent of the non-robotic group.

```{r}
#| label: fig-sim-grid-search-2
#| fig-cap: "**Sensitivity of observed treatment effect to conversion value and opioid proportion — modest effect** - As above, but with a modest true treatment effect, illustrating greater sensitivity to conversion value error and opioid proportion differences."

if (PRODUCTION_QUALITY) {
  trials = 100
} else {
  trials = 3
}

#Note that the robot_effect_size vairable is hard coded to the global_env above 
simulation2 <- run_simulation(trials=trials,
               conversion_values = seq(from=1, to=3.5, by = 0.2),
               morphine_proportion = seq(from =0.1, to= 0.9, by = 0.05)
               )
plot_grid_2 <- plot_grid(simulation2,
          potency_oxycodone_truth = potency_oxycodone_truth,
          robot_effect_mg = robot_effect_mg)

```

The parameters of this exemplar simulation can be mapped to the previous simple simulation. Firstly the values of $\delta p_i$ are $\epsilon$ are recovered for each trial in the grid search. These are then plotted side by side with the results of the first simulation when the same parameters as the exemplaer experiment is used. @fig-compare demonstrates that the experiment space for the example covered only values where $\Delta p_i < 0$ and for values of $ -0.5 < \epsilon < 1 $. Despite being a relativity narrow section of the possible gird space detailed in @sec-manuscript-results-simple with a modest effect size of `r robot_effect_mg` OME mg there are still values where the observed effect size is significantly different. 

```{r}
#| label: fig-compare
#| fig-cap: "**Exemplar simulation parameters mapped onto the simple simulation space** - Side-by-side comparison of the simple and exemplar simulation grids. The region explored by the exemplar experiment (Δp_i < 0; −0.5 < ε < 1) is overlaid on the full simple simulation space, showing that even within this narrow parameter region a modest effect size produces meaningfully variable observed outcomes."


label_function <- function(val) {
  paste0('mu["OME"]^"(control)" == ', val )
}

bounds <- c(min(simulation2$delta_p_oxycodone),
            max(simulation2$delta_p_oxycodone),
            min(simulation2$epsilon),
            max(simulation2$epsilon))

simple_results <- data.frame(p = numeric(), e = numeric(), diff = numeric())
for(p in seq(p_min, p_max, by = 0.05)){
  for(e in seq(e_min, e_max, by = 0.05)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu_tot,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = robot_effect_size -1
    )
    
    simple_results <- rbind(simple_results, data.frame(p = p, e = e, diff = diff))
  }
}


plot_simple_grid <- ggplot(simple_results, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "none")+
  symmetric_fill_scale(20)+
  geom_rect(ymin=bounds[1],
            ymax=bounds[2],
            xmin=bounds[3],
            xmax=bounds[4],
            colour="black",
            fill=alpha("grey",0))+
  annotate("text", label = "Parameter space of exemplar simulation", x=mean(bounds[3:4]), y=bounds[2]+0.1)

plot_grid_2 <- ggplot(simulation2, aes(x=conversion_value, y=morphine_proportion, fill=mean_diff_ome)) +
  geom_tile() +
  symmetric_fill_scale(20)+
  labs(y = TeX(r'(Proportion of post-op perscribing as morphine )'),
       x = TeX(r'($\c_{oxycodone}$ (Conversion Value for Oxycodone used) )')) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")
  
layout <- c(
  area(t = 1, l = 1, b = 10, r = 10),
  area(t = 6, l = 11, b = 7, r = 20)
)

plot_simple_grid + plot_grid_2 +
  plot_layout(design = layout)

```

### Simulation of ...et al 

The final simulation looks at a study performed by xx et al. This study looked at the effect of xxx on yyyy and found a observed effect size of xxx. 

They did not report the difference of proportion of medications between their treatment groups. We will apply conversion values that have been used across the literature to calculate. One issue is that we do not know the true conversion values of the medications. 

# Conclusion

We have demonstrated that the use of opioid equivalence metrics such as OME or MME are sensitivity to the proportions of different opioids and the conversion sets used. We have demonstrated that the difference measured between the observed $\mu_{\text{OME}}^{(A)}$ and $\mu_{\text{OME}}^{(B)}$ is a product of not only the association of the two groups (be it statistical or causal), but also the difference in proportions of medications within each group. This is explicitly demonstrated in @sec-ome-difference-proof. 

In addition, the opioid equivalence metric is also sensitivity to any error in the true conversion value of the underlying population and the conversion value used for calculations. As expressed across the literature, it is difficult to ascertain all of the inter-patient difference and inter-opioid interactions that would provide the true conversion value. None the less, any difference in the true population level conversion value and the values used will also lead to systematic under or over estimation of the true value. 

Where the opioid equivalence metric is being used as outcome metric in a comparison between groups, any difference in proportion of differing opioids or the conversion values will lead to a systematic under or over estimation of the true different. If either one of these is exactly correct then the error will cancel out but if both are not matched it can lead to an up to 3 fold error in the estimate of difference. 

This work is particular relevant to retrospective observational work where large datasets with varied treatment lend themselves well to the tool of opioid conversion values. However, even in a RCT where the protocol for analgesia may be fixed and patients randomised between groups, confounding may occur if analgesia protocols contain differing opioids allowing the possibility of diverging proportions between treatment groups. 

## Limitations

In the simulation studies we have only considered situations where medication choice is binary between the reference medication and the alternative (oxycodone). We hypothesise that in some situations with multiple opioids, due to Orthogonality, that multiple small errors in the estimate of $c_i$ and the true value of $c_i$ may generally cancel out and thus have little impact on the final estimate. This may be achieved using machine learning tools that consider the co-linearity between the features of the dose of each medication and the outcome of total opioid burden or OME. 

## Recommendations:

1. All studies using opioid equivalence metrics must publish the proportions of the differing opioids and conversion value sets used for each group. 

2. More work is required to estimate the true population level opioid conversion value sets to reduce the error between the true unmeasurable potency and the value sets used. 

## Generative AI statement

Generative artificial intelligence tools where used to aid the writing of this manuscript and in production of code. These tools included Gemini 3.1 models (Pro, Flash and Flash-Lite), Claude sonnet 4.6 (via Posit AI) and Github Copilot (via Posit Code Assistant). All ideas and methods are of the authors own, all text and code was reviewed and edited following generation.

## CRediT statement

```{r}
tbl_authors <- googlesheets4::read_sheet("123JiCxSHvA7Wiaz6tU25B7Wq27Ty_cV7n6UUqlhyvrY",
                                         sheet = 2)
aut <- plume::Plume$new(tbl_authors, dotted_initials = F)
aut$get_contributions(roles_first = T)
```


