---
title: "Manuscript"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r source_files}
#| echo: FALSE
#| message: FALSE
#| output: FALSE

# Source the founctions in simulaton.qmd with a temporary R script containing only code blocks
temp_r_file <- tempfile(fileext = ".R")
knitr::purl("simulation.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
knitr::purl("index.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
unlink(temp_r_file)
rm(temp_r_file)

set.seed(402)
```

```{r set theme}
#| warning: false
#| message: false

library(scales)
library(dplyr)
library(ggplot2)
library(patchwork)
library(latex2exp)
library(gt)

theme_academic <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      text = element_text(family = "sans"),
      plot.title = element_text(face = "bold", size = 14),
      strip.background = element_rect(fill = "white", color = "black")
    )
}
# Add scale_y_continuous(expand = expansion(mult = c(0, 0.05))) to plots 
# to ensure the Y axis starts at 0 and bars touch the X axis.

scale_fill_academic <- function() {
  scale_fill_viridis_d(option = "magma", begin = 0.3, end = 0.8)
}

```

# Background 

:::{.callout-warning}
Background drafted, some more citations on the various commentery on OME/metrics/issues from the literature to come. 
:::

## Origin of opioid equivalence and eqipotency

Opioid medications are an important pillar of analgesia in medicine, used across a range of specialities and patients. There are a wide range of opioids, some naturally derived and others synthetic, each with their own potency. While a wide range of factors may influence the potency of a specific opioid in a specific patient, there has long been a need to calculate the eqi-analgesic dose, or the dose of a medication that would provide an equal amount of analgesia compared to a reference. 

The need for agreed conversion values has lead to a number of publications over the last 20 years aiming to formalise the differing potencies. Much of this literature comes from the domain of palliative care in the context of opioid rotation and looked at multiple single opioid small studies to establish conversion values. McPherson described a range of limitations in the use of these equialangesic tables including the source of the data, lack of patient specific variables and bi-directionality of equivalence values. @mcphersonWhyEquianalgesicTables2020. Despite this, a set of approximate conversion values, taken with clinical experience offer a helpful set of starting points when prescribing different opioids 

The equi-analgesic conversion set has offered a new opportunity in research, the ability to quantify the total opioid burden for a group of patients.  With the increase in electronic health records, opportunities to now use total opioid use as either a feature or an outcome has become increasingly prevalent. In the last 5 years there have been 4,031 papers using the terms *"Milligram morphine equivalent"* (MME) or *"Oral morphine equivalent"* (OME) (@sec-appendix-pubmed-1). 

In 2016, Neilsen *et. al.* published on the first formal OME conversion tables for the intention of research @nielsenSynthesisOralMorphine2016. The authors aimed to source a range of national guidelines and recommended a single value for each opioid and route with a upper and lower quartile. This "common conversion set" paved the way for the idea that all opioid research should be using a single unifed conversion set. It was clear in the literature that using a range of different conversion sets would not produce equitable or comparable results\[**Add citations**\]. In an attempt to unify value sets, in 2022 the Center for disease control (CDC) produced their own value set an obligated any NIH researches working in opioid research to use their opioid conversions @dowellCDCClinicalPractice2022. 

A range of papers have highlighted some of the limitations and issues with opioid conversions sets in research. Namely the lack of accounting for inter-patient variability and the differential application of conversion sets. \[**Add citations**\]

Two aspects that have not been covered are the dependence on the proportion of each opioid and the error within the conversion values used. 
\[Add citations about who said/didn't say what\]

:::{.callout-warning}
We have not explained WHY OME/total opioid burden is a useful or interesting metric. This is an implicit assumption left to the reader, is this ok? 
:::

## Calculating opioid equivalence

In clinical practice the term **potency** is used to describe the relative strength of one opioid compared to another. The **eqi-potency** of a medication is commonly understood to be the dose required to achieve equal effect, and in the case of opioids, the equianalgesic effect.\[**Add citations**\] Thus @eq-potency describes the relationship between the conversion value $c$ and the potency $\rho$ of a given medication $i$ via a given route $r$.

$$c_{i,r} = \frac{1}{\rho_{i,r}}
$$ {#eq-potency} 

A worked example to demonstrate how opioid conversion is performed in clinical practice is provided in @sec-conversion-example.

The technique of comparing opioid equivalent doses has lead to the development of a new metric, the **total opioid burden**. By converting each opioid medication into it's equi-analgesic dose of the reference medication, then calculating the sum, the resulting value is an estimate of the total opioid analgesic effect administered. The total opioid burden metric is often presented as the weighted sum of opioids administered and provided in units of  of **oral morphine equivalent milligrams (OME mg)** or **milligram morphine equivalent (MME)**. 

We can define the total opioid burden $\text{OME}_j$ as the sum of the dose $d$ for medication $i$ in the set of medications $\mathbf{M}$ multiplied by the conversion value $c$ for a given route $r$ for patient $j$ (@eq-ome). $\text{OME}_j$ is often  expressed as a total of all opioids administered or within a specific time frame (*e.g.* 24 hours post-operative).

$$\text{OME}_j = \sum_{i\in\mathbf{M}} d_i c_{i,r}
$${#eq-ome}

From this, the arithmetic mean of OME values $\mu_{\text{OME}}^{(G)}$ for a given group $G$ is often calculated and used as an outcome in studies aiming to estimate the average opioid burden. As the $\mu_{\text{OME}}^{(G)}$ is a **mean of weighted sums** it can instead be expressed as a **weighted sum of means** whereby the mean dose of a given medication within the group $\bar{d_i}$ is weighted by the conversion values $c_{i,r}$ (@eq-mu-ome-final with rearrangement provided in @sec-equation-proof).

$$\begin{aligned}
\mu_{\text{OME}}^{(G)} &= \frac{1}{|G|} \sum_{j=G} \text{OME}_j\\
\mu_{\text{OME}}^{(G)}  &= \sum_{i\in\mathbf{M}} \bar{d_{i}} c_{i,r}
\end{aligned}
$${#eq-mu-ome-final}

@eq-mu-ome-final demonstrates that the group mean ($\mu_{\text{OME}}^{(G)}$) is dependant on the relative proportions of the medications within each group. For two groups $A$ and $B$ where the mean dose of a given opioid $\bar{d}_{i}$ is not then same then the resulting means OME values are not equivalent. Furthermore, the conversion value ($c_{i,r}$) is a scaler applied to the mean dose ${\bar{d_i}}$. Where the conversion value provided is not equal to the ground truth, a further bias will be introduced. If there is a  difference in proportions of medication, the bias introduced by the error in the conversion value will cause the resulting error to diverge. 


## Aims

We will aim to:

1. Simulate and describe the dependence of differing opioid proportions and errors in conversion values on the the OME values calculated. 

2. Investigate the methods of calculating total opioid burden in the peri-operative and surgical literature.

# Methods

## Part 1 - Simulations

Two simulated studies were performed to demonstrate the sensitivity of the opioid burden metric to differences in the proportions of medications between groups and the conversion values used. In both simulations, the outcome was the measured difference in the group mean total opioid burden.

The **first** (@sec-manuscript-simple-sim) uses a simplified simulation with no treatment effect between groups to understand the variation in results across the full range of parameters. The **second** simulation (@sec-manuscript-exemplar-sim) uses a hypothetical experimental setup with realistic values to establish the extent of the variation in results under realistic conditions. 

### Simple simulation {#sec-manuscript-simple-sim}

```{r sim simple}

p_min <- -1
p_max <- 1
e_min <- -1
e_max <- 1
n = 10000
mu_opioid_requirment_control = 10
sigma = 0.1
treatment_effect = 0
potency_oxycodone = 2

simple_results <- data.frame(p = numeric(), e = numeric(), diff = numeric())
for(p in seq(p_min, p_max, by = 0.05)){
  for(e in seq(e_min, e_max, by = 0.05)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu_opioid_requirment_control,
      potency_oxycodone = potency_oxycodone,
      error_in_c_oxycodone = e,
      treatment_effect = treatment_effect
    )
    
    simple_results <- rbind(simple_results, data.frame(p = p, e = e, diff = diff))
  }
}
```

`r comma(n)` samples were divided equally between a control and intervention group. An opioid of either morphine or oxycodone was then samples from a distribution set by $\Delta p_i$ where $\Delta p_i = p_i^{\text{(control)}} - p_i^{\text{(intervention)}}$. An opioid requirement was sampled from a normal distribution ($\mathbb{E} =$   `r mu_opioid_requirment_control` with $\sigma =$ `r sigma`) for both groups enforcing a treatment effect of `r percent(treatment_effect)`. For samples where the opioid was oxycodone, the opioid requirement was converted to an oxycodone requirement with a ground truth potency equal to `r potency_oxycodone` and then back into an OME mg value using a conversion value with a error term set to $\epsilon$ where $\delta c_i = c_i + (c_i * \epsilon)$. For samples where the opioid was morphine the opioid requirement is equal to the OME mg as it is the reference medication. 

The simulation was then performed for the range of grid values for ($\Delta p_i \in$ `r p_min`, `r p_max` and $\epsilon \in$ `r e_min`, `r e_max` with all other parameters fixed. The result for each study was the measured difference between the group means of the control and intervention groups ($\Delta \mu_{\text{OME}} = \mu_{\text{OME}}^{\text{(control)}} - \mu_{\text{OME}}^{\text{(intervention)}}$).


### Hypothetical study simulation {#sec-manuscript-exemplar-sim}

```{r}
case_volume <- 100
proportion_robotic <- 0.8
t_0 = as.Date("2020-01-01")
t_end = as.Date("2025-12-31")
t_robot = as.Date("2023-01-01")
proportion_morphine = 0.7
oxycodone_shortage = c(as.Date("2024-01-01"), as.Date("2024-12-31"))

mu_tot = 50
sd_tot = 10
potency_oxycodone_truth = 1.85
potency_oxycodone_ome = 2

annual_n = 30
robot_effect_size = 0.90
robot_effect_mg = mu_tot * (robot_effect_size-1)

p_morph_high = 1
p_morph_low = 0
conversion_value_low = 1
conversion_value_high = 3
```

A hypothetical study was designed as outlined in @nte-sim-design and then simulated. 

::: {#nte-sim-design .callout-note title="Simulated trial design" collapse=true appearance="simple"}
A research team wish to understand if the use of a robot for a surgical procedure, reduces the post-operative pain.

The average case volume is `r case_volume` procedures per year, these were all performed using non-robotic surgery until `r format(t_robot, "%e %b %y")` when a robot was acquired, and since `r scales::percent(proportion_robotic)` of cases are performed with the robot. All EHR data was collected between `r format(t_0, "%e %b %y")` and `r format(t_end, "%e %b %y")`.

All patients are treated with a Patient Controlled Analgesia (PCA) after their operation and on average the anaesthetist will prescribe a morphine PCA to `r scales::percent(proportion_morphine)` of patients and an oxycodone PCA to the remainder. No other opioid medication was administered. 

They design a retrospective cohort study for patients treated with a robotic procedure and those without. The primary outcome is the total opioid requirement, as measured in OME mg, in the first 24 hours post-operatively as a surrogate for the total pain caused by the surgical procedure.

During the study there was a national shortage of oxycodone from `r oxycodone_shortage[1]` to `r oxycodone_shortage[2]` and at which time all patients were prescribed a morphine PCA. Whilst the authors described this in their methods section, they did feel it would impact their results as the differing opioids were already accounted for in their OME metric. 
:::

The difference in proportion ($\Delta p_i$) between groups is simulated using a hypothetical medication shortage that due to temporal differences in the groups, effects the treatment group to a greater extent. The ground truth potency of oxycodone is set to `r potency_oxycodone_truth`. The post-operative opioid requirement was sampled from a normal distribution ($\mathbb{E} =$ `r mu_tot` with $\sigma =$ `r sd_tot`) for the non-robotic group with the average treatment effect of the intervention is set to `r robot_effect_mg` OME mg as informed by a power analysis in @sec-appendix-simulation-power. OME was calculated as per described previously.

The simulation was performed once with parameters specified in @nte-sim-design and then across the grid values for the rate of post-operative morphine prescribed between `r percent(p_morph_high)` and `r percent(p_morph_low)` with conversion values of oxycodone:morphine between `r conversion_value_low`:1 and `r conversion_value_high`:1. The result for each study was the measured difference between the group means of the control and intervention groups ($\Delta \mu_{\text{OME}}$), the observed difference in proportion ($\Delta p_i$) and the error in the conversion value ($\epsilon$).  

All simulations were performed in `r version$version.string` and all code is available in @sec-sim. 

## Part 2 - Literature Review

```{r}
# To get screening data:
# Covidence -> Export -> References subsection -> Title and absract screening -> Export as .csv -> Drag to data foler -> Rename to lit_screen.csn
lit_screen <- read.csv("data/lit_screen.csv")
lit_select <- read.csv("data/lit_select.csv")
lit_included <- read.csv("data/lit_included.csv")
lit_excluded <- read.csv("data/lit_excluded.csv")
lit_data <- read.csv("data/lit_data.csv") %>% filter(Reviewer.Name == "Dylan Whitaker")
conversion <- read.csv("data/conversion_sets.csv")

n_lit_data <- nrow(lit_data)

```
A rapid literature review of the use of opioid equivalence metrics in peri-operative and surgical studies published in 2025 was performed. A structured search of the PUBMED database as detailed in @sec-appendix-pubmed was performed on 7th July 2026. Papers were screened and data extracted from a random subset by a single author, using the Covidence platform @zotero-item-670 with a custom proforma (@sec-appendix-cov-extraction). Data extracted included the study design, methodology regarding total opioid metrics and outcomes.

:::{.callout-warning}
Do we need a second author to check a subset of the extracted articles? 
:::

```{r}
papers <- read.csv("data/exemplars.csv") 

excluded_papers <- papers %>% filter(is.na(opioid))

papers <- papers %>% filter(!is.na(opioid)) %>%
  mutate(control_mean_dose = case_when(is.na(control_mean_dose) ~ (control_q1+control_median_dose+control_q3)/3,
                                       .default = control_mean_dose),
         intervention_mean_dose = case_when(is.na(intervention_mean_dose) ~ (intervention_q1 + intervention_median_dose + intervention_q3)/3,
                                       .default = intervention_mean_dose),
         control_sd = (control_q3-control_q1)/1.35,
         control_sd = case_when(is.na(control_sd) ~ (control_q3-control_q1)/1.35,
                                .default = control_sd)
         )%>%
  mutate(Delta_p = intervention_mean_dose - control_mean_dose,
         delta_p = Delta_p/control_mean_dose) 

papers_summary <- papers %>% mutate(control_mean_ome = control_mean_dose*c_i,
         intervention_mean_ome = intervention_mean_dose*c_i) %>%
  group_by(authors) %>%
  summarise(control_total_ome = sum(control_mean_ome),
            intervention_total_ome = sum(intervention_mean_ome)) %>%
  mutate(Delta_ome = intervention_total_ome - control_total_ome,
         delta_ome = Delta_ome/control_total_ome)

exemplars_summary <- readxl::read_excel("data/exemplar_summary.xlsx")

exemplars_summary <- exemplars_summary %>%
  mutate(reported_Delta_ome = reported_intervnetion_group_mean - reported_control_group_mean,
         reported_delta_ome = reported_Delta_ome/reported_control_group_mean)

```

```{r}
# 1. Run the independent grid search (if not already done)
library(tidyr)
plot_data <- papers %>%
  group_by(authors) %>%
  group_modify(~ {
    current_data <- .x
    n_drugs <- nrow(current_data)
    
    # Map each drug to its range from the ranges list
    grid_list <- lapply(current_data$opioid, function(op) {
      # Match opioid name to ranges key (case-insensitive and stripping ' (assumed)')
      clean_op <- tolower(gsub(" \\(assumed\\)", "", op))
      if (clean_op %in% names(neilsen_ranges)) {
        neilsen_ranges[[clean_op]]
      } else {
        # Fallback to the existing c_i value if the opioid isn't in neilsen_ranges
        unique(current_data$c_i[current_data$opioid == op])
      }
    })
    
    names(grid_list) <- paste0("m_", 1:n_drugs)
    
    expand_grid(!!!grid_list) %>%
      rowwise() %>%
      mutate(
        control_total_ome = sum(current_data$control_mean_dose * c_across(starts_with("m_"))),
        intervention_total_ome = sum(current_data$intervention_mean_dose * c_across(starts_with("m_")))
      ) %>%
      ungroup()
  }) %>%
  mutate(
    Delta_ome = intervention_total_ome - control_total_ome,
    delta_ome = (Delta_ome) / control_total_ome
  )

papers_grid_summary <- plot_data %>%
  group_by(authors) %>%
  summarise(min_Delta = min(Delta_ome),
            max_Delta = max(Delta_ome),
            min_delta = min(delta_ome),
            max_delta = max(delta_ome)) %>%
  left_join(papers_summary%>%select(authors, delta_ome), join_by(authors))

```

For each paper where group average doses of each opioid were described, these were extracted. If only median values were provided, an estimated mean was calculated using Wan's 2014 approximation ($\hat{d_i} = \frac{Q_1 + M + Q_2}{3}$ ) to preserve the linearity with in the arithmetic mean @wanEstimatingSampleMean2014. Where opioid routes where not provided then IV was assumed and where a value set was not provided then the Nielsen set was used @nielsenSynthesisOralMorphine2016. 

The group mean of total opioid burden reported by the paper was extracted. Then using the mean doses and the conversion values provided, the group mean was calculated independently. Following this, all possible group means were calculates by applying each of the possible values provided in Table 1 of Neilsens Conversion set (summarised in @sec-appendix-neilsen-ranges), as the range of plausible conversion values an author may use @nielsenSynthesisOralMorphine2016.

# Results

## Part 1 - Simulation study 

### Simple simulation {#sec-manuscript-results-simple}

`r comma(nrow(simple_results))` different simulations where performed across the grid values of $\Delta p_{\text{oxycodone}}$ and $\epsilon$. The observed difference across the grid values demonstrates a normal distribution, centred around the true treatment effect of `r treatment_effect` (@fig-simple-density).

```{r}
#| label: fig-simple-density
#| fig-cap: "**Distribution of observed group mean difference across the simple simulation** -Density plot of the observed difference in group means (ΔOME) across all grid parameter combinations. The dashed red line indicates the true treatment effect of zero."

ggplot(simple_results, aes(x=diff))+
  geom_density(fill="blue", alpha = 0.8)+
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05)), limits = c(-10, 10)) +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed")+
  labs(y = "Density",
       x = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) 

```

When plotted against the grid values, it becomes clear that when the proportion of medication in the control and intervention groups are matched ($\Delta p_i = 0$), then the true effect size of 0 is recovered, regardless of any error in the conversion value ($\delta c_i$) used (@fig-simple-grid-1). Furthermore, where the conversion values used are correct (with respect to the ground truth), any difference in proportion is accounted for and the true effect is again recovered.

However, as the proportions between the control and treatment group deviate, resulting difference observed is biased by any error in the conversion values. Where the groups are completely unmatched ($\Delta p_i =$ 1 or -1) the resulting observed difference is the due entirely to the error in potency.

```{r}
#| label: fig-simple-grid-1
#| fig-cap: '**Observed OME group mean difference across opioid proportion difference and conversion value error** - Heatmap of the observed difference in group means ($\Delta \mu_{\text{OME}}$) across grid values of $\Delta p_i$ (difference in opioid proportion between groups) and $\epsilon$ (error in the oxycodone conversion value). Bias is zero when either parameter is zero and increases as both deviate from zero.'
#| fig-height: 6
#| fig-width: 6

plot_simple_grid <- ggplot(simple_results, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
 # scale_fill_gradientn(
#  colours = c("blue", "dodgerblue", "white", "orange", "red"))+
#  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
#  geom_hline(yintercept = 0, linetype = "dashed", color = "black")+
#  annotate("text",x=-0.7, y=0.05, label = "Proportion of medication equal")+
#  annotate("text",x=-0.7, y=0.95, label = "Control group: 100% morphine\nIntervention group: 0% morphine")+
 # annotate("text",x=-0.7, y=-0.95, label = "Control group: 100% oxycodone\nIntervention group: 0% oxycodone")+
  #annotate("text",x=-0.05, y=-0.7, label = "No error in conversion value", angle=90)+
  #annotate("text",x=0.7, y=-0.05, label = "Overestimate potency ->" )+
  #annotate("text",x=-0.7, y=-0.05, label = "<- Underestimate potency")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  symmetric_fill_scale(round(max(simple_results$diff))+0.01)

plot_simple_grid

```

@fig-simple-grid-1 demonstrates that the observed difference can be described with a rotated hyperbolic parabola with the scaling constant $k$.

$$
\Delta \mu_{\text{OME}} = k \cdot \Delta p_i \cdot \epsilon
$$ {#eq-results-simple-1}

@sec-appendix-sim-1 demonstrates that the scaling constant ($k$) is the expected dose for the control group as demonstrated in @eq-results-simple-2-proof). @sec-appendix-sim-2 demonstrates that as the true treatment effect acts to both translate and scale the observed treatment effect. 

### Hypothetical study simulation

```{r }
df <- sim_patient(
  mu_tot = mu_tot,
  potency_oxycodone_truth = potency_oxycodone_truth,
  potency_oxycodone_ome = potency_oxycodone_ome,
  proportion_morphine = proportion_morphine,
  oxycodone_shortage = oxycodone_shortage,
  t_0 = t_0,
  t_end = t_end,
  t_robot = t_robot,
  annual_n = annual_n,
  robot_effect_size = robot_effect_size
)

```

`r nrow(df)` procedures were simulated using the parameters described in @sec-manuscript-exemplar-sim. 

```{r}
#| label: tbl-sim-demo
#| tbl-cap: Demographic table of simulated study population
#| eval: true

library(gtsummary)

df |>
  tbl_summary(
    by = robotic,
    include = c(opioid, dose, ome),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      dose ~ "Dose (mg)",
      opioid ~ "Opioid",
      ome ~ "OME (mg)"
    ),
    digits = all_continuous() ~ 1
  ) |>
  add_overall() |>
  add_p() |>
  as_gt() |>
  gt::tab_options(table.width = gt::pct(100))
```

A simple demographics table for the study is shown in @tbl-sim-demo, this demonstrates the unequal distributions of opioids between groups. @fig-sim-ome-1 demonstrates this distribution of the total opioid burden (OME mg) of patients in the robotic and non-robotic patients. @fig-sim-ome-2 shows that within the robotic and non-robotic groups the proportion of each opioid is not equal, and the effect of the oxycodone shortage has had a larger impact on the robotic group.

```{r}
#| label: fig-sim-ome
#| layout-ncol: 2
#| fig-cap: "**Opioid prescribing and OME distribution by surgery type**"
#| fig-subcap: 
#|   - "**OME distribution by surgery type** - Kernel density of calculated OME (OME mg) for robotic and non-robotic surgery groups."
#|   - "**Opioid prescribing proportions by surgery type** - Proportion and count of each opioid prescribed within robotic and non-robotic surgery groups."
#| fig-width: 6
#| fig-height: 6

ggplot(df, aes(x=ome, fill=robotic))+
  geom_density(alpha=0.8)+
  theme_academic()+
  labs(x="OME (OME mg)",
        y="Density",
        fill="Robotic Surgery")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme(legend.position = "bottom")+
  paletteer::scale_fill_paletteer_d("MetBrewer::Austria", direction = -1)

df %>%
  count(robotic, opioid) %>%
  group_by(robotic) %>%
  mutate(proportion = n / sum(n)) %>%
  ggplot(aes(x = robotic, y = proportion, fill = opioid)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = paste0("n=",n)),
    position = position_dodge(width = 0.9),
    vjust = -0.4
  )+
  labs(y = "Proportion", x = "", fill = "Opioid") +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent_format()) +
  theme(legend.position = "bottom")+
  paletteer::scale_fill_paletteer_d("MetBrewer::Austria")

```


```{r}
obs_robotic_ome <- df %>% filter(robotic == "Robotic") %>% summarise(mean(ome)) %>% pull() 
obs_non_robotic_ome <- df %>% filter(robotic != "Robotic") %>% summarise(mean(ome)) %>% pull()
obs_robot_effect_mg <- obs_robotic_ome - obs_non_robotic_ome

result <- analyse_dataset(df)

if (PRODUCTION_QUALITY) {
  trials = 50
} else {
  trials = 3
}

```

The mean opioid burden value for the robotic surgery group was `r round(obs_robotic_ome, 1)` OME mg and for the non-robotic surgery group was `r  round(obs_non_robotic_ome, 1)` OME mg. This gives an observed effect size of the robot as `r round(obs_robot_effect_mg,1)` OME mg. Using a two sided t-test to assess if there is a significant difference in mean OME between the robotic and standard surgery groups, the mean difference was `r round(result$mean_diff_ome,1)` OME mg (95% CI `r paste(round(result$conf_int_ome, 1), collapse = " to ")`) with a p-value of `r pvalue(result$p_value_ome)`. Given the study was adequately powered (@sec-appendix-simulation-power), the research team were then able to conclude that robotic surgery is associated with a significant difference in total opioid burden. 

### Impact of conversion set and proportions

The hypothetical study was repeated using the grid values of different conversion values and different proportion of post-operative oxycodone prescribing. The prior was informed by ranges observed in the literature and the latter to influence the degree of different in opioid proportions between groups. For each set of grid values, `r trials` were performed and the mean observed difference and median Fishers Exact p-value extracted. Given a `r mu_tot` mg expected opioid requirement and a treatment effect of `r scales::percent(robot_effect_size-1)` mg we would expect the mean difference in OME of the robotic group to be `r (mu_tot * robot_effect_size)-mu_tot` mg OME.

```{r}
#| label: fig-sim-grid-search-1
#| fig-cap: "**Sensitivity of observed treatment effect to conversion value and opioid proportion** - Grid search results showing the observed treatment effect across oxycodone conversion values and post-operative morphine prescribing proportions. Purple cross denotes an median p value of Fishers Exact test <0.05. True treatment effect is -5 mg. The dashed line indicates the ground truth conversion value."
#| fig-height: 4
#| fig-width: 8
#| out-height: 3in
#| out-width: 6in


simulation1 <- run_simulation(trials=trials,
               conversion_values = seq(from=conversion_value_low, to=conversion_value_high, by = 0.1),
               morphine_proportion = seq(from=p_morph_low, to= p_morph_high, by = 0.05)
               )

plot_grid_1 <- plot_grid(simulation1,
          potency_oxycodone_truth = potency_oxycodone_truth,
          robot_effect_mg = 0)

plot_grid_1

```

The results of the `r comma(nrow(simulation1)*trials)`  simulations are shown in @fig-sim-grid-search-1.

Where the proportion of post-operative morphine prescribing is high, the exogenous effect of an oxycodone shortage has little impact to diverge the proportions of opioids between groups , and the true treatment effect of `r (mu_tot * robot_effect_size)-mu_tot` mg OME is recovered. However, as the impact of the oxycodone shortage causes further divergence of the proportions between groups there is wider variation in the observed difference. Where the conversion value supplied is close to the ground truth (indicated by the red line) then the error is low however a statistically significant result may not be reached. At the extreme, there are values where the observed difference is +`r round(max(simulation1$mean_diff_ome),2)` mg, inverting the direction of the treatment effect. 

```{r}
#| label: fig-compare
#| fig-cap: "**Exemplar simulation parameters mapped onto the simple simulation space** - Side-by-side comparison of the simple and exemplar simulation grids. The region explored by the exemplar experiment (Δp_i < 0; −0.5 < ε < 1) is overlaid on the full simple simulation space, showing that even within this narrow parameter region a modest effect size produces meaningfully variable observed outcomes."
#| fig-width: 8
#| fig-height: 4


label_function <- function(val) {
  paste0('mu["OME"]^"(control)" == ', val )
}

bounds <- c(min(simulation1$delta_p_oxycodone),
            max(simulation1$delta_p_oxycodone),
            min(simulation1$epsilon),
            max(simulation1$epsilon))

simple_results <- data.frame(p = numeric(), e = numeric(), diff = numeric())
for(p in seq(p_min, p_max, by = 0.05)){
  for(e in seq(e_min, e_max, by = 0.05)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu_tot,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = robot_effect_size -1
    )
    
    simple_results <- rbind(simple_results, data.frame(p = p, e = e, diff = diff))
  }
}


plot_simple_grid <- ggplot(simple_results, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "none")+
  symmetric_fill_scale(20)+
  geom_rect(ymin=bounds[1],
            ymax=bounds[2],
            xmin=bounds[3],
            xmax=bounds[4],
            colour="black",
            fill=alpha("grey",0))
 # annotate("text", label = "Parameter space of exemplar simulation", x=mean(bounds[3:4]), y=bounds[2]+0.1)

plot_grid_2 <- ggplot(simulation1, aes(x=conversion_value, y=morphine_proportion, fill=mean_diff_ome)) +
  geom_tile() +
  symmetric_fill_scale(20)+
  labs(y = TeX(r'(Proportion of post-op perscribing as morphine )'),
       x = TeX(r'($\c_{oxycodone}$ (Conversion Value for Oxycodone used) )'),
       fill = "Mean difference in OME") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
  scale_y_continuous(position = "right", labels = scales::percent)+
  geom_vline(xintercept = potency_oxycodone_truth, linetype = "dashed", color = "red")
  
layout <- c(
  area(t = 1, l = 1, b = 10, r = 10),
  area(t = 6, l = 11, b = 10, r = 20)
)

plot_simple_grid + plot_grid_2 +
  plot_layout(design = layout)
```

@fig-compare demonstrates that the experiment space for the example simulation values where $\Delta p_i < 0$ and for values of $-0.5 < \epsilon < 1$. Despite being a relativity narrow section of the possible gird space detailed in @sec-manuscript-results-simple with a modest effect size of `r robot_effect_mg` OME mg there are still values where the observed effect size is significantly different. 


## Part 2 - Literature Review 

### Reporting of opioid burden metrics

480 studies were identified in the rapid literature search, a sample of `r nrow(lit_data)` that fit the eligibility has data extracted. The majority of studies were cohort studies and most commonly looked at anaesthetic type interventions (@fig-study-types).

```{r}
#| label: fig-study-types
#| fig-cap: "**Study design and intervention types in included literature** - Bar charts summarising the study design and intervention type of studies included in the rapid literature review. Left panel shows count by study design; right panel shows count by intervention type."

# Plot for Study.type proportions
plot_lit_study_type <- ggplot(lit_data, aes(x = Study.type)) +
  geom_bar(aes(fill = Study.type)) +
  labs(x = "Study design",
       y = "Count") +
  theme_minimal() +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_academic()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# Plot for Intervention proportions
plot_lit_intervnetion_type <- ggplot(lit_data, aes(x = Intervention)) +
  geom_bar(aes(fill = Intervention)) +
  labs(x = "Intervention",
       y = "Count") +
  theme_minimal() +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_academic()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

plot_lit_study_type + plot_lit_intervnetion_type

```

With regards to the opioid conversion sets used, `r lit_data %>% filter(Opioid.conversion.set == "Cited") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.conversion.set == "Cited") %>% nrow()/n_lit_data)`) cited another source, whilst `r lit_data %>% filter(Opioid.conversion.set == "Included in text") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.conversion.set == "Included in text") %>% nrow()/n_lit_data)`) included the conversion values within the paper. `r lit_data %>% filter(Opioid.conversion.set == "Not described") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.conversion.set == "Not described") %>% nrow()/n_lit_data)`) papers did not include any citation or description of the opioid conversion sets used. Upon review of the the results sections, it was found that only `r lit_data %>% filter(Opioid.proportions.within.group == "Described") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.proportions.within.group == "Described") %>% nrow()/n_lit_data)`) papers described the proportions of different opioids between intervention groups (@fig-study-results). Of these `r nrow(lit_data %>% filter(Opioid.proportions.within.group == "Described"))` papers, `r nrow(excluded_papers)` provided count data only and not mean/median dose for each opioid within treatment groups. 

```{r}
#| label: fig-study-results
#| fig-cap: "**Opioid conversion set reporting and within-group proportion documentation** - Heatmap showing the count of studies by opioid conversion set type and whether opioid proportions within groups were reported. Cell counts are displayed as text labels."

# Prepare data for heatmap: sum of occurrences per combination
heatmap_data <- lit_data |>
  group_by(Opioid.conversion.set, Opioid.proportions.within.group) |>
  summarise(count = n(), .groups = "drop")

# Prepare data with a conditional color column for labels
ggplot(heatmap_data, aes(y = Opioid.proportions.within.group, x = Opioid.conversion.set, fill = count)) +
  geom_tile() +
  geom_label(aes(label = paste0("n=",count),
                 color = after_scale(ifelse(fill > 0.5, "black", "white"))),
             fill = alpha("white", 0.7), size = 3.5) +
  labs(y = "Proportion of opioids within groups",
       x = "Opioid conversion set",
       fill = "Count") +
  theme_minimal()+
 theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(angle = 45),
    panel.grid = element_blank(),
    legend.position = "none")+
  symmetric_fill_scale(max(heatmap_data$count))
```

### Impact of varying conversion values on studies

@tbl-lit-summary details the `r length(unique(exemplars_summary$Ref))` studies where opioid proportions could be recovered. Of note, only Baran cited a opioid conversion set @baranOptimizingAnalgesiaMinimally2025, all other authors did not describe which value sets were used. Two studies were randomised controlled trials (RCT), whilst the remainder were retrospective cohort studies. With regards to the interventions analysed, in this group, all were concerning analgesia either during or after the surgical procedure. 

```{r}
#| label: tbl-lit-summary
#| tbl-cap: "Summary table of litrature articles that presented breakdown by opioid groups."

exemplars_summary <- exemplars_summary %>%
  left_join(papers_summary, by="authors") %>%
  left_join(papers %>% select(authors, authors), join_by(authors), multiple = "first") %>%
  left_join(papers_grid_summary %>% select(-delta_ome), by = "authors") %>%
  mutate(reported_diff = paste0(round(reported_Delta_ome,1) , "mg (", percent(reported_delta_ome, accuracy = 0.1), ")"),
         calculated_diff = paste0(round(Delta_ome, 1), "mg (", percent(delta_ome, accuracy = 0.1), ")"),
         min_diff = paste0(round(min_Delta, 1), "mg (", percent(min_delta, accuracy = 0.1), ")"),
         max_diff = paste0(round(max_Delta, 1), "mg (", percent(max_delta, accuracy = 0.1), ")")
         ) 
  
exemplars_summary_table <- exemplars_summary %>%
  mutate(Ref = as.integer(Ref),
         authors = case_when(Ref==303 ~ "Gold 2025",
                             .default = authors)) %>%
  left_join(lit_data, join_by("Ref" == "Covidence..") ) %>%
  select(authors, note, Title, Study.type,population, control_group, intervention_group, Opioid.conversion.set) %>%
  unique() 

gt_table_data <- exemplars_summary_table %>%
  filter(authors != "Gold 2025") %>%
  select(-note)

# Identify rows with notes for footnotes
note_indices <- which(!is.na(exemplars_summary_table$note) & exemplars_summary_table$note != "")
notes_to_add <- exemplars_summary_table$note[note_indices]

tbl <- gt_table_data %>%
  gt() %>%
  cols_label(
    authors = "Authors",
    Title = "Title",
    Study.type = "Study design",
    population = "Population",
    control_group = "Control group",
    intervention_group = "Intervention group",
    Opioid.conversion.set = "Opioid conversion set included"
  )

# Programmatically add footnotes
for (i in seq_along(note_indices)) {
  tbl <- tbl %>%
    tab_footnote(
      footnote = notes_to_add[i],
      locations = cells_body(columns = authors, rows = note_indices[i])
    )
}

tbl %>%
  opt_footnote_marks(marks = "numbers") %>%
  tab_options(table.font.size = "small",
              container.width = pct(100)) %>%
  cols_width(
    Title ~ pct(40),
    population ~ pct(20)
  )

```


@tbl-lit-summary-outcomes provides three sets of results for the studies detailed. The first is the study reported outcomes, this is the mean total opioid burden for the control and intervention groups. From this an observed difference is calculate with a percentage change from the control. The second column set is the de-novo calculating, by using the mean dose of each opioid within each group and conversion set provided (or Neilsens if not) to attempt to calculate the control and intervention group means. From these de-novo estimates a second difference of the treatment group was calculated. 

Finally, regardless of the conversion sets used, the de-novo calculation was repeated across the possible grid values of the Neilsen conversion set. The minimum and maximum possible differences were then extracted. 

```{r}
#| label: tbl-lit-summary-outcomes
#| tbl-cap: "Table of reported outcomes and de-novo calalculated outcomes."


exemplars_summary_table <- exemplars_summary %>%
  select(authors,
         reported_control_group_mean, reported_intervnetion_group_mean, reported_diff,
         control_total_ome, intervention_total_ome, calculated_diff,
         min_diff, max_diff)

exemplars_summary_table %>%
  gt() %>%
  tab_spanner(
    label = "Study reported outcome",
    columns = c(reported_control_group_mean, reported_intervnetion_group_mean, reported_diff)
  ) %>%
  tab_spanner(
    label = "De Novo calculated outcome",
    columns = c(control_total_ome, intervention_total_ome, calculated_diff)
  ) %>%
  tab_spanner(
    label = "Range of estimates",
    columns = c(min_diff, max_diff)
  ) %>%
  cols_label(
    authors = "Authors",
    reported_control_group_mean = "Control group",
    reported_intervnetion_group_mean = "Intervention group",
    reported_diff = "Difference",
    control_total_ome = "Control group",
    intervention_total_ome = "Intervention group",
    calculated_diff = "Difference",
    min_diff = "Lower bound",
    max_diff = "Upper bound"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = c(reported_diff, calculated_diff))
  ) %>%
  fmt_number(
    columns = where(is.numeric),
    decimals = 1
  ) %>%
  tab_options(
  table.font.size = "small",
  container.width = pct(100)
  )
```

This table highlights the difficulty in re-calculating the total opioid burdens that the authors have produced. This is likely due to a number of assumptions as details in the methods such as estimating mean values from medians and assuming medication routes. Only one study, Fedorka 2025, showed no sensitivity to the conversion value set used @fedorkaLiposomalBupivacaineBupivacaine2025. This is because, in the RCT only oxycodone was used in the post-operative period, as so whilst the group estimates may vary, the treatment effect percentage is fixed at 10%.  


@fig-lit-outcomes demonstrates the reported difference in opioid burden (red cross) compared to the de-novo calculated difference (green cross). When the the grid search of the possible combinations of the values used by Neilsen was performed, the minimum and maximum values are demonstrated by the orange whiskers with the distribution within each given in @fig-lit-outcome-dist.

```{r}
#| label: fig-lit-outcomes
#| fig-cap: "Point and whisker plot of the reported and de-novo calcualted difference in the intervention group for opioid outcomes. Red cross shows the reported percentage intervention group outcome, green cross's the de novo point estiamte using hte values reported by neilsen and the oragne error bar the range using the grid values across the ranges reproted by Neilsen. All effects given as the percentage change in the intervention group compared to the control."

ggplot(exemplars_summary, aes(y=authors))+
  geom_errorbar(aes(xmin=min_delta, xmax=max_delta), color = "orange")+
  geom_point(aes(x=reported_delta_ome),colour = "red", shape = 4)+
  geom_point(aes(x=delta_ome),colour = "green", shape=3)+
  geom_vline(xintercept = 0, color = "black") +
  theme_academic()+
  theme(axis.line.y = element_blank())+
  labs(x="Change in opioid use in intervention group",
       y="")+
  scale_x_continuous(labels = percent,
                     breaks = seq(-0.6, 0.3, by = 0.1))

```

### Conversion values used across the literature

Of the `r lit_data %>% filter(Opioid.conversion.set == "Cited") %>% nrow()` papers where a conversion value set was cited, the `r nrow(conversion)` unique source value sets where reviewed. Generally across the literature, conversion values with the intention of research were in the minority (`r percent((nrow(conversion %>% filter(aim == "for research")))/(nrow(conversion)))`) and most had compiled the conversion value sets for use in clinical care. 

:::{.callout-warning}
Can expand this section but not sure if its worth it?
:::


# Discussion

By defining and re-arranging the equations for total opioid burden and group average opioid burden, we have shown that any difference in proportion or error in the conversion value introduces an error in the observed difference measured. This relatonships can be mapped to hyperbolic parabola that is shaped by the underlying treatment effect and the size of the group means. The saddle geometery raises an important issue in comparing group means; if a conversion factor for the medication $i$ is over-estimated ($\epsilon_i > 0$) and prevalence is simultaneously higher in the intervention group ($\Delta p_i > 0$), the error in the treatment effect compounds quadratically. Conversely, opposing directional errors will collapse and at the extreme values may invert the direction of the observed treatment effect.

This phenomenon could be explained by Simpsons paradox, whereby the observed treatment effect is conditioned on the proportion of the the medications within each group, thus creating the observation of a false or biased treatment effect @pearlBookWhyNew2019. 

Across the literature, the methods and results describing the total opioid burden metrics is generally poor. It was found that only `r scales::percent(lit_data %>% filter(Opioid.proportions.within.group == "Described")  %>% filter(Opioid.conversion.set != "Not described") %>% nrow()/n_lit_data)` of papers reported both the conversion value sets used and the proportions of opioids between groups. Re-creating the opioid burden metrics from the limited papers where this was possible, showed signficiant differences between the author reported results and the de novo calculation. 

It was noted that a number of papers sourced their perceiving information from the Pennsylvania Prescription Drug monitoring program (PPDMP), of which none of these papers cited the conversion values used @adalbertPrescriptionDrugMonitoring2023. Upon investigation further, the technical notes from the PPDMP describe MME values are calculated using conversion factors, but they are not stated or detailed further @zotero-item-700. Whilst datasets that contain pre-calculated MME values offer an attractive data source, it should be made clear how derived metrics are calculated. With the increase of EHR and digital datasets, this need for clear documentation and transparency is becoming increasingly important to ensure the metrics used are true and accurate reflections of the underling data \[Could cite JMIR paper here.\]. 


## Limitations

In the simulation studies we have only considered situations where medication choice is binary between the reference medication and the alternative (oxycodone). We hypothesise that in some situations with multiple opioids, due to orthogonality, that multiple small errors in the estimate of $c_i$ and the true value of $c_i$ may generally cancel out and thus have little impact on the final estimate. This may be achieved using machine learning tools that consider the co-linearity between the features of the dose of each medication and the outcome of total opioid burden or OME. 

This work is particular relevant to retrospective observational work where large datasets with varied treatment lend themselves well to the tool of opioid conversion values. However, even in a RCT where the protocol for analgesia may be fixed and patients randomised between groups, confounding may occur if analgesia protocols contain differing opioids allowing the possibility of diverging proportions between treatment groups. However, the degree of error introduced is likely if a high quality randomisation method is used. 

In our simulations we have used a simulated "ground truth" value for oxycodone as this cannot be measured. The broad range of conversion values produced in the literature has been primarily generated for clinical care. By co-opting these general rule-of-thumb measures authors are not able to quantify or communicate any level of uncertainty.

The literature search was intended to provide a snapshot of current academic practice and hence was not exhaustive by only using a sample of the papers in 2025 with an intentionally specific search criteria. Thus this rapid lit review may not be fully representative of opioid burdens metrics used across the academic domain.  
 
# Conclusion

We have demonstrated that the use of total opioid burden metrics such as OME or MME are sensitivity to the proportions of different opioids and the conversion sets used. We have demonstrated that the difference measured between the observed $\mu_{\text{OME}}^{(A)}$ and $\mu_{\text{OME}}^{(B)}$ is a product of not only the association of the two groups (be it statistical or causal), but also the difference in proportions of medications within each group.

In addition, the opioid equivalence metric is also sensitivity to any error in the true conversion value of the underlying population and the conversion value used for calculations. As expressed across the literature, it is difficult to ascertain all of the inter-patient difference and inter-opioid interactions that would provide the true conversion value. None the less, any difference in the true population level conversion value and the values used will also lead to systematic under or over estimation of the true value. 

## Further work

In retrospective work it will not be possible to alter the proportions of different opioids within groups. Further work is needed to understand if a new formula for the total opioid burden can be generated that will account or control this. 

Furthermore, new methods are needed to attempt to estimate the true potency values of different opioids. A data driven methodology to attempt to derive the conversion values de-novo may offer a promising solution given the relatively small evidence base that all current conversion values are based upon. Additionally a baysian approach may allow for quantification of the uncertainty within the conversion values. 

## Recommendations:

1. All studies using opioid equivalence metrics must publish the proportions of the differing opioids and conversion value sets used for each group. 

2. More work is required to estimate the true population level opioid conversion value sets to reduce the error between the true unmeasurable potency and the value sets used. 

:::{.callout-warning}
The further work and recommendations are rather similar and could be consolidated. 
:::

## Funding statement

DW is supported by the Engineering and Physical Sciences Research Council (EP/Y034929/1). MW and SH are supported by the National Institute for Health Research (NIHR) Central London Patient Safety Research Collaboration (NIHR 204297). SH and DW are supported by the NIHR University College London Hospitals National Health Service Foundation Trust and University College London Biomedical Research Centre. The views expressed in this study are those of the authors and do not necessarily reflect those of the NIHR or the Department of Health and Social Care.

## Acknowledgements

The authors wish to thank the following the for their support. Dr Thomas Frost for early direction and support. University College London Hospitals National Health Service Foundation Trust Critical Care Special Interest Project Scheme.

Generative artificial intelligence tools where used to aid the production of code, figures and tables. These tools included Gemini 3.1 models (Pro, Flash and Flash-Lite), Claude sonnet 4.6 (via Posit AI) and Github Copilot (via Posit Code Assistant). All ideas and methods are of the authors own, all text and code was reviewed and edited following generation.

## CRediT statement

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: asis
#| eval: false

tbl_authors <- googlesheets4::read_sheet("123JiCxSHvA7Wiaz6tU25B7Wq27Ty_cV7n6UUqlhyvrY",sheet = 2)
aut <- plume::Plume$new(tbl_authors, dotted_initials = F)
aut$get_contributions(roles_first = T)
```

Conceptualization: DW, MW and SH
Data curation: DW
Formal analysis: DW
Funding acquisition: MW and SH
Methodology: DW
Supervision: MW and SH
Visualization: DW
Writing - original draft: DW
Writing - review & editing: DW, MW and SH

::: {.content-visible when-format="pdf"}
\listoffigures
:::