---
title: "Manuscript"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r source_files}
#| echo: FALSE
#| message: FALSE
#| output: FALSE

# Source the founctions in simulaton.qmd with a temporary R script containing only code blocks
temp_r_file <- tempfile(fileext = ".R")
knitr::purl("simulation.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
knitr::purl("index.qmd", output = temp_r_file, quiet = TRUE)
source(temp_r_file, echo = FALSE)
unlink(temp_r_file)
rm(temp_r_file)
```

```{r set theme}
#| warning: false
#| message: false

library(scales)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(patchwork)

theme_academic <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black"),
      text = element_text(family = "sans"),
      plot.title = element_text(face = "bold", size = 14),
      strip.background = element_rect(fill = "white", color = "black")
    )
}
# Add scale_y_continuous(expand = expansion(mult = c(0, 0.05))) to plots 
# to ensure the Y axis starts at 0 and bars touch the X axis.

scale_fill_academic <- function() {
  scale_fill_viridis_d(option = "magma", begin = 0.3, end = 0.8)
}

```


## Origin of opioid equivalence and eqipotency

Opioid medications are an important pillar of analgesia in medicine, used across a range of specialities and patients. There are a wide range of opioids, some naturally derived with others synthetic, each with their own potency. Whilst a wide range of factors may influence the potency of a specific opioid in a specific patient, there has long been a need to calculate the eqi-analgesic dose, or the dose of a medication that would provide an equal amount of analgesia compared to a reference. 

Much of the literature on opioid equivalence comes from the domain of palliative care in the context of opioid switching. McPherson described a range of limitations in the use of equialangesic tables including the source of the data, lack of patient specific variables and bi-directionality of equivalence values. @mcphersonWhyEquianalgesicTables2020.

In a clinical context, prescribers will use often use a single round conversion value for all patients. This provides a range of benefits including medication safety and reducing the cognitive burden in high stress environments. 

## Opioid equivalence in research

With the increased of EHR data offering a rich sources of medication data, opportunities to now use total opioid use as either a feature or an outcome have become more prevalent. In the last 5 years there have been 4,031 papers using the terms *"Milligram morphine equivalent"* (MME) or *"Oral morphine equivalent"* (OME) (see @sec-appendix-pubmed-1). 

The use of this metric has likely arisen due to the need to calculate the total opioid burden across a range of possible treatment s, and the need to compare between groups. In 2016 Neilsen et al published on the first formal OME conversion tables @nielsenSynthesisOralMorphine2016. They sources a range of interntail clinical guidleins and recommended a single value for each opioid and route with a upper and lower quartile. 

It was clear in the literature that using a range of different conversion sets would not produce equitable or comparable resultsm in 2022 the CDC 

## Section that when consistent with the conversion set, it is assumed that any differences in drug or route are accounted for.

An important aspect of the OME metric that has not been highlighted in the literature is the dependence on the proportion of each medication within each group.


## Opioid equivalence in clinical practise

Often in clinical practice the term potency is used to describe the relative strength of one opioid compared to another. The potency of a medication is commonly understood to be the dose required to achieve equal effect, and in the case of opioids, the equianalgesic effect. Thus @eq-potency describes the relationship between the conversion value $c$ and the potency $\rho$ of a given medication $i$ via a given route $r$.

$$c_{i,r} = \frac{1}{\rho_{i,r}}
$$ {#eq-potency} 

The value of $c$, then allows for the calculation of the equivalent effective dose of a target medication $d_{\text{target}}$, given a dose of a reference medication $d_{\text{referance}}$, as shown in @eq-equivelent-dose. 

$$
d_{\text{referance}} \times c_{i,r} = d_{\text{target}}
$$ {#eq-equivelent-dose}

A worked example of this is process of opioid conversion calculation is provided in @sec-conversion-example. This approach is used clinically for opioid conversion or rotation and is the origin of the opioid equivalence system. As discusses above, with a small set of opioid equivalence values, it's possible to convert doses between a wide range of different opioids. 

## Opioid equivalence in research

This technique of comparing opioid equivalent doses has lead to the development of a new metric, the sum opioid burden.  Assuming that pain is well controlled, this the sum of the doses of oral morphine equivalence is a measure of the opioid burden allows for the estimation of the total pain a patient is experiencing as shown in @eq-pain-burden. 
$$
\begin{alignat}{2}
& & \mathit{Pain}_{\text{total}} &= \mathit{Pain}_{\text{experienced}} + \text{Analgesia} \\
& & \text{Opioid burden} &\approx \text{Analgesia}  \\[1em]
&\text{Where: } \qquad & \mathit{Pain}_{\text{experienced}} &\approx 0 \\
&\text{then: } & \mathit{Pain}_{\text{total}} &\approx \text{Analgesia}
\end{alignat}
$$ {#eq-pain-burden}

The ability to now approximate the total amount of pain a patient is experiencing is a powerful metric for research as there is now a easy to collect, specific and quantified value to estimate the amount of pain a patient is in. This offers an attractive primary outcome for any study looking to reduce the pain experienced though a surgical procedure. 

## Defining OME

To provide a value with a unit a reference medication is selected to ground the conversions. This is usually expressed in units of equivalent dose of oral morphine with the units of OME mg^[The use of oral morphine as the reference medication is likely due to the background in palliative care and the availability/use of oral morphine across healthcare systems]. 

The opioid burden metric is often presented as the weighted sum of opioids administered. We can define $\text{OME}_j$ as the sum of the dose $d$ for medication $i$ in the set of medications $\mathbf{M}$ multiplied by the conversion value $c$ for a given route $r$ for patient $j$ as shown in @eq-ome.^[Medications administered in different dose units are generally converted to a single dose unit (often **mg**) and the conversion values updated to reflect this but in some cases they may be kept in their administered unit (**e.g. mcg**) and the conversion shifted by an appropriate factor.] $\text{OME}_j$ can often be expressed as a total of all opioids administered or within a specific time frame (*e.g.* 24 hours post-operative).

$$\text{OME}_j = \sum_{i\in\mathbf{M}} d_i c_{i,r}
$${#eq-ome}

Whilst $\text{OME}_j$ provides an estimate of a specific patients opioid burden, for use within research, further statistical tools are often applied for hypothesis testing. Given a likley parametric distribution, the arithmetic mean of a groups OME value $\mu_{\text{OME}}$ is often calculated as the sum of OME values in the group $G$ divided by the group size as shown in as @eq-mu-ome-1.
$$
\mu_{\text{OME}}^{(G)} = \frac{1}{|G|} \sum_{j=G} \text{OME}_j
$${#eq-mu-ome-1}

As the mean OME is a **mean of weighted sums** it can instead be expressed as a **weighted sum of mean doses** whereby the mean dose of a given medication within the group $\bar{d_i}$ is weighted by the conversion values $c_{i,r}$ as shown in @eq-mu-ome-final. The rearrangement is demonstrated in @sec-equation-proof.

$$\mu_{\text{OME}}^{(G)}  = \sum_{i\in\mathbf{M}} \bar{d_{i}} c_{i,r} 
$${#eq-mu-ome-final}

@eq-mu-ome-final shows that the group mean $\mu_{\text{OME}}$ calculated for group $G$ is sensitive to the relative proportions of the medications within each group. For two groups $A$ and $B$ where the mean dose of a given opioid $i$ is not then same ($\bar{d_{i}}^{(A)} \neq \bar{d_{i}}^{(B)}$) then the resulting means OME values are not equivalent ($\mu_{\text{OME}}^{(A)} \not\equiv \mu_{\text{OME}}^{(B)}$). 


## Influence of conversion values 

From @eq-mu-ome-final we see that the conversion value $c_{i,r}$ is a scaler applied to the mean dose ${\bar{d_i}}$. We have described the lineage of the conversion values from a estimate used in clinical practice to their integration in metrics used in research. 

The accuracy of the conversion values (how close the value used for $c_{i,r}$ is to the true value) is not well described. For example inter-personal variation may have a significant impact on the potency of a medication for a given patient however on a population level they may offer a "average conversion value set". However given that the 

We have demonstrated that the difference measured between the observed $\mu_{\text{OME}}^{(A)}$ and $\mu_{\text{OME}}^{(B)}$ is a product of not only the association of the two groups (be it statistical or causal), but also the difference in proportions of medications within each group. This is explicitly demonstrated in @sec-ome-difference-proof. 


## Aims

In this paper we will aim to demonstrate:

1. Describe how the methods of calculating OME values in the peri-operative and surgical literature.

2. Simulate and describe the influence of differing medication proportions on the results of an experiment. 

3. Demonstrate how this influence can effect results using examples from the literature. 

# Methods

## Rapid Literature Review

A rapid literature review of the use of opioid equivalence metrics in peri-operative and surgical studies over 2025 was performed. A structured search of the PUBMED database as detailed in @sec-appendix-pubmed on 7th July 2026. Papers were screened and data extracted using the Covidence platform @zotero-item-670 with a custom proforma (@sec-appendix-cov-extraction). 

```{r}
# TO get screening data:
# Covidence -> Export -> References subsection -> Title and absract screening -> Export as .csv -> Drag to data foler -> Rename to lit_screen.csn
lit_screen <- read.csv("data/lit_screen.csv")
lit_select <- read.csv("data/lit_select.csv")
lit_included <- read.csv("data/lit_included.csv")
lit_excluded <- read.csv("data/lit_excluded.csv")
lit_data <- read.csv("data/lit_data.csv") %>% filter(Reviewer.Name == "Dylan Whitaker")

n_lit_data <- nrow(lit_data)

```

## Simulated study

Two simulated studies were performed to demonstrate the sensitivity of the OME metric to differences in the proportions of medications between groups and the conversion values used. 

### Simple simulation {#sec-manuscript-simple-sim}

The first simulation aimed to quantify the relationship between the difference in proportion of a binary choice of medication between two groups, $\Delta p_i$ and the error in the conversion value used for the non-reference medication $\delta c_{i}$ on the observed difference in mean OME between the two groups ($\Delta \mu_{\text{OME}}$).

$$
\begin{align}
\text{for } i \in \{\text{morphine, oxycodone} \} \\
\Delta p_i = \Delta p_{\text{oxycodone}} = p_{\text{oxycodone}}^{\text{(control)}} - p_{\text{oxycodone}}^{\text{(intervention)}} \\
\delta c_i = c_{\text{oxycodone}} + (c_{\text{oxycodone}} * \epsilon) \\
\Delta \mu_{\text{OME}} = \mu_{\text{OME}}^{\text{(control)}} - \mu_{\text{OME}}^{\text{(intervention)}}
\end{align}
$$

```{r sim simple}

p_min <- -1
p_max <- 1
e_min <- -1
e_max <- 1
n = 10000
mu_opioid_requirment_control = 10
sigma = 0.1
treatment_effect = 0.5

simple_results <- data.frame(p = numeric(), e = numeric(), diff = numeric())
for(p in seq(p_min, p_max, by = 0.05)){
  for(e in seq(e_min, e_max, by = 0.05)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu_opioid_requirment_control,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = treatment_effect
    )
    
    simple_results <- rbind(simple_results, data.frame(p = p, e = e, diff = diff))
  }
}
```

Across the grid of values for $\Delta p_i$ and $\delta c_{i}$, `r comma(n)` total samples of two groups (control and intervention) with the observed difference in mean OME between the two groups was calculated. The expected mean dose was $\mathbb{E}$(`r mu_opioid_requirment_control`) with $\sigma$(`r sigma`)  For this simulation, the effect size between the control and treatment groups was set to `r treatment_effect`,thus $\mathbb{E}(\text{OME})_{\text{control}} = \mathbb{E}(\text{OME})_{\text{intervention}}$ and hence the expected value of $\Delta_{\text{control},\text{intervention}}\mu_{\text{OME}}$ is 0. 

### Exemplar simulation study

```{r}
case_volume <- 100
proportion_robotic <- 0.8
t_0 = as.Date("2020-01-01")
t_end = as.Date("2025-12-31")
t_robot = as.Date("2023-01-01")
proportion_morphine = 0.7
oxycodone_shortage = c(as.Date("2024-01-01"), as.Date("2024-12-31"))

mu_tot = 50
sd_tot = 10
potency_oxycodone_truth = 1.85
potency_oxycodone_ome = 2

annual_n = 30
robot_effect_size = 0.8
robot_effect_mg = mu_tot * (robot_effect_size-1)
```

Following this, a realistic exemplar study was performed to demonstrate the applicability of the this relationship on a peri-opererative study. The summary of the simulation is provided in @nte-sim-design. The difference in proportion between groups is simulated using a medication shortage that shifts baseline practice, this shortage effects one group to a greater extent thus scaling the change in proportion in a divergent manner. In this simulation the effect size of the robot is set to `r robot_effect_mg` OME mg. 

::: {#nte-sim-design .callout-note title="Simulated trial design" collapse=true}
A research team at St Somewhere Hospital want to establish if the use of a robot for a surgical procedure, reduces the post-operative opioid requirement.

They design a cohort controlled trial with two intervention groups:
  1) robot-assisted surgery - known as "robotic"
  2) standard open surgery - known as "non-robotic"
  
The primary outcome is the total opioid requirement, as measured in OME mg, in the first 24 hours post-operatively as a surrogate for the total pain caused by the surgical procedure.

The usual case volume is `r case_volume` procedures per year, these were all performed using standard surgery until `r t_robot` when a robot was acquired, and since `r scales::percent(proportion_robotic)` of cases are performed with the robot. Data was collected between `r t_0` and `r t_end`.

All patients are treated with a Patient Controlled Analgesia (PCA) after their operation and on average the anaesthetist will prescribe a morphine PCA to `r scales::percent(proportion_morphine)` of patients and an oxycodone PCA to the remainder. No other opioid medication was administered. 

During the study there was a national shortage of oxycodone from `r oxycodone_shortage[1]` to `r oxycodone_shortage[2]` and at which time all patients were prescribed a morphine PCA. Whilst the authors described this in their methods section, they did feel it would impact their results as the differing opioids were already accounted for in their OME metric. 

Patients in the standard surgery group have an expected opioid requirement ($\mu_{\text{OME}}$) of `r mu_tot`mg (sd `r sd_tot`) of morphine equivalents in the first 24 hours post-operatively. The authors expect that the robot-assisted surgery will reduce the opioid requirement by `r scales::percent(1-robot_effect_size)` and thus the expected opioid requirement for this group is `r mu_tot * (robot_effect_size)`mg of morphine equivalents in the first 24 hours post-operatively. The ground truth oxycodone potency is `r scales::percent(potency_oxycodone_truth)` relative to morphine morphine and thus the conversion value for oxycodone ($c_{\text{oxycodone}}$) is `r 1/potency_oxycodone_truth`. 
:::

All simulations were performed in `r version$version.string` and all code is available in @sec-sim. 

# Results

## Rapid Literature Review 

Of the 480 studies identified in the rapid literature search, *`r nrow(lit_included)` proceeded to data extraction and `r nrow(lit_excluded)` were ...* . 

The distributions of the study designs and intervention types are shown in @fig-study-types. 
```{r}
#| label: fig-study-types
#| fig-title: Study types

# Plot for Study.type proportions
plot_lit_study_type <- ggplot(lit_data, aes(x = Study.type)) +
  geom_bar(aes(fill = Study.type)) +
  labs(title = "Proportion of Study Types",
       x = "Study Type",
       y = "Count") +
  theme_minimal() +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_academic()+
  theme(legend.position = "none")

# Plot for Intervention proportions
plot_lit_intervnetion_type <- ggplot(lit_data, aes(x = Intervention)) +
  geom_bar(aes(fill = Intervention)) +
  labs(title = "Proportion of Interventions",
       x = "Intervention",
       y = "Count") +
  theme_minimal() +
  theme_academic()+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_academic()+
  theme(legend.position = "none")

plot_lit_study_type / plot_lit_intervnetion_type

```

With regards to the opioid conversion sets used `r lit_data %>% filter(Opioid.conversion.set == "Cited") %>% nrow()` cited another paper, whilst `r lit_data %>% filter(Opioid.conversion.set == "Included in text") %>% nrow()` included the conversion values in the methods. However for `r lit_data %>% filter(Opioid.conversion.set == "Not described") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.conversion.set == "Not described") %>% nrow()/n_lit_data)`) papers did not include any reference or citation to the opioid conversion sets used. Only `r lit_data %>% filter(Opioid.proportions.within.group == "Described") %>% nrow()` (`r scales::percent(lit_data %>% filter(Opioid.proportions.within.group == "Described") %>% nrow()/n_lit_data)`) *paper* described the proportions of opioids used between intervention groups as shown in @fig-study-results. 

```{r}
#| label: fig-study-results
#| fig-title: Proportion of studies reporting conversion set and opiooid proportions

# Prepare data for heatmap: sum of occurrences per combination
heatmap_data <- lit_data |>
  group_by(Opioid.conversion.set, Opioid.proportions.within.group) |>
  summarise(count = n(), .groups = "drop")

# Prepare data with a conditional color column for labels
ggplot(heatmap_data, aes(x = Opioid.proportions.within.group, y = Opioid.conversion.set, fill = count)) +
  geom_tile() +
  geom_label(aes(label = paste0("n=",count), color = after_scale(ifelse(fill > 0.5, "black", "white"))),
             fill = alpha("white", 0.7), size = 3.5) +
  scale_fill_viridis_c(option = "A", direction = -1) +
  labs(title = "Heatmap of Opioid Conversion Set vs Proportions",
       x = "Proportion within group",
       y = "Opioid conversion set",
       fill = "Count") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

## Simulation study 

### Simple simulation {#sec-manuscript-results-simple}

`r comma(nrow(simple_results))` different trials where performed across the grid values of $\Delta p_{\text{oxycodone}}$ (between `r p_max`, where all of the control group were treated with morphine and all of the control group were treated with oxycodone, and `r p_min`, the inverse) and $\delta c_{\text{oxycodone}}$. 

```{r plot simple grid}
#| label: fig-simple-grid
#| title: "Heatmap of the difference in proportion between group and the error in conversion value" 

library(latex2exp)

plot_simple_grid <- ggplot(simple_results, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom")+
  scale_fill_gradientn(
  colours = c("blue", "dodgerblue", "white", "orange", "red"))+
  geom_vline(xintercept = 0, linetype = "dashed", color = "black")+
  geom_hline(yintercept = 0, linetype = "dashed", color = "black")+
  annotate("text",x=-0.7, y=0.05, label = "Proportion of medication equal")+
  annotate("text",x=-0.7, y=0.95, label = "Control group: 100% morphine\nIntervention group: 0% morphine")+
  annotate("text",x=-0.7, y=-0.95, label = "Control group: 100% oxycodone\nIntervention group: 0% oxycodone")+
  annotate("text",x=-0.05, y=-0.7, label = "No error in conversion value", angle=90)+
  annotate("text",x=0.7, y=-0.05, label = "Overestimate potency ->" )+
  annotate("text",x=-0.7, y=-0.05, label = "<- Underestimate potency")

plot_simple_grid

```

@fig-simple-grid demonstrates the interaction effect on the observed difference between the control group and treatment group. Where the proportion of medication in the control and intervention groups are the same ($\Delta p_i = 0$) then we note that the true effect size of 0 is recovered, regardless of any error in the conversion value ($\delta c_i$) used. However, as the proportions between the control and treatment group deviate, resulting difference observed is biased by any error in the conversion values. Where the groups are completely unmatched ($\Delta p_i \in 1,-1$) the resulting observed difference is the due entirely to the error in potency as shown in @eq-results-simple-1.

$$
\Delta \mu_{\text{OME}} = -1 \cdot \Delta p_i \cdot \delta c_i
$$ {#eq-results-simple-1}

This scaling constant is the expected dose for the control group as demonstrated in @sec-appendix-sim-1 with @fig-appendix-simple-sim-mu. This relationship can be expressed shown in @eq-results-simple-2 (as expanded in in @eq-results-simple-2-proof).

$$
\mu^\text{intervention} = \mu^\text{control}[ 1 + \Delta p_i \cdot \delta c_i]
$$ {#eq-results-simple-2}

However, in all of the simulations described thus far the treatment effect was set at 0 meaning $\mathbb{E}(\text{OME}^\text{control}) = \mathbb{E}(\text{OME}^\text{intervention})$. This demonstrated that despite have no true treatment effect, difference in the proportion and error in conversion size, can lead to the observation of a treatment effect. This phenomenon can be explained by Simpsons paradox, whereby the observed treatment effect is conditioned on the proportion of the the medications within each group, thus creating the observation of a false treatment effect @pearlBookWhyNew2019. 

@sec-appendix-sim-2 demonstrates that as the treatment effect is increased (of decreased) it stretched the hyperbolic parabola in the Z direction. 

This saddle geometry illustrates a critical vulnerability in aggregate OME comparisons: if a conversion factor for the  medication $i$ is over-estimated ($\epsilon_i > 0$) and prevalence is simultaneously higher in the intervention group ($\Delta p_i > 0$), the error in the effect size compounds quadratically. Conversely, opposing directional errors will collapse and at the extreme values may invert the direction of the observed treatment effect.

### Study set up and interpretation

```{r }
df <- sim_patient(
  mu_tot = mu_tot,
  potency_oxycodone_truth = potency_oxycodone_truth,
  potency_oxycodone_ome = potency_oxycodone_ome,
  proportion_morphine = proportion_morphine,
  oxycodone_shortage = oxycodone_shortage,
  t_0 = t_0,
  t_end = t_end,
  t_robot = t_robot,
  annual_n = annual_n,
  robot_effect_size = robot_effect_size
)

```

`r nrow(df)` procedures were simulated using the parameters described in @nte-sim-design. The OME value for each patient was calculated use $c_{\text{morphine}}$ as 1 (as the reference medication) and $c_{\text{oxycodone}}$ as `r 1/potency_oxycodone_ome`, in contract to the ground truth value of `r 1/potency_oxycodone_truth`. The mean dose (mg), opioid and OME value for each group is shown in @tbl-sim-demo.

```{r}
#| label: tbl-sim-demo
#| tbl-title: Demographic table of simulated study population
library(tableone)

table_1 <- CreateTableOne(data=df,
               vars = c("dose", "opioid", "ome"),
               strata = "robotic",
               addOverall = TRUE,
               test=FALSE)

kableone(table_1)
```

::: {#fig-sim-density layout-ncol=3}
```{r}
#| label: fig-sim-dose-density
#| fig-title: "Density plot of dose by medication"

ggplot(df, aes(x=dose, fill=opioid))+
  geom_density(alpha=0.5)+
  theme_academic()+
  labs(,
        x="OME (mg)",
        y="Density",
        fill="Opioid")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) 
```

```{r}
#| label: fig-sim-ome-density
#| fig-title: "Density plot of ome by surgery type"

ggplot(df, aes(x=ome, fill=robotic))+
  geom_density(alpha=0.5)+
  theme_academic()+
  labs(title="Density plot of OME by robotic surgery",
        x="OME (mg)",
        y="Density",
        fill="Robotic Surgery")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) 

```

```{r}
#| label: fig-sim-ome-proportion
#| fig-title: "Bar chart demonstrating the proporotion of mediations by by surgery type"

df %>%
  count(robotic, opioid) %>%
  group_by(robotic) %>%
  mutate(proportion = n / sum(n)) %>%
  ggplot(aes(x = robotic, y = proportion, fill = opioid)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = paste0("n=",n)),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 3
  )+
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y = "Proportion of opioid", x = "") +
  theme_minimal()

```
:::

@fig-sim-density Demonstrates the distribution of the dose of each medication (@fig-sim-dose-density) and the distribution of OME values by surgery type (@fig-sim-ome-density). @fig-sim-ome-proportion shoes that within the robotic and non-robotic groups the proportion of each opioid is not equal, and the effect of the oxycodone shortage has had a larger impact on the robotic group.

```{r}
obs_robotic_ome <- df %>% filter(robotic == "Robotic") %>% summarise(mean(ome)) %>% pull() 
obs_non_robotic_ome <- df %>% filter(robotic != "Robotic") %>% summarise(mean(ome)) %>% pull()
obs_robot_effect_mg <- obs_robotic_ome - obs_non_robotic_ome
```

The mean OME value for the robotic surgery group was `r round(obs_robotic_ome, 1)` OME mg and for the non-robotic surgery group was `r  round(obs_non_robotic_ome, 1)` OME mg. This gives an observed effect size of the robot as `r round(obs_robot_effect_mg,1)` OME mg. 

```{r}
result <- analyse_dataset(df)
```

Using a two sided t-test to assess if there is a significant difference in mean OME between the robotic and standard surgery groups, the mean difference was `r round(result$mean_diff_ome,1)` OME mg (95% CI `r paste(round(result$conf_int_ome, 1), collapse = " to ")`) with a p-value of `r signif(result$p_value_ome, 3)`.

To assess if the difference in proportion of morphine and oxycodone was significant a Fishers Exact test was performed and found a p-value of `r signif(result$chisq_result_medication, 3)`.

Thus the researched from St Somewhere Hospital were able to conclude that the robotic surgery was associated with a significantly different post-operative OME dose. 

### Impact of conversion set and proportions

The study was then recreated using a grid search of different conversion values and differences proportion of post-operative oxycodone prescribing. 

```{r}
#| label: fig-sim-grid-search-1

if (PRODUCTION_QUALITY) {
  trials = 100
} else {
  trials = 3
}

simulation1 <- run_simulation(trials=trials,
               conversion_values = seq(from=1, to=4, by = 0.2),
               morphine_proportion = seq(from =0.1, to= 0.9, by = 0.05)
               )
plot_grid(simulation1,
          potency_oxycodone_truth = potency_oxycodone_truth,
          robot_effect_mg = robot_effect_mg)

```

Given a `r mu_tot` mg expected opioid requirement and a treatment effect of `r scales::percent(robot_effect_size-1)` mg we would expect the mean difference in OME of the robotic group to be `r (mu_tot * robot_effect_size)-mu_tot` mg OME.

The results shown in @fig-sim-grid-search show that where the proportions of morphine morphine prescribed post-op, (thus meaning a shortage of oxycodone has little influence on prescribing patterns and hence the effect of the exogenous shock is little) then across the values of $c_{\text{Oxycodone}}$, the true difference in treatment groups is well observed. With all situations being statistically significant. However as the proportion of post-operative prescribing shifts towards oxycodone, the observed difference in OME between the two groups becomes highly sensitive to the conversion value used. This is due to the exogenous effect of the oxycodone shortage impacting the robotic group more than the the standard surgery group. 

In this study set up, the true effect of the robot is powerful with an true effect of -10mg which is equal to 1 SD. Power calculations in @sec-appendix-simulation-power demonstrate that this is a very highly powered study. 

```{r}
robot_effect_mg = -4
robot_effect_size = (mu_tot + robot_effect_mg)/mu_tot

```

The experiment was run again but with a more modest effect of `r robot_effect_mg`mg or a post-operative OME `r scales::percent(robot_effect_size)` percent of the non-robotic group.

```{r}
#| label: fig-sim-grid-search-2

if (PRODUCTION_QUALITY) {
  trials = 100
} else {
  trials = 3
}

#Note that the robot_effect_size vairable is hard coded to the global_env above 
simulation2 <- run_simulation(trials=trials,
               conversion_values = seq(from=1, to=3.5, by = 0.2),
               morphine_proportion = seq(from =0.1, to= 0.9, by = 0.05)
               )
plot_grid_2 <- plot_grid(simulation2,
          potency_oxycodone_truth = potency_oxycodone_truth,
          robot_effect_mg = robot_effect_mg)

```

The parameters of this exemplar simulation can be mapped to the previous simple simulation. Firstly the values of $\delta p_i$ are $\epsilon$ are recovered for each trial in the grid search. These are then plotted side by side with the results of the first simulation when the same parameters as the exemplaer experiment is used. @fig-compare demonstrates that the experiment space for the example covered only values where $\Delta p_i < 0$ and for values of $ -0.5 < \epsilon < 1 $. Despite being a relativity narrow section of the possible gird space detailed in @sec-manuscript-results-simple with a modest effect size of `r robot_effect_mg` OME mg there are still values where the observed effect size is significantly different. 

```{r}
#| label: fig-compare
#| fig-title: Comparsion of simple simulation results and exemplar simulation results. 

bounds <- c(min(simulation2$delta_p_oxycodone),
            max(simulation2$delta_p_oxycodone),
            min(simulation2$epsilon),
            max(simulation2$epsilon))

simple_results <- data.frame(p = numeric(), e = numeric(), diff = numeric())
for(p in seq(p_min, p_max, by = 0.05)){
  for(e in seq(e_min, e_max, by = 0.05)){
    
    diff <- sim_simple(
      n = n,
      delta_p_i = p,
      sigma = sigma,
      mu_opioid_requirment_control = mu_tot,
      potency_oxycodone = 2,
      error_in_c_oxycodone = e,
      treatment_effect = robot_effect_size -1
    )
    
    simple_results <- rbind(simple_results, data.frame(p = p, e = e, diff = diff))
  }
}


plot_simple_grid <- ggplot(simple_results, aes(y = p, x = e, fill = diff)) +
  geom_tile() +
  labs(,
       y = TeX(r'($\Delta  p_i$ (Difference in proportion))'),
       x = TeX(r'($\epsilon$ (Error in Conversion Value for Oxycodone) )'),
       fill = TeX(r'(Observed difference ($\Delta \mu_{{OME}})$)')
       ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "none")+
  symmetric_fill_scale(20)+
  geom_rect(ymin=bounds[1],
            ymax=bounds[2],
            xmin=bounds[3],
            xmax=bounds[4],
            colour="black",
            fill=alpha("grey",0))

plot_grid_2 <- ggplot(simulation2, aes(x=conversion_value, y=morphine_proportion, fill=mean_diff_ome)) +
  geom_tile() +
  symmetric_fill_scale(20)+
  labs(y = TeX(r'(Proportion of post-op perscribing as morphine )'),
       x = TeX(r'($\c_{oxycodone}$ (Conversion Value for Oxycodone used) )')) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "none")
  

plot_simple_grid + plot_grid_2

layout <- c(
  area(t = 1, l = 1, b = 10, r = 10),
  area(t = 6, l = 11, b = 7, r = 20)
)
plot_simple_grid + plot_grid_2 +
  plot_layout(design = layout)

```

### Simulation of ...et al 

The final simulation looks at a study performed by xx et al. This study looked at the effect of xxx on yyyy and found a observed effect size of xxx. 

They did not report the difference of proportion of medications between their treatment groups. We will apply conversion values that have been used across the literature to calculate. One issue is that we do not know the true conversion values of the medications. 

# Conclusion


## Implications




## Limitations
In the simulation studies we have only considered situations where medication choice is binary between the reference medication and the alternative (oxycodone). We hypothesise that in some situations with multiple opioids, due to Orthogonality, that multiple small errors in the estimate of $c_i$ and the true value of $c_i$ may generally cancel out and thus have little impact on the final estimate. This may be achived using machine learning tools that consider the co-linearity between the features of the dose of each medication and the outocme of total opioid burden or OME. 


## Generative AI statement

Generative artificial intelligence tools where used to aid the writing of this manuscript and in production of code. These tools included Gemini 3.1 models (Pro, Flash and Flash-Lite), Claude sonnet 4.6 (via Posit AI) and Github Copilot (via Posit Code Assistant). 
All ideas and methods are of the authors own, all text and code was reviewed and edited following generation.

# Appendix



