---
editor_options: 
  chunk_output_type: console
---

# Code for simulation study {#sec-sim}

## Code repository

The code will be hosted on a public github. 

## Simple simulation {#sec-sim-simple}

The `sim_simple` function simulates a simple experiment with two equally groups, control and treatment, for total `n` samples. `delta_p_i` controls the difference in proportion of medications `i` between the two groups. For each patient a expected opioid requirement is sampled from a normal distribution with mean `mu_opioid_requirment_control` and standard deviation `sigma`. The treatment group has a mean opioid requirement that is increased by the factor of `treatment_effect`.
For samples where `i` is `"oxycodone"`, the expected opioid requirement is converted to a oxycodone dose using `potency_oxycodone` and then converted back into expected dose of morphine (the OME) using the conversion factor `potency_oxycodone` +/- an error term `error_in_c_oxycodone`.

The function then returns the difference between the OME dose between the control and treatment group. 

```{r sim_simple}
#| echo: true
#| collape: true

sim_simple <- function(n, # total number of patients
                       delta_p_i, # Difference in proportion of i between groups,
                       i = c("morphine", "oxycodone"), # The two medications
                       mu_opioid_requirment_control = 1, 
                       sigma = 0.1,
                       treatment_effect = 0,
                       potency_oxycodone,
                       error_in_c_oxycodone = 0){
  
   group_control = data.frame(
     group = "control",
     drug = sample(i, size = n/2, replace = TRUE, prob = c(0.5+delta_p_i/2, 0.5-delta_p_i/2)),
     opioid_req = rnorm(n/2, mean = mu_opioid_requirment_control, sd = sigma)
   )
  
  group_treatment = data.frame(
     group = "treatment",
     drug = sample(i, size = n/2, replace = TRUE, prob = c(0.5-delta_p_i/2, 0.5+delta_p_i/2)),
     opioid_req = rnorm(n/2, mean = mu_opioid_requirment_control+(mu_opioid_requirment_control*treatment_effect), sd = sigma)
   )
  
  result <- rbind(group_control,group_treatment) %>%
    mutate(dose_recived = ifelse(drug == "morphine", opioid_req, opioid_req/potency_oxycodone),
           ome = ifelse(drug == "morphine", dose_recived, dose_recived*(potency_oxycodone+(potency_oxycodone*error_in_c_oxycodone))
                        ))
  
  analysis <- result %>%
    group_by(group) %>%
    summarise(proportion_morphine = sum(drug == "morphine")/n(),
              mean_ome = mean(ome),
              n = n(),
              .groups = "drop") 
    
  return(analysis$mean_ome[2] - analysis$mean_ome[1])
  
}

```

## Exemplar simulation {#sec-sim-exemplar}

The `sim_patient` function produces a study dataset by seeding procedure dates given the annual case load and the study period. Then for each procedure sampling whether the procedure is robotic or standard, and finally sampling which opioid is used.

The function then estimates the mean analgesia requirement based on treatment and opioid and samples the dose from a normal distribution.

```{r sim_patient}
#| echo: true
#| collape: true

sim_patient <- function(
  mu_tot = 50,
  potency_oxycodone_truth = 1.85,
  potency_oxycodone_ome = 2,
  proportion_morphine = 0.7,
  oxycodone_shortage = c(as.Date("2024-01-01"), as.Date("2024-12-31")),
  t_0 = as.Date("2020-01-01"),
  t_end = as.Date("2025-12-31"),
  t_robot = as.Date("2023-01-01"),
  annual_n = 30,
  robot_effect_size = 0.8
){
  # Calculate total number of patients over the study period
  total_days <- as.numeric(t_end - t_0)
  total_years <- total_days / 365.25
  total_n <- round(total_years * annual_n)

  # Sample random dates
  dates <- as.Date(runif(total_n, min = as.numeric(t_0), max = as.numeric(t_end)), origin = "1970-01-01")

  # Sample if procedure is robotic (after 2023, 80% chance)
  robotic <- ifelse(dates >= t_robot, rbinom(total_n, 1, 0.8) == 1, FALSE)

  # Sample which opioid is used
  # During 2024, 100% morphine. Otherwise, 40% oxycodone, 60% morphine.
  is_shortage <- dates >= oxycodone_shortage[1] & dates <= oxycodone_shortage[2]
  opioid <- ifelse(is_shortage, "M",
    ifelse(rbinom(total_n, 1, proportion_morphine) == 1, "M", "O")
  )

  # Estimate mean analgesia requirement based on treatment
  mu <- ifelse(robotic, mu_tot * robot_effect_size, mu_tot)

  # Set mean dose depending on opioid type (using true potency)
  mu_op <- ifelse(opioid == "M", mu, mu / potency_oxycodone_truth)

  # Sample dose from normal distribution
  dose <- rnorm(total_n, mean = mu_op, sd = 10)

  # Round the doses to an integer of milligrams and prevent negative doses
  dose <- pmax(0, round(dose))

  # Create final dataframe
  df <- data.frame(
    date = dates,
    robotic = robotic,
    opioid = opioid,
    dose = dose
  )

  # Sort by date for chronological sense
  df <- df[order(df$date), ]
  rownames(df) <- NULL
  
  df <- df %>%
      mutate(ome = case_when(
        opioid == "M" ~ dose,
        opioid == "O" ~ dose * 1*potency_oxycodone_ome
      ))
 
  df <- df %>% mutate(
    robotic = factor(robotic, labels = c("Non-Robotic", "Robotic")),
    opioid = factor(opioid, levels = c("M", "O"), labels = c("Morphine", "Oxycodone"))
  )

  return(df)
}
```

The `analyse_dataset` function acts to hypothesis test the difference in mean OME between robotic and non-robotic procedures.

```{r analyse_dataset}
#| echo: true
#| collape: true

analyse_dataset <- function(df){
  # Perform t-test for difference in mean OME between robotic and non-robotic procedures
  t_test_result <- t.test(ome ~ robotic, data = df)

  # analyse the difference between the opioid proportions
  fishers_result <- fisher.test(table(df$robotic, df$opioid)) 
  
  # Return results as a list
  return(list(
    mean_diff_ome = diff(t_test_result$estimate),
    p_value_ome = t_test_result$p.value,
    conf_int_ome = t_test_result$conf.int,
    chisq_result_medication = fishers_result$p.value
  ))
}

```

The `run simulation` function will run $t$ trials for each of the values of the conversion values $c_{\text{Oxycodone}}$ and the proportion of morphine perceribed post-operativly. The results will be stored in a data frame and then summarized using the mean of the results of the `analyse_dataset` function.  

```{r run_simulation}
#| echo: true
#| collape: true


run_simulation <- function(trials = 5,
                          conversion_values = seq(from=1, to=4, by = 0.2),
                          morphine_proportion = seq(from =0.1, to= 0.9, by = 0.05))
                          {

  for(t in 1:trials){
    for(i in conversion_values){
      for(j in morphine_proportion){
    
    df <- sim_patient(
      mu_tot = mu_tot,
      potency_oxycodone_truth = potency_oxycodone_truth,
      potency_oxycodone_ome = i,
      proportion_morphine = j,
      oxycodone_shortage = oxycodone_shortage,
      t_0 = t_0,
      t_end = t_end,
      t_robot = t_robot,
      annual_n = annual_n,
      robot_effect_size = robot_effect_size
    )
    
    delta_p_oxycodone <- df |>
      group_by(robotic) |>
      summarise(p = mean(opioid == "Oxycodone")) |>
      pull(p) |>
      diff()
    
    epsilon <- (i - potency_oxycodone_truth)/potency_oxycodone_truth
    
    
    result <- analyse_dataset(df)
    
    if(t == 1 & i == conversion_values[1] & j == morphine_proportion[1]){
      
      results_grid <- data.frame(
        trial = t,
        conversion_value = i,
        morphine_proportion = j,
        mean_diff_ome = result$mean_diff_ome,
        p_value_ome = result$p_value_ome,
        p_value_fishers = result$chisq_result_medication,
        delta_p_oxycodone = delta_p_oxycodone,
        epsilon = epsilon
      )
    }else{
    # Store results
    results_grid <- rbind(results_grid, data.frame(
      trial=t,
      conversion_value = i,
      morphine_proportion = j,
      mean_diff_ome = result$mean_diff_ome,
      p_value_ome = result$p_value_ome,
      p_value_fishers = result$chisq_result_medication,
      delta_p_oxycodone = delta_p_oxycodone,
      epsilon = epsilon
    ))
    }
    
  }
  }
}

  results_grid <- results_grid %>% group_by(conversion_value, morphine_proportion) %>%
    summarise(mean_diff_ome = mean(mean_diff_ome),
            p_value_ome = mean(p_value_ome),
            p_value_fishers = mean(p_value_fishers),
            delta_p_oxycodone = mean(delta_p_oxycodone),
            epsilon = mean(epsilon),
            .groups = "drop") %>%
    mutate(significant = ifelse(p_value_ome < 0.05, "*", ""))

  return(results_grid)

} 

```

`plot_grid()` function plots the heatmap of the results of the exemplar simulation. 

```{r plot_grid}
#| echo: true
#| collape: true


plot_grid <- function(results_grid,
                      potency_oxycodone_truth,
                      robot_effect_mg){

midpoint <- robot_effect_mg
min_val <- min(results_grid$mean_diff_ome, na.rm = TRUE)  # adjust column name as needed
max_val <- max(results_grid$mean_diff_ome, na.rm = TRUE)

  ggplot(results_grid, aes(x=conversion_value, y=morphine_proportion, fill=mean_diff_ome)) +
  geom_tile() +
  geom_point(data = subset(results_grid, significant == "*"), shape = 4, color = "purple") +
  symmetric_fill_scale(20)+
  labs(title = ,
       x = TeX(r'(Value used for $c_{oxycodone}$)'),
       y = "Proportion of post-operative morphine perscribing",
       fill = "Mean difference in OME") +
   geom_vline(xintercept = potency_oxycodone_truth, linetype = "dashed", color = "red")+
    theme_minimal() +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )
 
}

```

`symetic_fill_scale` is a colour gradient scale that allows control of the scale to ensure consistancy between plots 

```{r}


symmetric_fill_scale <- function(limit, type="fill") {
  
  if(type=="fill"){
  scale_fill_gradientn(
    colours = c("darkblue", "blue", "dodgerblue", "white", "orange", "red", "darkred"),
    values  = scales::rescale(c(-limit, -limit/2, 0, limit/2, limit)),
    limits  = c(-limit, limit)
  )
    }else{
    scale_colour_gradientn(
    colours = c("darkblue", "blue", "dodgerblue", "white", "orange", "red", "darkred"),
    values  = scales::rescale(c(-limit, -limit/2, 0, limit/2, limit)),
    limits  = c(-limit, limit)
  )
  }
}

```
