---
title: "Code for simulation study"
---

# Functions

The `sim_patient` function produces a study dataset by seeding procedure dates given the annual case load and the study period. Then for each procedure sampling whether the procedure is robotic or standard, and finally sampling which opioid is used.

The function then estimates the mean analgesia requirement based on treatment and opioid and samples the dose from a normal distribution.

```{r sim_patient }
#| echo: true
#| collape: true

sim_patient <- function(
  mu_tot = 50,
  potency_oxycodone_truth = 1.85,
  potency_oxycodone_ome = 2,
  proportion_morphine = 0.7,
  oxycodone_shortage = c(as.Date("2024-01-01"), as.Date("2024-12-31")),
  t_0 = as.Date("2020-01-01"),
  t_end = as.Date("2025-12-31"),
  t_robot = as.Date("2023-01-01"),
  annual_n = 30,
  robot_effect_size = 0.8
){
  # Calculate total number of patients over the study period
  total_days <- as.numeric(t_end - t_0)
  total_years <- total_days / 365.25
  total_n <- round(total_years * annual_n)

  # Sample random dates
  dates <- as.Date(runif(total_n, min = as.numeric(t_0), max = as.numeric(t_end)), origin = "1970-01-01")

  # Sample if procedure is robotic (after 2023, 80% chance)
  robotic <- ifelse(dates >= t_robot, rbinom(total_n, 1, 0.8) == 1, FALSE)

  # Sample which opioid is used
  # During 2024, 100% morphine. Otherwise, 40% oxycodone, 60% morphine.
  is_shortage <- dates >= oxycodone_shortage[1] & dates <= oxycodone_shortage[2]
  opioid <- ifelse(is_shortage, "M",
    ifelse(rbinom(total_n, 1, proportion_morphine) == 1, "M", "O")
  )

  # Estimate mean analgesia requirement based on treatment
  mu <- ifelse(robotic, mu_tot * robot_effect_size, mu_tot)

  # Set mean dose depending on opioid type (using true potency)
  mu_op <- ifelse(opioid == "M", mu, mu / potency_oxycodone_truth)

  # Sample dose from normal distribution
  dose <- rnorm(total_n, mean = mu_op, sd = 10)

  # Round the doses to an integer of milligrams and prevent negative doses
  dose <- pmax(0, round(dose))

  # Create final dataframe
  df <- data.frame(
    date = dates,
    robotic = robotic,
    opioid = opioid,
    dose = dose
  )

  # Sort by date for chronological sense
  df <- df[order(df$date), ]
  rownames(df) <- NULL
  
  df <- df %>%
      mutate(ome = case_when(
        opioid == "M" ~ dose,
        opioid == "O" ~ dose * 1*potency_oxycodone_ome
      ))
 
  df <- df %>% mutate(
    robotic = factor(robotic, labels = c("Non-Robotic", "Robotic")),
    opioid = factor(opioid, levels = c("M", "O"), labels = c("Morphine", "Oxycodone"))
  )

  return(df)
}
```

The `analyse_dataset` function acts to hypothesis test the difference in mean OME between robotic and non-robotic procedures.

```{r analyse_dataset}
#| echo: true
#| collape: true

analyse_dataset <- function(df){
  # Perform t-test for difference in mean OME between robotic and non-robotic procedures
  t_test_result <- t.test(ome ~ robotic, data = df)

  
  # analyse the difference between the opioid proportions
  fishers_result <- fisher.test(table(df$robotic, df$opioid)) 
  
  # Return results as a list
  return(list(
    mean_diff_ome = diff(t_test_result$estimate),
    p_value_ome = t_test_result$p.value,
    conf_int_ome = t_test_result$conf.int,
    chisq_result_medication = fishers_result$p.value
  ))
}

```
